### 2023-01-23 README.Rmd
First commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# se.norway.shoredate.14c

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2023-01-24 analysis/paper/paper.Rmd
Including raw data and initial script for SPD of shoreline dates

---
title: "Contrasting summed probability distributions of shoreline and radiocarbon dates on the Norwegian Skagerrak coast"
# author: 
# - 'Isak Roalkvam'
# - 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - IsakRoalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [iakh]
      # correspondence: true
#   - name: Steinar Solheim
#     email: steinar.solheim@khm.uio.no
#     institute: [mch]
#     correspondence: false
institute:
  - iakh: Institute of Archaeology, Conservation and History, University of Oslo
#   - mch: Museum of Cultural History, University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

<!-- Keywords: `r rmarkdown::metadata$keywords` -->

<!-- Highlights: `r rmarkdown::metadata$highlights` -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

knitr::opts_knit$set(
  eval.after = "fig.cap"
)

# Set-up so that I can call floatbarrier using regfloat=TRUE in the chunk 
# headings. See: 
# https://gist.github.com/burchill/8873d2ade156b27e92a238a774ce2758
knitr::knit_hooks$set(plot = function (x, options) {
  float_correct <- function(f, y, opts)  {
    if (is.null(opts$regfloat) || opts$regfloat==FALSE)
      paste0(f(y, opts), "\n\n\\FloatBarrier\n")
    else
      f(y, opts)
  }
  if (!is.null(options$out.width) || !is.null(options$out.height) ||
      !is.null(options$out.extra) || options$fig.align != "default" ||
      !is.null(options$fig.subcap)) {
    if (is.null(options$fig.scap))
      options$fig.scap = NA
    return(float_correct(knitr:::hook_plot_tex, x, options))
  }
  return(float_correct(knitr:::hook_plot_md_base, x, options))
})
```

```{r packages, echo = FALSE}
library(here)
library(dplyr)
library(kableExtra)
library(sf)
```

```{r data, include = FALSE}
surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
```

# Introduction
Population size is regarded as one of the primary drivers of cultural variation, and is of critical importance to our understanding of past human societies [@shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018a; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017]. This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

# Background
Large parts of the post-glacial landscape of Northern Scandinavia is characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. As coastal foragers appear to have predominantly settled on or close to the contemporaneous shoreline, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement---a method known as shoreline dating. This is not least useful for dating the large number of surveyed sites in the region where other temporal data such as radiocarbon dates or typological indicators in artefact inventories that often follow with an excavation are not available.

The frequency of shoreline dated sites has also been compared to RSPDs in the past [@solheim2018; @tallavaara2020]. However, this has been done by finding point estimates of shoreline dates that are then aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not take into account the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of this analysis indicate that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility [see @roalkvam2023 for details]. Thus, the geographical and temporal limits outlined above are also used for this study.

<!-- This paper is decidedly exploratory. While some speculative thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. The data have thus been explored using a wide range of techniques in an attempt to unpick patterns and grasp the relationship between the variables.  -->

## Population dynamics and summed probabilities

The dates as data approach is dependent on there being a direct link between the past generation of material that ultimately become ^14^C samples, and population size. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and if used as a proxy for population size is therefore dependent on there being a connection between site count and population size. If a comparison of these proxies do not or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflect true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to explain any observed pattern.

To what degree the radiocarbon record is determined by past population numbers might vary both geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns and settlement nucleation and dispersion [@palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted. Before any such fluctuations are given any substantive interpretation, however, there are a host of methodological issues that have to be considered.

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represents the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data---the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that an SPD cannot be used directly to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, as introduced by [@shennan2013]. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, or exponential or logistic distributions, fit to the observed RSPD by means of regression. These models are chosen *a priori* with reference to common long-term population dynamics. The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside this envelope is used to estimate a p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these should be interpreted [@timpson2021]. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of regression will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model can be rejected as an explanation of the data, and does not provide statistical justification for interpreting the deviations themselves as meaningful demographic signals, as has often been the case [see review by @crema2022; @timpson2021]. Instead of only applying a null-hypothesis significance testing approach using a few models chosen *a priori* to fit to the data, the analysis performed here instead follows the model-fitting approach of @timpson2021.

<!-- @timpson2021 have recently presented a framework for fitting continuous piecewise linear (CPL) models along with more commonly applied models to ^14^C-data and evaluating their relative performance. The CPL models consist of a series of linear pieces, representing phases, defined by n hinges, represented demographic events. These piecewise models allows for the identification of key demographic events and a conservative description of population dynamics between these. One major benefit of the approach is thus also that it provides a clearer inferential justification for a substantive interpretation of the results. Schwarz criterion, apparently misnamed BIC [],  -->

# Methods and data

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2022. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites without relevant ^14^C-dates that were originally shoreline dated were included in the SSPD along with the retained surveyed sites. Any sites situated at elevations that result in a shoreline date earlier than 9460 BCE were then excluded. No sites are yet verifiable known to be older than around 9300 BCE [@glorstad2016], and so while the shoreline dates indicate that some surveyed sites are potentially older than this, they have not been properly investigated and . giving a total of 934 shoreline dated sites in the final SSPD. The borders of the municipalities where the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were excluded due the issue of marine reservoir effects [@nielsen2019, 83].

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in an repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [@Marwick2017].

```{r tab1}
Quality <- c(1, 2, 3, 4, 5, 6)
Definition <- c("Site delineated by use of a GNSS-device, or a securely georeferenced record. Extensive database entry.", "Secure spatial data. Slight disturbance of the site or somewhat lacking database record.", "Secure spatial data. Damaged site, such as outskirts of a quarry, and/or very limited database entry.", "Surveyed by archaeologists. However, the database entry is extremely limited/unclear, the site geometry is only given\\\\as a point or small automatically generated circle, and/or finds are from the topsoil of a field.", "Likely site but uncertain spatial information. Typical example is recurring stray finds in a field or other larger area.", "Single stray find or unverified claims/suggestions of possible site.")
Count <- c(nrow(filter(surveyed, quality == 1)), nrow(filter(surveyed, quality == 2)), nrow(filter(surveyed, quality == 3)), nrow(filter(surveyed, quality == 4)),
           nrow(filter(surveyed, quality == 5)), nrow(filter(surveyed, quality == 6)))

df <- data.frame(Definition, Quality, Count)

kableExtra::kable(df, booktabs = TRUE,
      caption = "Quality scoring of site records of surveyed sites retrieved from the national heritage database Askeladden. The scoring system was first used in Roalkvam (2020).", escape = FALSE) %>% kableExtra::kable_styling(latex_options= c("striped", "scale_down"), stripe_color = "gray!20",)
```

```{r map, echo = FALSE, fig.cap = "Map of the study area and analysed sites. Black lines indicate the borders between municipalities. The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1. Of the excavated sites, 102 have been dated by means of shoreline dating.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/map.png"))
```

## Summed probability of calibrated radiocarbon dates

Analysis of the ^14^C-dates were done using the R packages *rcarbon* [@crema2021]. To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned and dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other. When calibrated, these are summed and normalised to sum to unity before being included in the final RSPD. Individual ^14^C-dates were not normalised. This follows from a peculiarity in the calibration procedure which can lead the probability mass of a date to sum to more than one [@weninger2015]. As a consequence of this, normalising the dates before summing has been shown to create spurious spikes in the RSPD. All calibration was done using the IntCal20 calibration curve [@reimer2020]. Edge-effects were handled by only including dates that fall inside the range between xx--2500 BCE with more than 50% probability [cf. @timpson2021].

The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013] by fitting two models to the observed RSPD. Here this was an exponential model, fit by means of regression, and a uniform model. A series of individual calendar years are then drawn from this distribution with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors are added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 1000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations were then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed curve deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null model can be rejected.

## Summed probability of shoreline dated sites

Summing the probability of the shoreline dated sites and the model-fitting procedures follows the same fundamental structure as that used for the radiocarbon dates, and is based on re-purposed programming code from *rcarbon*. However, idiosyncrasies in the dating method does necessitate some adjustments. First, an illustration of the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate). Four geologic reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here. These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [e.g. @svendsen1987]. The first step in the dating procedure is therefore to interpolate the shoreline displacement to the location to be dated.  This is done by inverse distance weighting [e.g. @conolly2020], interpolating the sea-level for each calendar year along the geologic shoreline displacement curves based on the distance from the site to be dated to the isobases of the displacement curves. The analysis in @roalkvam2023 was based on simulating the distance between sites and shoreline by drawing on independent temporal data on site use as evidenced by ^14^C-dates. Sites dated to before 2500 BCE were found to have a likely elevation above sea-level that was reasonably approximated  by a gamma distribution with a decay ratio of 0.168. When shoreline dating, this probability is sequentially stepped through at increments of 0.1 m and transferred to the calendar scale by uniformly distributing the probability across calendar years in the range between the lower and upper limit of the interpolated displacement curve. Here this is done with resoltuon of 10 years on the calendar scale. This procedure gives the the shoreline date of a site. The procedure is thus similar to how a radiocarbon date is calibrated, and insights from studies that deal with the numerical treatment of radiocarbon dates and their summation can consequently be useful as analogy for understanding the same properties for the shoreline dates.

```{r shoredate, echo = FALSE, fig.cap = "A) Example location relative to the isobases of the displacement curves.  B) The geologic displacement curves and the curve interpolated to the example location. C) Resulting shoreline date in light grey on the x-axis. The black line underneath marks the 95\\% HDR. The dashed line marks the elevation of the example location. The exponential function on the y-axis decays with ratio $\\lambda$ and represents the likely elevation of a site above sea-level when it was in use. ", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

<!-- The SSPD was subjected to the same procedure for CPL model fitting and comparison as teh RSPD above. Dates with 50% probability distribution  -->

In figure  the shoreline dated sites are plotted according to their elevation. However, due to the variable uplift rates within the study region, the same elevation does not equate to the same date and so this does not directly tell us much about their temporal distribution. As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date than the same elevation further to the north-east. If the elevation of the sites is instead transformed to point-estimate of their date, used by couplng the mean elevation of the site and the mean of the displacement curve interpolated to each site location, the temporal distribution instead looks like the one provided in figure. This more resembles how the dates have usually been treated.  are instead are transformed to point-estimates of shoreline dates for each site  Looking at the distribution across elevations. Furthermore, one way that help in understanding the is to see tham as equivalent to uncalibrated radiocarbon dates. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed, as this should not create spurious spikes in the SSPD.

In much the same way as characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on the same isobase. With the analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites (Figure \@ref(fig:incpolys)). These intervals were then assigned a weight based on how the density of observed sites is distributed among them.

```{r incpolys, echo = FALSE, fig.cap = "Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 943) as distributed across the 2km wide line segments that run perpendicular to the shoreline gradient. A displacement curve has been interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/incpolys.png"))
```

Following from the commutative nature of the shoreline dates, the Monte Carlo simulation is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. This is equivalent to the *calsample* sampling method from *rcarbon* [@crema2021]. The calendar date is then 'uncalibrated'---to follow the RSPD terminology---to an elevation range from which a single elevation value is then drawn with a uniform probability, using intervals of 5cm. This elevation value is then shoreline dated using the displacement curve for the relevant 2km interval. The dates are then summed, and the process repeated a total of 1000 times. Finally, the 97.5% highest and 2.5% lowest summed probability at each calendar year across all simulations is retrieved to create the 95% critical envelope and find the global p-value.

# Results

```{r spduniform, echo = FALSE, fig.cap = "[Shoreline sims in this version have only been run 50/1000 times] A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model. B) Same as A, but with a exponential model. C) Summed probability of calibrated radiocarbon dates (n =, bins = 678) compared to a uniform null model. D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/spds_uniform.png"))
```

<!-- # Discussion -->

<!-- # Conclusion -->

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

### 2023-01-24 R/shoreline_spd.R

@@ -1,0 +1,88 @@
+library(tidyverse)
+library(sf)
+library(terra)
+library(here)
+library(shoredate)
+
+# For reproducibility
+set.seed(1)
+
+# Load data
+surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
+excavated <- st_zm(read_sf(here("analysis/data/raw_data/excavated_sites.gpkg")))
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+site_dat <- read.csv(here("analysis/data/raw_data/sites.csv"))
+
+
+###### Prepare site data for shoreline dating ######
+
+# Exclude sites that haven't been excavated and which haven't been shoreline
+# dated
+shore_sites <- site_dat %>% dplyr::filter(investigated == "t") %>%
+  dplyr::filter(dating_method == "shore_typo" & !is.na(reported_earliest_bce)
+                & !is.na(reported_min_elev))
+
+# Join shoreline dated excavated sites with the spatial data for the sites
+excshore <- st_as_sf(
+  dplyr::left_join(shore_sites, excavated,
+                    by = c("name" = "site_name", "ask_id"))
+  ) %>%
+  dplyr::filter(!(is.na(st_dimension(geom))))
+
+# Find mean reported elevation for the excavated sites
+excshore$elev <- rowMeans(subset(st_drop_geometry(excshore),
+                                 select = c(reported_min_elev,
+                                            reported_max_elev)))
+
+# Find elevation of surveyed sites
+surveyed$elev <- terra::extract(dtm, vect(surveyed), fun = mean)[, -1]
+
+# Exclude surveyed sites with a quality score worse than 3
+survq <- dplyr::filter(surveyed, quality < 4)
+
+#
+# surveyed <- surveyed %>% dplyr::mutate(lab = ifelse(quality < 4, "t", "f"))
+# excavated$lab <- "y"
+
+# Select and rename columns needed to perform shoreline dating
+rcolexcshore <- excshore %>% dplyr::select(ask_id, elev)
+rcolsurvq <- survq  %>% dplyr::select(askeladden_id, elev) %>%
+  dplyr::rename(ask_id = askeladden_id)
+
+# Combine the data
+sites <-rbind(rcolexcshore, rcolsurvq)
+
+###### Shoreline dating ######
+step_reso = 0.1
+
+shorelinedates <- shoredate::shoreline_date(sites,
[…]

### 2023-02-28 analysis/paper/paper.Rmd
Add files via upload

@@
 ---
-title: "Contrasting summed probability distributions of shoreline and radiocarbon dates on the Norwegian Skagerrak coast"
+title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Stone Age on the Norwegian Skagerrak coast"
 # author: 
@@
 Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015].
-What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017].
-This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).
+What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021].
+This paper reports on and beings to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).
 
@@
 This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement---a method known as shoreline dating.
-This is not least useful for dating the large number of surveyed sites in the region where other temporal data such as radiocarbon dates or typological indicators in artefact inventories that often follow with an excavation are not available.
+This is not least useful for dating the large number of surveyed sites in the region where other temporal data that often follow with an excavation are not available, such as radiocarbon dates or typological indicators in artefact inventories.
 
@@
 The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)).
-The results of this analysis indicate that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility [see @roalkvam2023 for details].
+The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility.
 Thus, the geographical and temporal limits outlined above are also used for this study.
 
-<!-- This paper is decidedly exploratory.
+This paper is decidedly exploratory.
 While some speculative thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses.
-The data have thus been explored using a wide range of techniques in an attempt to unpick patterns and grasp the relationship between the variables.
--->
+The data have thus been explored using a set of standard models to begin to unpick patterns and grasp the relationship between the variables.
+Furthermore, the dates as data approach is dependent on there being a direct link between the past generation of material that ultimately become ^14^C samples, and population size.
+The sum of shoreline dated sites, on the other hand, is determined by site frequency, and if used as a proxy for population size is therefore dependent on there being a connection between site count and population size.
+If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics.
+The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.
 
@@
 
-The dates as data approach is dependent on there being a direct link between the past generation of material that ultimately become ^14^C samples, and population size.
-The sum of shoreline dated sites, on the other hand, is determined by site frequency, and if used as a proxy for population size is therefore dependent on there being a connection between site count and population size.
-If a comparison of these proxies do not or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflect true population dynamics.
-The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to explain any observed pattern.
-
 To what degree the radiocarbon record is determined by past population numbers might vary both geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations.
 One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations.
-Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns and settlement nucleation and dispersion [@palmisano2017].
-However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted.
+Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017].
+However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted and possibly weighted differently according some criteria.
 Before any such fluctuations are given any substantive interpretation, however, there are a host of methodological issues that have to be considered.
@@
 It is the combined representation of a range of possible explanations for the data---the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021].
-This means that an SPD cannot be used directly to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018].
+This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021].
 While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.
 
-The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, as introduced by [@shennan2013].
+The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, as introduced by @shennan2013 and later expanded upon by @timpson2014.
 This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model.
-These null-models are typically a uniform, or exponential or logistic distributions, fit to the observed RSPD by means of regression.
-These models are chosen *a priori* with reference to common long-term population dynamics.
+These null-models are typically a uniform, exponential or logistic distribution.
+These are chosen *a priori* with reference to common long-term population dynamics and are parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE).
 The result from these simulations are then used to create a 95% critical envelope representing the null model.
-The proportion of the observed RSPD that falls outside this envelope is used to estimate a p-value indicating whether the null model can be rejected.
+The proportion of the observed RSPD that falls outside this envelope is used to estimate a global p-value indicating whether the null model can be rejected.
 In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model.
-However, care has to be taken in how these should be interpreted [@timpson2021].
+However, care has to be taken in how these are interpreted.
 First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to.
 Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used.
-While a model fit by means of regression will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine.
-Finally, the p-value only indicates whether or not the null-model can be rejected as an explanation of the data, and does not provide statistical justification for interpreting the deviations themselves as meaningful demographic signals, as has often been the case [see review by @crema2022; @timpson2021].
-Instead of only applying a null-hypothesis significance testing approach using a few models chosen *a priori* to fit to the data, the analysis performed here instead follows the model-fitting approach of @timpson2021.
+While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine.
+Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting local deviations themselves as meaningful demographic signals, as has often been the case [see review by @crema2022; @timpson2021].
 
-<!-- @timpson2021 have recently presented a framework for fitting continuous piecewise linear (CPL) models along with more commonly applied models to ^14^C-data and evaluating their relative performance.
-The CPL models consist of a series of linear pieces, representing phases, defined by n hinges, represented demographic events.
-These piecewise models allows for the identification of key demographic events and a conservative description of population dynamics between these.
-One major benefit of the approach is thus also that it provides a clearer inferential justification for a substantive interpretation of the results.
-Schwarz criterion, apparently misnamed BIC [],  -->
+## Model comparison and continous piece-wise linear models
+
+The procedure outlined above represent the standard approach when modelling population dynamics in archaeology.
+This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of sites over time could be made.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves.
+As a result, by improving on @goldstein, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics, published with the accompanying R package  *ADMUR*.
+This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally testing the best model using the Monte Carlo approach outlined above.
+The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these.
+One major benefit of the approach is thus that it provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
+Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.
 
@@
 All sites with a quality score of 4 or worse were excluded from further analysis.
+Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded.
+This marks the latest start-date among the employed displacement curves, and no sites are yet known to be older than around 9300 BCE [@glorstad2016].
 Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2022.
 Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis.
-The 102 excavated sites without relevant ^14^C-dates that were originally shoreline dated were included in the SSPD along with the retained surveyed sites.
-Any sites situated at elevations that result in a shoreline date earlier than 9460 BCE were then excluded.
-No sites are yet verifiable known to be older than around 9300 BCE [@glorstad2016], and so while the shoreline dates indicate that some surveyed sites are potentially older than this, they have not been properly investigated and .
-giving a total of 934 shoreline dated sites in the final SSPD.
-The borders of the municipalities where the shoreline dated sites are located were used to limit the radiocarbon sample.
+The 102 excavated sites in the dataset without relevant ^14^C-dates and that were originally shoreline dated in the reports were included in the SSPD along with the retained surveyed sites.
+This gave a total of  shoreline dated sites in the final SSPD.
+
+The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample.
 Radiocarbon dates were taken from @roalkvam2023 and @solheim.
-Dates done on food crusts were excluded due the issue of marine reservoir effects [@nielsen2019, 83].
+Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates.
+Following both from the point made by @freeman concerning the comparison of radiocarbon dates from populations with different , as noted above,  and from the fact that shoreline dating appears less reliable after 4000 BCE and looses its utility c.
+2500 BCE, the Mesolithic was the main focus of this analysis.
+Consequently, calibrated ^14^C-dates with a probability falling later than 2500 BCE were excluded from the analysis.
+Dates falling after 4000 BCE, and the first small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects, but these results should be treated with for the reasons noted above.
 
 All analyses done in this study were performed using the R programming language [@rcoreteam].
-Underlying data and programming code used for the paper is available in an repository at X.
+Underlying data and programming code used for the paper is available in a version-controlled online repository at X.
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [@Marwick2017].
+Analysis of the ^14^C-dates were done using the R package *rcarbon* [@crema2021].
+The R package *shoredate* was used for performing and handling the shoreline dating of sites.
+Finally, the package *ADMUR* [@timpson2021] was used for model comparison and selection, as well as for estimating .
 
@@
 
-Analysis of the ^14^C-dates were done using the R packages *rcarbon* [@crema2021].
-To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned and dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other.
-When calibrated, these are summed and normalised to sum to unity before being included in the final RSPD.
-Individual ^14^C-dates were not normalised.
+To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned using the *binPrep* function from *rcarbon*, where dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other.
+When calibrated, these are summed and divided by the number of dates in the bin before being included in the final RSPD.
+Furthermore, the ^14^C-dates were not normalised.
 This follows from a peculiarity in the calibration procedure which can lead the probability mass of a date to sum to more than one [@weninger2015].
-As a consequence of this, normalising the dates before summing has been shown to create spurious spikes in the RSPD.
+As a consequence of this, normalising the dates before summing has been shown to create spurious spikes in the RSPD [e.g.
+@crema2021].
 All calibration was done using the IntCal20 calibration curve [@reimer2020].
-Edge-effects were handled by only including dates that fall inside the range between xx--2500 BCE with more than 50% probability [cf.
-@timpson2021].
 
-The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013] by fitting two models to the observed RSPD.
-Here this was an exponential model, fit by means of regression, and a uniform model.
-A series of individual calendar years are then drawn from this distribution with replacement, the number of which equals the number of bins in the observed RSPD.
-These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors are added to the date.
+The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD.
+Following @timpson2021, the model parameters were identified by maximum likelihood search, using the differential evolution optimization algorithm DEoptimR.
+For each model, a series of individual calendar years are drawn from the distribution, with replacement, the number of which equals the number of bins in the observed RSPD.
+These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date.
 These are then calibrated back to the calendar scale and finally summed.
 Here, this procedure was repeated 1000 times for each null model.
-The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations were then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
-The degree to which the observed curve deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null model can be rejected.
+The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
+The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.
 
@@
 
-Summing the probability of the shoreline dated sites and the model-fitting procedures follows the same fundamental structure as that used for the radiocarbon dates, and is based on re-purposed programming code from *rcarbon*.
-However, idiosyncrasies in the dating method does necessitate some adjustments.
-First, an illustration of the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate).
+Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon*.
+However, idiosyncrasies in the dating method did necessitate some adjustments.
+To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate).
 Four geologic reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here.
-These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [e.g.
-@svendsen1987].
-The first step in the dating procedure is therefore to interpolate the shoreline displacement to the location to be dated.
+These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987].
+The first step in the dating procedure is to interpolate the shoreline displacement to the location to be dated.
 This is done by inverse distance weighting [e.g.
-@conolly2020], interpolating the sea-level for each calendar year along the geologic shoreline displacement curves based on the distance from the site to be dated to the isobases of the displacement curves.
-The analysis in @roalkvam2023 was based on simulating the distance between sites and shoreline by drawing on independent temporal data on site use as evidenced by ^14^C-dates.
-Sites dated to before 2500 BCE were found to have a likely elevation above sea-level that was reasonably approximated  by a gamma distribution with a decay ratio of 0.168.
-When shoreline dating, this probability is sequentially stepped through at increments of 0.1 m and transferred to the calendar scale by uniformly distributing the probability across calendar years in the range between the lower and upper limit of the interpolated displacement curve.
-Here this is done with resoltuon of 10 years on the calendar scale.
-This procedure gives the the shoreline date of a site.
-The procedure is thus similar to how a radiocarbon date is calibrated, and insights from studies that deal with the numerical treatment of radiocarbon dates and their summation can consequently be useful as analogy for understanding the same properties for the shoreline dates.
+@conolly2020], interpolating the sea-level to the location of a site to be dated based on it distance from the isobases of the displacement curves.
+The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that was reasonably approximated by a gamma distribution with shape =  and scale = .
+When shoreline dating, the gamma distribution is here sequentially stepped through at increments of 0.1m and transferred to the calendar year scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve.
+This procedure gives the the shoreline date of a site [see @roalkvam2023 for details].
+Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf.
+@weninger2015, 545], each shoreline date is normalised to sum to unity.
+To reduce the computational cost of the simulation procedures, the gamma distribution is here stepped through at increments of 0.1m and the calendar scale is kept at a resolution of 5 years.
 
@@
 
-In figure  the shoreline dated sites are plotted according to their elevation.
-However, due to the variable uplift rates within the study region, the same elevation does not equate to the same date and so this does not directly tell us much about their temporal distribution.
-As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date than the same elevation further to the north-east.
-If the elevation of the sites is instead transformed to point-estimate of their date, used by couplng the mean elevation of the site and the mean of the displacement curve interpolated to each site location, the temporal distribution instead looks like the one provided in figure.
-This more resembles how the dates have usually been treated.
-are instead are transformed to point-estimates of shoreline dates for each site  Looking at the distribution across elevations.
-Furthermore, one way that help in understanding the is to see tham as equivalent to uncalibrated radiocarbon dates.
-Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf.
-@weninger2015, 545], each shoreline date is normalised to sum to unity before being summed, as this should not create spurious spikes in the SSPD.
+In Figure  the shoreline dated sites are plotted according to their elevation.
+However, due to the variable uplift rates within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution.
+As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east.
+If the elevation of the sites is instead transformed to point-estimated shoreline dates, done by coupling the mean elevation of the site and the mean of the displacement curve interpolated to each site location, the temporal distribution instead looks like the one provided in figure.
+This follows the lines of how shoreline dates have usually been treated when the frequency distribution of shoreline dated sites has been investigated in the past.
+However, this does not properly account for the uncertainty associated with the dates, and is thus similar to summarising point-estimates of ^14^C-dates.
+This problem is further accentuated by the fact that shoreline dates are associated with more uncertainty than ^14^C-dates.
+Additionally, given that the displacement curves have a more rapid drop early in the Holocene and at locations further to the north-east, this means that this also results in both a geographical and temporal bias in the final results.
+The SSPD resulting from using the probabilistic method of shoreline dating is given in Figure.
 
 In much the same way as characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site.
-As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on the same isobase.
-With the analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve.
+As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase.
+With analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve.
 As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites (Figure \@ref(fig:incpolys)).
@@
 
-```{r incpolys, echo = FALSE, fig.cap = "Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 943) as distributed across the 2km wide line segments that run perpendicular to the shoreline gradient.
-A displacement curve has been interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
+```{r incpolys, echo = FALSE, fig.cap = "A) Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 921) as distributed across the 2km wide line segments that run perpendicular to the shoreline gradient.
+B) The displacement curves interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
 knitr::include_graphics(here("analysis/figures/incpolys.png"))
@@
 Following from the commutative nature of the shoreline dates, the Monte Carlo simulation is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice.
-Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval.
-This is equivalent to the *calsample* sampling method from *rcarbon* [@crema2021].
-The calendar date is then 'uncalibrated'---to follow the RSPD terminology---to an elevation range from which a single elevation value is then drawn with a uniform probability, using intervals of 5cm.
-This elevation value is then shoreline dated using the displacement curve for the relevant 2km interval.
-The dates are then summed, and the process repeated a total of 1000 times.
-Finally, the 97.5% highest and 2.5% lowest summed probability at each calendar year across all simulations is retrieved to create the 95% critical envelope and find the global p-value.
+This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021].
+Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2km interval.
+The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability, using intervals of 5cm.
+As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution of the ^14^C age.
+Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2km interval with the same gamma distribution for all samples.
+Having dated the number of samples equalling the number of shoreline dated sites these were then summed, and the entire process repeated a total of 1000 times.
+The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs.
+This was done using re-purposed code from *rcarbon*.
+
+## Continous piece-wise linear models
+The procedures outlined above represent standard approaches when modelling population dynamics in archaeology.
+These can be helpful for determining whether the population behaviour has followed the trajectories of theoretical models that , and was an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of sites over time could be made.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves -- although this has frequently been the case.
+As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics, published with the accompanying  *ADMUR* R package.
+This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above.
+The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these.
+Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.
+
+## Comparing model performance
+
+For each dating method, the relative performance of the exponential, logistic, uniform and CPL models was compared.
+This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*.
+For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase.
+The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, ], found by estimating the likelihood using the final SPD.
+From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) for each model was then found.
+This penalises the models for the number of parameters in use to avoid over-fitting.
 
@@
 
-```{r spduniform, echo = FALSE, fig.cap = "[Shoreline sims in this version have only been run 50/1000 times] A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model.
+```{r spduniform, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates.
+A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model.
 B) Same as A, but with a exponential model.
@@
 D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
-knitr::include_graphics(here("analysis/figures/spds_uniform.png"))
+knitr::include_graphics(here("analysis/figures/mc.png"))
 ```
+Comparing the the directions and devitions from the standard models, fig indicates some process of overall general increase in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
+However, with global p-values all below X,  all models can be rejected as explaining the data they have been fit to.
+Based on common demographic developments and previous suggestions, there is some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments.
+However, as introduced above, several authors have warned against over-interpreting such deviation.
+Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.
 
-<!-- # Discussion -->
+In the following, some of these deviations will nonetheless be narratively linked to some major developments that have previously been proposed to characterise the Stone Age of south-eastern Norway, with the goal of suggesting some potential.
+It is, however, important to underscore that the *ad-hoc* coupling of the results achieved here with these developments should only be considered suggestions that can offer some potential avenues for further inquiry, for example using the approach outlined by @timpson2021.
+
+# Discussion
+
+It is firmly established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE.
+While the SSPD starts from c.
+, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand.
+Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations, ranging from taphonomic loss, to diverse cultural factors, such as the potential use of quickly deteriorating seal blubber for burning, which could also be related to a possible lack of reliable access to firewood.
+While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump in ^14^C-dates at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind this pattern.
+
+While all the considered models can be rejected as explaining the analysed data, they do nonetheless indicate that some process of decrease in site frequency occurs throughout the period, contrasted by a general increase in the frequency of ^14^C-dated material.
+The most immediate explanation of this mismatch would be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+The modelled process of logistic decay markedly under-predicts the number of sites in the period X--X, and a deviation is also apparent in the period X--X, where the number of observed sites is significantly lower than what would be expected under the logistic model.
+While we are wary of over-interpreting these signals, they strike us as warranting more consideration than the other deviations that are barely outside the critical envelope, or are of a short duration.
+The deviations from after 4000 BCE should also be treated with further care, following both from questions concerning the reliability of shoreline dating after 4000 BCE [cf.
+@roalkvam2023], and the noise in the critical envelope following from the practical implementation of the simulation procedures.
+
+The initial peak in the SSPD would be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
+As evidenced by both genetic data and technological analyses of lithic inventories, this initial phase of human occupation appears to be followed by an influx of people from the .
+If this narrative is coupled to the data observed here, this process would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, while the end of this drop corresponds to the first ^14^C-dates in the RSPD -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is related to cultural processes.
+In the period X--X following this, there is also a marked postive deviation from the RSPD under the exponential model, indicting a population boom, which, although barely breaking the critical envelope of the logitic model, is also evident in the SSPD.
+
+While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is difficult to speculate about.
+However, the positive deviation in the SSPD from the logistic model at c.
+does correspond to a similar deviation of the RSPD from the exponential model.
+Matching negative deviations, on the other hand, can be observed around the middle of the 7th century BCE and in the period after 3500 BCE.
+While we are, as mentioned, cautions in giving these much weight in the SSPD alone, the correspondence of these deviations across the two proxies gives some reason to suggests that these might reflect meaningful demographic signals.
+This would also i
 
 <!-- # Conclusion -->
+The finding that the frequency of radiocarbon dates and frequency of shoreline dated sites, in general, follow inverse trajectories is a valuable insight.
+Although a more precise understanding of this relationship
 

### 2023-02-28 R/03_functions.R

@@ -53,11 +53,12 @@ simulation_summary <- function(spd, simulation_results, cut_off, nsim){
   # Format observed SPD and retrieve the probability
   fSPD <- as.data.frame(spd) %>%
     filter(sum.bce <= cut_off) %>%
-    select(sum.probability)
+    select(sum.probability) %>%
+    unlist(use.names = FALSE)
 
   Zsim <- t(apply(sim, 1, scale))
   zLo <- apply(Zsim, 1, quantile, prob = 0.025, na.rm = TRUE)
-  zHi <- apply(Zsim, 1, quantile, prob = 0.975,na.rm = TRUE)
+  zHi <- apply(Zsim, 1, quantile, prob = 0.975, na.rm = TRUE)
 
   Zscore_empirical <- (fSPD - apply(sim, 1, mean))/apply(sim, 1, sd)
   busts <- which(Zscore_empirical < zLo)

### 2023-02-28 R/04_shoreline_monte_carlo.R

@@ -3,8 +3,11 @@ library(shoredate)
 library(dplyr)
 library(sf)
 library(here)
+library(patchwork)
+library(ADMUR)
+library(DEoptimR)
 
-set.seed(1)
+set.seed(42)
 
 # Source functions
 source(here("R/03_functions.R"))
@@ -45,15 +48,15 @@ fit <- nls(y ~ (exp(a + b * x)),
                              y = sumdatesdf$sum.probability),
            start = list(a = 0, b = 0))
 est <- predict(fit, list(x = sumdatesdf$sum.bce))
-predgrid <- data.frame(bce = sumdatesdf$sum.bce, prob_dens = est)
+expgrid <- data.frame(bce = sumdatesdf$sum.bce, prob_dens = est)
 
 sumdatesdf <- filter(sumdatesdf, sum.bce <= -2500)
-predgrid <- filter(predgrid, bce <= -2500)
+expgrid <- filter(expgrid, bce <= -2500)
 
 # Inspect SPD and fitted model
 ggplot() +
   geom_line(data = sumdatesdf, aes(x = sum.bce, y = sum.probability)) +
-  geom_line(data = predgrid, aes(x = bce, y = prob_dens),
+  geom_line(data = expgrid, aes(x = bce, y = prob_dens),
             linetype = "dashed", colour = "grey40") +
   scale_x_continuous(breaks = seq(-10000, 2500, 1000)) +
   theme_bw()
@@ -63,8 +66,8 @@ random_dates <- data.frame(matrix(nrow = ssize*nsim, ncol = 3))
 names(random_dates) <- c("sample", "simn", "displacement_curve")
 
 
-random_dates$sample = sample(predgrid$bce, replace = TRUE,
-                             size = ssize*nsim, prob = predgrid$prob_dens)
+random_dates$sample = sample(expgrid$bce, replace = TRUE,
+                             size = ssize*nsim, prob = expgrid$prob_dens)
 random_dates$simn <- rep(1:nsim, each = nrow(random_dates)/nsim)
 random_dates$displacement_curve <- incpolys[sample(incpolys$id, replace = TRUE,
                                               size = ssize*nsim,
@@ -124,8 +127,7 @@ simresults <- do.call(rbind.data.frame, results) %>%
   group_by(simn) %>%
   mutate(prob_sum_normalised = prob_sum / sum(prob_sum, na.rm = TRUE))
 
-
-modelsum <- simulation_summary(sumdates, simresults, cut_off = -2500, nsim)
[…]

### 2023-02-28 R/05_radiocarbon.R

@@ -1,98 +1,277 @@
 library(dplyr)
 library(rcarbon)
-
+library(ADMUR)
+library(patchwork)
 
 c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
 c14 <- c14 %>% filter(context != "Food crust")
 
 caldates <- calibrate(x = c14$c14_bp, normalised = FALSE,
                       errors = c14$error, calCurves = "intcal20")
+c14bins <- binPrep(sites =  c14$site_name, ages = c14$c14_bp, h = 200)
 
-c14bins = binPrep(sites = c14$site_name, ages = c14$c14_bp, h = 200)
-c14spd <- spd(caldates, bins = c14bins, timeRange = c(11000, 2000))
+# SPD of entire range in use for fitting models
+c14spd <- spd(caldates, bins = c14bins, timeRange = c(12000, 3500))
 
-nsim <- 10
+# Inspect the impact of binning
+# binsense(x = caldates, y = c14$site_name, h = seq(0,500,100),
+#         timeRange=c(12000,3700))
 
-### Exponential
-expnull <- modelTest(x = caldates, errors = c14$error, bins = c14bins, nsim = nsim,
-                     timeRange = c(12000, 4500), model = "exponential")
+# Fit exponential model
+expfit <- modelTest(x = caldates, errors = c14$error, bins = c14bins,
+                    model = "exponential", timeRange = c(12000, 3500),
+                    fitonly = TRUE)
 
-plot(expnull, calendar = "BCAD")
-
-ggplot(data = expnull$result, aes(x = (calBP-2000)*-1)) +
-  geom_ribbon(aes(ymin = lo, ymax = hi),
-              fill = "grey", alpha = 0.9) +
-  geom_line(data = expnull$fit, aes(x = (calBP-2000)*-1, y = PrDens),
-            linetype = "dashed", colour = "grey40") +
-  labs(x = "BCE", y = "Summed probability") +
-  scale_x_continuous(breaks = seq(-10000, 2000, 1000),
-                     limits = c(-10000, -2500)) +
-  geom_line(aes(y = PrDens)) +
-  theme_bw()
-
-### Logistic
-
+# Fit logistic model
 grd_c14 <- c14spd$grid %>% filter(PrDens > 0)
 x <- grd_c14$calBP
 y <- grd_c14$PrDens
 logfit <- nls(y ~ SSlogis(x, Asym, xmid, scal),
[…]

### 2023-02-28 R/02_displacement_prep.R
Re-did incoloys plot

@@ -128,7 +128,7 @@ displt <- ggplot(dispcurves) +
                                     group = id,
                                     alpha = dens)) +
   labs(title = "B", x = "BCE/CE", y = "Meters above present sea-level") +
-  geom_vline(xintercept = -9465) +
+  scale_x_continuous(expand = c(0,0), limits = c(-9469, 2000)) +
   theme_bw() +
   theme(legend.position = "none")
 

### 2023-04-21 analysis/paper/paper.Rmd
ADMUR for SSPD

@@
 library(sf)
+library(shoredate)
 ```
@@
 surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
+sites <- st_read(here("analysis/data/derived_data/combined_sites.gpkg"))
 ```
@@
 The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g.
-@crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018a; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021].
+@crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018a; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021, see also @persson2009].
 The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach.
@@
 What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021].
-This paper reports on and beings to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).
+This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).
 
@@
 As coastal foragers appear to have predominantly settled on or close to the contemporaneous shoreline, this can be utilised to assign an approximate date to the sites.
-This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement---a method known as shoreline dating.
+This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating.
 This is not least useful for dating the large number of surveyed sites in the region where other temporal data that often follow with an excavation are not available, such as radiocarbon dates or typological indicators in artefact inventories.
@@
 The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility.
-Thus, the geographical and temporal limits outlined above are also used for this study.
+Thus, this geographical and temporal limit are also used for this study.
 
@@
 As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model.
-It is the combined representation of a range of possible explanations for the data---the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021].
+It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021].
 This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021].
@@
 However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves.
-As a result, by improving on @goldstein, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics, published with the accompanying R package  *ADMUR*.
+As a result, and by improving on @goldstein, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics based RSPDs, published with the accompanying R package  *ADMUR*.
 This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally testing the best model using the Monte Carlo approach outlined above.
-The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these.
-One major benefit of the approach is thus that it provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
+The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these.
+The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
 Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.
@@
 Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded.
-This marks the latest start-date among the employed displacement curves, and no sites are yet known to be older than around 9300 BCE [@glorstad2016].
+This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016].
 Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2022.
@@
 Consequently, calibrated ^14^C-dates with a probability falling later than 2500 BCE were excluded from the analysis.
-Dates falling after 4000 BCE, and the first small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects, but these results should be treated with for the reasons noted above.
+Dates falling after 4000 BCE and the first small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.
 
@@
 
-```{r map, echo = FALSE, fig.cap = "Map of the study area and analysed sites.
+```{r map, echo = FALSE, fig.cap = "Map of the study area and shoreline dated sites.
 Black lines indicate the borders between municipalities.
-The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1.
-Of the excavated sites, 102 have been dated by means of shoreline dating.", out.width = "500px", fig.align="center", regfloat = TRUE}
+The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1.", out.width = "500px", fig.align="center", regfloat = TRUE}
+
+# Map prepared in 01_shoreline_spd.R 
 knitr::include_graphics(here("analysis/figures/map.png"))
@@
 
-To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned using the *binPrep* function from *rcarbon*, where dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other.
+To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned using the *binPrep()* function from *rcarbon*, where dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other.
 When calibrated, these are summed and divided by the number of dates in the bin before being included in the final RSPD.
@@
 The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD.
-Following @timpson2021, the model parameters were identified by maximum likelihood search, using the differential evolution optimization algorithm DEoptimR.
+Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*.
+The Monte Carlo simulations were then performed using the *modelTest()* function from *rcarbon*.
 For each model, a series of individual calendar years are drawn from the distribution, with replacement, the number of which equals the number of bins in the observed RSPD.
@@
 These are then calibrated back to the calendar scale and finally summed.
-Here, this procedure was repeated 1000 times for each null model.
+Here, this procedure was repeated 10000 times for each null model.
 The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
@@
 @weninger2015, 545], each shoreline date is normalised to sum to unity.
-To reduce the computational cost of the simulation procedures, the gamma distribution is here stepped through at increments of 0.1m and the calendar scale is kept at a resolution of 5 years.
+To reduce the computational cost of the simulation procedures, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.
 
@@
 ", out.width = "500px", fig.align="center", regfloat = TRUE}
-knitr::include_graphics(here("analysis/figures/shoredate.png"))
+target_site <- sites[1,]
+target_plot(target_site)
+
 ```
@@
 As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east.
-If the elevation of the sites is instead transformed to point-estimated shoreline dates, done by coupling the mean elevation of the site and the mean of the displacement curve interpolated to each site location, the temporal distribution instead looks like the one provided in figure.
-This follows the lines of how shoreline dates have usually been treated when the frequency distribution of shoreline dated sites has been investigated in the past.
-However, this does not properly account for the uncertainty associated with the dates, and is thus similar to summarising point-estimates of ^14^C-dates.
-This problem is further accentuated by the fact that shoreline dates are associated with more uncertainty than ^14^C-dates.
-Additionally, given that the displacement curves have a more rapid drop early in the Holocene and at locations further to the north-east, this means that this also results in both a geographical and temporal bias in the final results.
-The SSPD resulting from using the probabilistic method of shoreline dating is given in Figure.
+If the elevation of the sites is instead transformed to point-estimated shoreline dates by finding the mean shoreline date for the sites, without accounting for the likely distance to the shoreline, the temporal distribution instead looks like the one provided in Figure.
+This is similar to how shoreline dates have been treated when the frequency distribution of shoreline dated sites has been investigated in the past [@breivik; @solheim; @fossum; @mjærum; @tallavaara].
+However, apart from not accounting for the distance between the site and the sea, this does not properly account for the uncertainty associated with the dates, and is thus similar to summarising point-estimates of ^14^C-dates.
+This problem is further accentuated by the fact that shoreline dating, as implemented here, is associated with more uncertainty than ^14^C-dates.
+Additionally, given that the displacement curves have a more rapid drop early in the Holocene and at locations further to the north-east in the study area, this means that this also results in both a geographical and temporal bias in the final results.
+The SSPD resulting from using the probabilistic method of shoreline dating is given in Figure, which clearly demonstrates that accounting for these dimensions gives a different result compared to the traditional approach.
 
-In much the same way as characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site.
+However, in the same way is with RSPDs, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above.
+Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site.
 As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase.
 With analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve.
-As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites (Figure \@ref(fig:incpolys)).
-These intervals were then assigned a weight based on how the density of observed sites is distributed among them.
+As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites.
+These intervals were then assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).
 
@@
 Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2km interval.
-The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability, using intervals of 5cm.
-As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution of the ^14^C age.
-Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2km interval with the same gamma distribution for all samples.
-Having dated the number of samples equalling the number of shoreline dated sites these were then summed, and the entire process repeated a total of 1000 times.
+The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability, using intervals of 5 cm.
+As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age.
+Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples.
+Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times.
 The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs.
@@
 The procedures outlined above represent standard approaches when modelling population dynamics in archaeology.
-These can be helpful for determining whether the population behaviour has followed the trajectories of theoretical models that , and was an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of sites over time could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves -- although this has frequently been the case.
-As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics, published with the accompanying  *ADMUR* R package.
+These can be helpful for determining whether the population behaviour has followed the trajectories of theoretical models that , and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of sites over time could be made.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case.
+As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics.
 This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above.
@@
 ```
-Comparing the the directions and devitions from the standard models, fig indicates some process of overall general increase in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
-However, with global p-values all below X,  all models can be rejected as explaining the data they have been fit to.
-Based on common demographic developments and previous suggestions, there is some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments.
+Comparing the the directions and devitions from the three standard models, Figure appears to indicate some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
+However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD.
+Based on common demographic developments and previous suggestions, there is possibly some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments.
 However, as introduced above, several authors have warned against over-interpreting such deviation.
@@
 , thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand.
-Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations, ranging from taphonomic loss, to diverse cultural factors, such as the potential use of quickly deteriorating seal blubber for burning, which could also be related to a possible lack of reliable access to firewood.
+Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations, ranging from taphonomic loss, to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a possible lack of reliable access to firewood.
 While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump in ^14^C-dates at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind this pattern.
@@
 In the period X--X following this, there is also a marked postive deviation from the RSPD under the exponential model, indicting a population boom, which, although barely breaking the critical envelope of the logitic model, is also evident in the SSPD.
+
+@berg-hansen2022 have argued that there is an increase in sites in the first half of the MM in the northern part of the study area of this paper.
+However, their analysis is seemingly also based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
+Apart from the limitations with using point-based estimates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
+This analysis instead indicates a drop in the number of sites through this time-interval
 

### 2023-04-21 R/01_shoreline_spd.R

@@ -3,6 +3,7 @@ library(sf)
 library(terra)
 library(here)
 library(shoredate)
+library(patchwork)
 
 # For reproducibility
 set.seed(1)
@@ -55,6 +56,19 @@ sites <- rbind(rcolexcshore, rcolsurvq)
 ###### Shoreline dating ######
 step_reso = 0.1
 
+# Find weighted means dates for figure X in text
+wmean_dates <- shoreline_date(sites, elevation = dtm,
+                              elev_reso = step_reso,
+                              cal_reso = 5,
+                              model = "none",
+                              verbose = TRUE)
+
+wmeans <- c()
+for(i in 1:length(wmean_dates)){
+  wmeans <- c(wmeans, wmean_dates[[i]][[1]]$weighted_mean)
+}
+wmeans <- wmeans[wmeans  <= -2500]
+
 shorelinedates <- shoreline_date(sites,
                                  elevation = dtm,
                                  elev_reso = step_reso,
@@ -63,8 +77,14 @@ shorelinedates <- shoreline_date(sites,
                                  verbose = TRUE)
 
 
-save(step_reso, shorelinedates,
+save(step_reso, wmeans, shorelinedates,
      file = here("analysis/data/derived_data/sdates.RData"))
+
+# Sum the probability distributions for the shoreline dates
+sumdates <- sum_shoredates(shorelinedates)
+sumdatesdf <- as.data.frame(sumdates) %>%
+  filter(sum.probability != 0)
+sumdatesdf <- filter(sumdatesdf, sum.bce <= -2500)
 
 # Exclude sites that were given a NA date (elevation equates to a date older
 # than 9465 BCE or younger than 2500 BCE).
@@ -78,6 +98,139 @@ out_of_range <- c("159969", "94301-1", "144111-1", "144113-1", "144114-1",
 
 combined_sites <- filter(sites, !ask_id %in% out_of_range)
 
+wplot <- ggplot() +
+  geom_histogram(aes(x = wmeans, y = ..density..),
+                 binwidth = 200,
+                 col = "black", fill = "grey", alpha = 0.7) +
+  geom_density(aes(x = wmeans),
+               col = "black") +
+  scale_x_continuous(breaks = seq(-10000, 2500, 1000)) +
+  labs(title = "Mean shoreline date", x = "BCE", y = "Density") +
+  theme_bw()
+
+splot <- ggplot() +
[…]

### 2023-04-21 R/03_functions.R

@@ -40,6 +40,8 @@ reverse_shoredate <- function(shoreline_date,
 }
 
 
+# Reverse a single calendar date to a corresponding elevation
+# Repurposed from rcarbon
 simulation_summary <- function(spd, simulation_results, cut_off, nsim){
 
   # Reformat data frame of normalised simulated SPDs
@@ -53,6 +55,7 @@ simulation_summary <- function(spd, simulation_results, cut_off, nsim){
   # Format observed SPD and retrieve the probability
   fSPD <- as.data.frame(spd) %>%
     filter(sum.bce <= cut_off) %>%
+    mutate(sum.probability = sum.probability/(sum(sum.probability) * 5)) %>%
     select(sum.probability) %>%
     unlist(use.names = FALSE)
 

### 2023-04-21 R/04_shoreline_monte_carlo.R

@@ -467,8 +467,9 @@ cpl_8 <- JDEoptim(lower = rep(0, 15), upper = rep(1, 15), fn = objectiveFunction
                   PDarray = pd, type = 'CPL', NP = 260, maxiter = 1000 * 15,
                   trace = TRUE)
 save(cpl_8, file = here("analysis/data/derived_data/cpl_8.RData"))
+load(here("analysis/data/derived_data/cpl_8.RData"))
 
-save(cpl_1, cpl_2, cpl_3, cpl_4, cpl_5, cpl_6, cpl_7, exp, logi,
+save(cpl_1, cpl_2, cpl_3, cpl_4, cpl_5, cpl_6, cpl_7, cpl_8, exp, logi,
      file = here("analysis/data/derived_data/shoremodels.RData"))
 load(here("analysis/data/derived_data/shoremodels.RData"))
 
@@ -484,7 +485,8 @@ shore_lik <- c("Exponential" = -exp$value,
              "CPL-4" = -cpl_4$value,
              "CPL-5" = -cpl_5$value,
              "CPL-6" = -cpl_6$value,
-             "CPL-7" = -cpl_7$value)
+             "CPL-7" = -cpl_7$value,
+             "CPL-8" = -cpl_8$value)
 
 shore_bic <- c(log(ssize)*1 - 2*shore_lik[1],
                log(ssize)*2 - 2*shore_lik[2],
@@ -494,7 +496,8 @@ shore_bic <- c(log(ssize)*1 - 2*shore_lik[1],
                log(ssize)*7 - 2*shore_lik[6],
                log(ssize)*9 - 2*shore_lik[7],
                log(ssize)*11 - 2*shore_lik[8],
-               log(ssize)*13 - 2*shore_lik[9])
+               log(ssize)*13 - 2*shore_lik[9],
+               log(ssize)*15 - 2*shore_lik[10])
 
 shore_aic <- c(2*1 - 2*shore_lik[1],
                2*2 - 2*shore_lik[2],
@@ -525,18 +528,28 @@ cpl6 <- convertPars(pars = cpl_6$par, years = -9445:-2500, type='CPL')
 cpl6$model <- "CPL-6"
 cpl7 <- convertPars(pars = cpl_7$par, years = -9445:-2500, type='CPL')
 cpl7$model <- "CPL-7"
+cpl8 <- convertPars(pars = cpl_8$par, years = -9445:-2500, type='CPL')
+cpl8$model <- "CPL-8"
 
[…]

### 2023-04-21 R/05_radiocarbon.R

@@ -2,6 +2,7 @@ library(dplyr)
 library(rcarbon)
 library(ADMUR)
 library(patchwork)
+library(ggplot2)
 
 c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
 c14 <- c14 %>% filter(context != "Food crust")
@@ -200,7 +201,6 @@ x <- subcal2
 binNames <- unique(subbins2)
 
 calyears <- data.frame(calBP = seq(12000, 4500, -1))
-caldateyears <- seq(caldateTR[1],caldateTR[2],-1)
 binnedMatrix <- matrix(NA, nrow= nrow(calyears), ncol = length(binNames))
 colnames(binnedMatrix) <- binNames
 rownames(binnedMatrix) <- rev(calyears[,1])
@@ -226,26 +226,67 @@ for(b in 1:length(binNames)){
 
 pd <- as.data.frame(binnedMatrix)
 
-# Rename model columns to match ADMUR syntax
-expmodel <- expnull$fit %>%
-  rename("year" = "calBP", "pdf" = "PrDens") %>%
-  filter(year < 10554) %>%
-  select(year, pdf)
-
-logmodel <- lognull$fit %>%
-  rename("year" = "calBP", "pdf" = "PrDens") %>%
-  select(year, pdf)
-
-unimodel <- uninull$fit %>%
-  rename("year" = "calBP", "pdf" = "PrDens") %>%
-  filter(year < 10554) %>%
-  select(year, pdf)
+# # Rename model columns to match ADMUR syntax
+# expmodel <- expnull$fit %>%
+#   rename("year" = "calBP", "pdf" = "PrDens") %>%
+#   filter(year < 10554) %>%
+#   select(year, pdf)
+#
+# logmodel <- lognull$fit %>%
+#   rename("year" = "calBP", "pdf" = "PrDens") %>%
+#   select(year, pdf)
+#
+# unimodel <- uninull$fit %>%
+#   rename("year" = "calBP", "pdf" = "PrDens") %>%
+#   filter(year < 10554) %>%
+#   select(year, pdf)
 
 # Make year range in pd match that of the models
 pd <- pd[as.numeric(rownames(pd)) <= max(expmodel$year),]
-
+plotPD(pd)
 # Check that SPD and models matches plt4-plt6 , above
 SPD <- as.data.frame(rowSums(pd))
+plotPD(SPD/(sum(SPD) * 5))
+
[…]

### 2023-04-21 R/0x_shoreline_monte_carlo2.R

@@ -1,0 +1,407 @@
+library(ggplot2)
+library(shoredate)
+library(dplyr)
+library(sf)
+library(here)
+library(patchwork)
+library(ADMUR)
+library(DEoptimR)
+
+set.seed(42)
+
+# Source functions
+source(here("R/03_functions.R"))
+
+# Load combined sites
+sites <- st_read(here("analysis/data/derived_data/combined_sites.gpkg"))
+
+# Load polygons with displacement curves
+load(here("analysis/data/derived_data/displacement_polygons.RData"))
+
+# Load shoreline dates
+load(here("analysis/data/derived_data/sdates.RData"))
+
+dates_dfs <- lapply(shorelinedates, as.data.frame)
+
+# Combine these into a single data frame
+sdates <- do.call(rbind, dates_dfs)
+
+# Assign simulation results for each site a unique ID
+sdates$site <- rep(1:length(dates_dfs) , each = nrow(sdates)/length(dates_dfs))
+
+pd <- as.data.frame(sdates) %>%
+  dplyr::filter(bce <= -2500 & bce >= -9445) %>%
+  dplyr::group_by(bce) %>%
+  dplyr::filter(!is.na(probability)) %>%
+  dplyr::group_by(site) %>%
+  tidyr::pivot_wider(names_from = site, values_from = probability) %>%
+  tibble::column_to_rownames("bce")
+
+pd <- sweep(pd, 2, colSums(pd),"/")
+
+SPD <- as.data.frame(rowSums(pd))
+SPD <- SPD/( sum(SPD) *5 )
+
+# Maximum likelihood search using DEoptimR
+exp <- JDEoptim(lower = -0.01, upper = 0.01, fn = objectiveFunction,
+                PDarray = pd, type = "exp", NP = 20)
+logi <- JDEoptim(lower = c(0, 0000), upper = c(1, 10000),
+                 fn = objectiveFunction, PDarray = pd, type = 'logistic',
+                 NP = 40, trace = TRUE)
+# No search required for uniform
+unif <- -objectiveFunction(pars = NULL, PDarray = pd, type = 'uniform')
+
+expp <- convertPars(pars = exp$par, years = -9445:-2500, type = 'exp')
+expp$model <- "Exponential"
+logip <- convertPars(pars = logi$par, years = -9445:-2500, type = 'logistic')
+logip$model <- "Logistic"
[…]

### 2023-04-21 R/0x_shoreline_monte_carlo2.R
Monte Carlo2

@@ -404,4 +404,9 @@ uniplt <- ggplot() +
                      limits = c(-10000, -2500),
                      expand = expansion(mult = c(0, 0))) +
   theme_bw()
-uniplt
+
+
+expplt + logiplt + uniplt
+
+ggsave(here::here("analysis/figures/shoreline_mc.png"),
+       units = "px", width = 4000, height = 1250)

### 2023-04-24 analysis/paper/paper.Rmd
ADMUR

@@
 ```
-Comparing the the directions and devitions from the three standard models, Figure appears to indicate some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
+Comparing the the directions and deviations from the three standard models, Figure appears to indicate some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
 However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD.
@@
 Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.
+
+For the RSPD the rejected logistic model nonetheless offers the best approximation of the data.
+If taken as an indication of demographic developments this would indicate a first migration of people that is more substantial that what would be expected under the logistic null.
+Further deviations are then evident with the negative deviation and the poistive deviation c.
+These are, however, brief and can be left as little more than suggestions for possible meaningful demographic events.
 

### 2023-04-24 R/05_radiocarbon.R

@@ -3,6 +3,7 @@ library(rcarbon)
 library(ADMUR)
 library(patchwork)
 library(ggplot2)
+library(DEoptimR)
 
 c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
 c14 <- c14 %>% filter(context != "Food crust")

### 2023-04-24 R/0x_radiocarbon_cpl.R

@@ -1,0 +1,101 @@
+library(ADMUR)
+library(DEoptimR)
+
+set.seed(42)
+
+# Load data from radiocarbon_mc.R: models fit for MC, the pd and the min and
+# max ages
+load(file = here("analysis/data/derived_data/rcarbon_pd_models.RData"))
+
+# Fit CPL-models to radiocarbon data
+cpl_1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction, PDarray = pd,
+                  type = 'CPL', NP = 20, trace = TRUE)
+cpl_2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 60,  maxiter = 400 * 3,
+                  trace = TRUE)
+cpl_3 <- JDEoptim(lower = rep(0, 5), upper = rep(1, 5), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 100, maxiter = 400 * 5,
+                  trace = TRUE)
+cpl_4 <- JDEoptim(lower = rep(0, 7), upper = rep(1, 7), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 140, maxiter = 400 * 7,
+                  trace = TRUE)
+cpl_5 <- JDEoptim(lower = rep(0, 9), upper = rep(1, 9), fn = objectiveFunction,
+                  PDarray = pd, type='CPL', NP = 180, maxiter = 400 * 9,
+                  trace = TRUE)
+cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 220, maxiter =  400 * 11,
+                  trace = TRUE)
+
+save(exp, logi, cpl_1, cpl_2, cpl_3, cpl_4, cpl_5,
+          file = here("analysis/data/derived_data/rcarbon_models.RData"))
+
+expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
+expp$model <- "Exponential"
+logip <- convertPars(pars = logi$par, years = minage:maxage, type = 'logistic')
+logip$model <- "Logistic"
[…]

### 2023-04-24 R/0x_radiocarbon_mc2.R

@@ -1,0 +1,253 @@
+library(dplyr)
+library(rcarbon)
+library(ADMUR)
+library(patchwork)
+library(ggplot2)
+library(DEoptimR)
+
+set.seed(42)
+
+c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
+c14 <- c14 %>% filter(context != "Food crust")
+
+caldates <- calibrate(x = c14$c14_bp, normalised = FALSE,
+                      errors = c14$error, calCurves = "intcal20")
+c14bins <- binPrep(sites =  c14$site_name, ages = c14$c14_bp, h = 200)
+
+# SPD of entire range in use for fitting models
+c14spd <- spd(caldates, bins = c14bins, timeRange = c(12000, 3500))
+
+
+# Using rcarbon, calibrate each date individually or sum dates that are binned
+subcal2 <- subset(caldates, BP <= 12000 & BP >= 4500, p = 0.05)
+c14sub2 <- c14[which.CalDates(caldates, BP <= 12000 &
+                                BP >= 4500, p = 0.05),]
+subbins2 = binPrep(sites =  c14sub2$site_name, ages = c14sub2$c14_bp, h = 200)
+
+# Retrieve binned and unbinned dates (based on code from rcarbon::spd())
+x <- subcal2
+binNames <- unique(subbins2)
+
+calyears <- data.frame(calBP = seq(12000, 4500, -1))
+binnedMatrix <- matrix(NA, nrow= nrow(calyears), ncol = length(binNames))
+colnames(binnedMatrix) <- binNames
+rownames(binnedMatrix) <- rev(calyears[,1])
+
+for(b in 1:length(binNames)){
+  index <- which(subbins2 == binNames[b])
+  slist <- x$grids[index]
+  slist <- lapply(slist,FUN = function(x) merge(calyears, x, all.x = TRUE))
+  slist <- rapply(slist, f = function(x) ifelse(is.na(x), 0, x),
+                  how = "replace")
+  slist <- lapply(slist, FUN = function(x) x[with(x, order(-calBP)), ])
+
+  tmp <- do.call(rbind, slist)
+  tmp <- aggregate(tmp$PrDens, by = list(Category = tmp$calBP), FUN = sum)
+  spdtmp <- tmp[,-1]
+  spdtmp <- as.data.frame(spdtmp)
+  rownames(spdtmp) <- tmp[, 1]
+  names(spdtmp) <- "PrDens"
+
+  # tmp <- lapply(slist,`[`,2)
[…]

### 2023-06-11 analysis/paper/paper.Rmd
Monte Carlo simulation, shoreline dates

@@
 ---
-title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Stone Age on the Norwegian Skagerrak coast"
+title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
 # author: 
@@
       - --lua-filter=../templates/pagebreak.lua
+header-includes:
+  \usepackage{placeins}
 bibliography: references.bib
@@
 library(dplyr)
+library(ggplot2)
+library(patchwork)
 library(kableExtra)
@@
 # Introduction
-Population size is regarded as one of the primary drivers of cultural variation, and is of critical importance to our understanding of past human societies [@shennan2000].
+Population size is regarded as an important driver of cultural variation, and is of critical importance to our understanding of past human societies [e.g.
+@french2021; @shennan2000].
 The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g.
-@crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018a; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021, see also @persson2009].
+@crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].
 The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach.
@@
 Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017].
-However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted and possibly weighted differently according some criteria.
+However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
 Before any such fluctuations are given any substantive interpretation, however, there are a host of methodological issues that have to be considered.
@@
 The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure.
-This is because the summed probabilities can no longer been seen as probabilities, but rather represents the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022].
+This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022].
 As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model.
@@
 These null-models are typically a uniform, exponential or logistic distribution.
-These are chosen *a priori* with reference to common long-term population dynamics and are parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE).
+These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE).
 The result from these simulations are then used to create a 95% critical envelope representing the null model.
-The proportion of the observed RSPD that falls outside this envelope is used to estimate a global p-value indicating whether the null model can be rejected.
+The proportion of the observed RSPD that falls outside this envelope is then used to estimate a global p-value indicating whether the null model can be rejected.
 In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model.
@@
 The procedure outlined above represent the standard approach when modelling population dynamics in archaeology.
-This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of sites over time could be made.
+This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of shoreline dated sites over time could be made.
 However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves.
-As a result, and by improving on @goldstein, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics based RSPDs, published with the accompanying R package  *ADMUR*.
-This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally testing the best model using the Monte Carlo approach outlined above.
-The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these.
+As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*.
+This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test in line with the Monte Carlo approach outlined above.
+The benefit of the CPL models is that these define discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these.
 The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
@@
 This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016].
-Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2022.
+Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019.
 Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis.
@@
 Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates.
-Following both from the point made by @freeman concerning the comparison of radiocarbon dates from populations with different , as noted above,  and from the fact that shoreline dating appears less reliable after 4000 BCE and looses its utility c.
-2500 BCE, the Mesolithic was the main focus of this analysis.
-Consequently, calibrated ^14^C-dates with a probability falling later than 2500 BCE were excluded from the analysis.
-Dates falling after 4000 BCE and the first small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.
+Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c.
+2500 BCE, the Mesolithic is the main focus of this analysis.
+Consequently, probabilities falling later than 2500 BCE were excluded from the analysis.
+Probabilities falling after 4000 BCE and the first, possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.
 
@@
 Underlying data and programming code used for the paper is available in a version-controlled online repository at X.
-This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [@Marwick2017].
+This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
 Analysis of the ^14^C-dates were done using the R package *rcarbon* [@crema2021].
-The R package *shoredate* was used for performing and handling the shoreline dating of sites.
-Finally, the package *ADMUR* [@timpson2021] was used for model comparison and selection, as well as for estimating .
+The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b].
+Finally, the package *ADMUR* [@timpson2021] was used for model comparison.
 
@@
 # Map prepared in 01_shoreline_spd.R 
-knitr::include_graphics(here("analysis/figures/map.png"))
+knitr::include_graphics(here("analysis/figures/map1.png"))
 ```
@@
 The Monte Carlo simulations were then performed using the *modelTest()* function from *rcarbon*.
-For each model, a series of individual calendar years are drawn from the distribution, with replacement, the number of which equals the number of bins in the observed RSPD.
+For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD.
 These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date.
@@
 However, idiosyncrasies in the dating method did necessitate some adjustments.
-To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate).
-Four geologic reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here.
+To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate) and outlined below.
+
+Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B).
 These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987].
@@
 This is done by inverse distance weighting [e.g.
-@conolly2020], interpolating the sea-level to the location of a site to be dated based on it distance from the isobases of the displacement curves.
-The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that was reasonably approximated by a gamma distribution with shape =  and scale = .
-When shoreline dating, the gamma distribution is here sequentially stepped through at increments of 0.1m and transferred to the calendar year scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve.
-This procedure gives the the shoreline date of a site [see @roalkvam2023 for details].
+@conolly2020], interpolating the relative sea-level change to the location of a site to be dated, based on it's distance from the isobases of the displacement curves.
+
+```{r tpq_ranges, include = FALSE}
+# Load data from mean_tpq_ranges to be used in paragraph below
+load(here("analysis/data/derived_data/disp_tpq_ranges.RData"))
+```
+
+With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline.
+Figure \@ref(fig:shoredate)C indicates where the elevation of the example site intersects the interpolated displacement curve.
+As it can be reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
+However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbritrary error of $\pm$ 100 years to this point-estimate.
+However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use, limits, as with TPQ dates in general, the further inferential steps that can be taken.
+Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
+This last point can be illustrated by finding the range between the upper and the lower limit of the four geological displacement curves for every meter from the marine limit and down to elevations that do not give an earliest date younger than 2500 BCE.
+This gives a mean TPQ range of `r round(tpqmean)` years with a standard deviation of `r round(tpqsd)` years across all four displacement curves, with this range varying significantly over time and between curves.
+
+The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833.
+This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D).
+When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve.
+This procedure thus gives the the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details].
 Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf.
 @weninger2015, 545], each shoreline date is normalised to sum to unity.
-To reduce the computational cost of the simulation procedures, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.
+To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.
 
-```{r shoredate, echo = FALSE, fig.cap = "A) Example location relative to the isobases of the displacement curves.
-B) The geologic displacement curves and the curve interpolated to the example location.
-C) Resulting shoreline date in light grey on the x-axis.
-The black line underneath marks the 95\\% HDR.
-The dashed line marks the elevation of the example location.
-The exponential function on the y-axis decays with ratio $\\lambda$ and represents the likely elevation of a site above sea-level when it was in use.
-", out.width = "500px", fig.align="center", regfloat = TRUE}
-target_site <- sites[1,]
-target_plot(target_site)
+```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.
+B) The geological displacement curves in the region and the curve interpolated to the example site.
+C) The black line on the y-axis represents the elevation of the site.
+This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date.
+D) The shoreline dating approach used for this study, where the gamma distrbution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
 
+knitr::include_graphics(here("analysis/figures/example_shoredate.png"))
 ```
 
-<!-- The SSPD was subjected to the same procedure for CPL model fitting and comparison as teh RSPD above. Dates with 50% probability distribution  -->
+The implications of different approaches to shoreline dating and summing the results is further illustrated in Figure \@ref(fig:sumshore).
+In Figure \@ref(fig:sumshore)A  the sites to be shoreline dated are plotted according to their elevation.
+Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution.
+For example, as the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east.
 
-In Figure  the shoreline dated sites are plotted according to their elevation.
-However, due to the variable uplift rates within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution.
-As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east.
-If the elevation of the sites is instead transformed to point-estimated shoreline dates by finding the mean shoreline date for the sites, without accounting for the likely distance to the shoreline, the temporal distribution instead looks like the one provided in Figure.
-This is similar to how shoreline dates have been treated when the frequency distribution of shoreline dated sites has been investigated in the past [@breivik; @solheim; @fossum; @mjærum; @tallavaara].
-However, apart from not accounting for the distance between the site and the sea, this does not properly account for the uncertainty associated with the dates, and is thus similar to summarising point-estimates of ^14^C-dates.
-This problem is further accentuated by the fact that shoreline dating, as implemented here, is associated with more uncertainty than ^14^C-dates.
-Additionally, given that the displacement curves have a more rapid drop early in the Holocene and at locations further to the north-east in the study area, this means that this also results in both a geographical and temporal bias in the final results.
-The SSPD resulting from using the probabilistic method of shoreline dating is given in Figure, which clearly demonstrates that accounting for these dimensions gives a different result compared to the traditional approach.
+When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
+These point-estimates are then aggregated in bins of 200 or 500 years [e.g.
+@breivik2014; @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020].
+This approach is illustrated in Figure \@ref(fig:sumshore)B.
+The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, clearly demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c.
+8000 BCE.
+
+```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by elevation.
+The binwidth is 4 m.
+B) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates.
+C) The result of instead summing the probability of the shoreline dates achievied with the method employed here, which accounts for the likely distance between sites and shoreline.
+The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}
+
+knitr::include_graphics(here("analysis/figures/summed_examples.png"))
+```
 
@@
 
-```{r incpolys, echo = FALSE, fig.cap = "A) Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 921) as distributed across the 2km wide line segments that run perpendicular to the shoreline gradient.
+```{r incpolys, echo = FALSE, fig.cap = "A) Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 921) as distributed across 2 km wide line segments that run perpendicular to the shoreline gradient.
 B) The displacement curves interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
@@
 
-Following from the commutative nature of the shoreline dates, the Monte Carlo simulation is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice.
+Following from the commutative nature of the shoreline dates dealt with here, the Monte Carlo simulation with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice.
 This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021].
-Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2km interval.
-The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability, using intervals of 5 cm.
+Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval.
+The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm.
 As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age.
-Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples.
+Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was then shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples.
 Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times.
@@
 The procedures outlined above represent standard approaches when modelling population dynamics in archaeology.
-These can be helpful for determining whether the population behaviour has followed the trajectories of theoretical models that , and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of sites over time could be made.
+These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shorebound sites over time could be made.
 However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case.
@@
 For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase.
-The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, ], found by estimating the likelihood using the final SPD.
+The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021], found by estimating the likelihood using the final SPD.
 From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) for each model was then found.
@@
 
-For the RSPD the rejected logistic model nonetheless offers the best approximation of the data.
+For the RSPD the rejected logistic model offers the best approximation of the data of the compared models.
 If taken as an indication of demographic developments this would indicate a first migration of people that is more substantial that what would be expected under the logistic null.
@@
 While the SSPD starts from c.
-, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand.
-Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations, ranging from taphonomic loss, to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a possible lack of reliable access to firewood.
+9450 BCE, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand.
+Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations, ranging from taphonomic loss, to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a lack of reliable access to firewood.
 While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump in ^14^C-dates at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind this pattern.
 
-While all the considered models can be rejected as explaining the analysed data, they do nonetheless indicate that some process of decrease in site frequency occurs throughout the period, contrasted by a general increase in the frequency of ^14^C-dated material.
+While all the considered models can be rejected as explaining the analysed data, they do nonetheless indicate that some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
 The most immediate explanation of this mismatch would be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
-The modelled process of logistic decay markedly under-predicts the number of sites in the period X--X, and a deviation is also apparent in the period X--X, where the number of observed sites is significantly lower than what would be expected under the logistic model.
+The modelled process of logistic decay markedly under-predicts the SSPD in the period X--X, and a deviation is also apparent in the period X--X, where the number of observed sites is significantly lower than what would be expected under the logistic model.
 While we are wary of over-interpreting these signals, they strike us as warranting more consideration than the other deviations that are barely outside the critical envelope, or are of a short duration.
 The deviations from after 4000 BCE should also be treated with further care, following both from questions concerning the reliability of shoreline dating after 4000 BCE [cf.
-@roalkvam2023], and the noise in the critical envelope following from the practical implementation of the simulation procedures.
+@roalkvam2023].
 
@@
 Apart from the limitations with using point-based estimates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
-This analysis instead indicates a drop in the number of sites through this time-interval
-
+The analysis undertaken here instead indicates a drop in the number of sites through this time-interval.
 While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is difficult to speculate about.
@@
 
-<!-- # Conclusion -->
+While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
+Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
+@riede2012; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
+@roalkvam2022].
+Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of importance to unpack their relationship [e.g.
+@ordonez2022; @bergsvik2022; @jorgensen2020].
+
+# Conclusion
 The finding that the frequency of radiocarbon dates and frequency of shoreline dated sites, in general, follow inverse trajectories is a valuable insight.
 Although a more precise understanding of this relationship
+
+Furthermore, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application [see @roalkvam2023].
+While we still believe the approach represents a considerable improvement over previous approaches, the results should
 

### 2023-06-11 R/01_shoreline_spd.R

@@ -80,6 +80,8 @@ shorelinedates <- shoreline_date(sites,
 save(step_reso, wmeans, shorelinedates,
      file = here("analysis/data/derived_data/sdates.RData"))
 
+load(file = here("analysis/data/derived_data/sdates.RData"))
+
 # Sum the probability distributions for the shoreline dates
 sumdates <- sum_shoredates(shorelinedates)
 sumdatesdf <- as.data.frame(sumdates) %>%
@@ -101,25 +103,28 @@ combined_sites <- filter(sites, !ask_id %in% out_of_range)
 wplot <- ggplot() +
   geom_histogram(aes(x = wmeans, y = ..density..),
                  binwidth = 200,
-                 col = "black", fill = "grey", alpha = 0.7) +
-  geom_density(aes(x = wmeans),
-               col = "black") +
-  scale_x_continuous(breaks = seq(-10000, 2500, 1000)) +
-  labs(title = "Mean shoreline date", x = "BCE", y = "Density") +
+                 col = "grey", fill = "grey") +
+  # geom_density(aes(x = wmeans),
+  #              col = "black") +
+  scale_x_continuous(breaks = seq(-10000, 2500, 1000), limits = c(-9500, -2500)) +
+  labs(title = "Mean TPQ dates", x = "BCE", y = "Density") +
   theme_bw()
 
 splot <- ggplot() +
-  geom_line(data = sumdatesdf, aes(x = sum.bce, y = sum.probability)) +
+  # geom_line(data = sumdatesdf, aes(x = sum.bce, y = sum.probability)) +
+  geom_bar(data = sumdatesdf, aes(x = sum.bce, y = sum.probability),
+                 stat = "identity",
+                 col = "grey", fill = "grey") +
   scale_x_continuous(breaks = seq(-10000, 2500, 1000)) +
-  labs(title = "SSPD", x = "BCE", y = "Summed probability") +
+  labs(title = "Summed probability distribution\nof shoreline dates", x = "BCE", y = "Summed probability") +
   theme_bw()
 
 eplot <- ggplot() +
   geom_histogram(aes(x = combined_sites$elev, y = ..density..),
                  binwidth = 4,
-                 col = "black", fill = "grey", alpha = 0.7) +
-  geom_density(aes(x = combined_sites$elev),
-             col = "black", alpha = 0.7) +
+                 col = "grey", fill = "grey") +
+  # geom_density(aes(x = combined_sites$elev),
[…]

### 2023-06-11 R/0x_radiocarbon_cpl.R

@@ -29,6 +29,8 @@ cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
 save(exp, logi, cpl_1, cpl_2, cpl_3, cpl_4, cpl_5,
           file = here("analysis/data/derived_data/rcarbon_models.RData"))
 
+load(file = here("analysis/data/derived_data/rcarbon_models.RData"))
+
 expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
 expp$model <- "Exponential"
 logip <- convertPars(pars = logi$par, years = minage:maxage, type = 'logistic')
@@ -60,6 +62,9 @@ rcarbon_lik <- c("Exponential" = -exp$value,
 # "CPL-3" = -cpl_3$value,
 # "CPL-4" = -cpl_4$value
 
+# Effective sample size
+ssize <- ncol(pd)
+
 rcarbon_bic <- c(log(ssize)*1 - 2*rcarbon_lik[1], # Exponential
                log(ssize)*2 - 2*rcarbon_lik[2], # Logistic
                log(ssize)*0  - 2*rcarbon_lik[3], # Uniform
@@ -67,7 +72,8 @@ rcarbon_bic <- c(log(ssize)*1 - 2*rcarbon_lik[1], # Exponential
                log(ssize)*3 - 2*rcarbon_lik[5], # CPL-2
                log(ssize)*5 - 2*rcarbon_lik[6], # CPL-3
                log(ssize)*7 - 2*rcarbon_lik[7], # CPL-4
-               log(ssize)*9 - 2*rcarbon_lik[8]) # CPL-5
+               log(ssize)*9 - 2*rcarbon_lik[8] # CPL-5
+               ) #log(ssize)*11 - 2*rcarbon_lik[9] # CPL-6
 
 # log(ssize)*5 - 2*shore_lik[6],
 # log(ssize)*7 - 2*shore_lik[6],
@@ -94,8 +100,8 @@ cplplt <- ggplot() +
 
 
 bicplt <- ggplot() +
-  geom_point(aes(names(rcarbon_bic), rcarbon_bic), size = 3) +
-  labs(x = "Model", y = "BIC") +
+  geom_point(aes(rcarbon_bic, names(rcarbon_bic)), size = 3) +
+  labs(x = "BIC", y = "Model") +
   theme_bw()
 
 cplplt + bicplt

### 2023-06-11 R/0x_shoreline_cpl.R

@@ -1,0 +1,84 @@
+library(ADMUR)
+library(DEoptimR)
+
+set.seed(42)
+
+# Load data from radiocarbon_mc.R: models fit for MC, the pd and the min and
+# max ages
+load(file = here::here("analysis/data/derived_data/shore_pd_models.RData"))
+
+# Fit CPL-models to shoreline dates
+cpl_1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction, PDarray = pd,
+                  type = 'CPL', NP = 20, trace = TRUE)
+cpl_2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 60,  maxiter = 400 * 3,
+                  trace = TRUE)
+cpl_3 <- JDEoptim(lower = rep(0, 5), upper = rep(1, 5), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 100, maxiter = 400 * 10,
+                  trace = TRUE)
+cpl_4 <- JDEoptim(lower = rep(0, 7), upper = rep(1, 7), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 140, maxiter = 400 * 8,
+                  trace = TRUE)
+cpl_5 <- JDEoptim(lower = rep(0, 9), upper = rep(1, 9), fn = objectiveFunction,
+                  PDarray = pd, type='CPL', NP = 180, maxiter = 400 * 9,
+                  trace = TRUE)
+cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 220, maxiter =  400 * 11,
+                  trace = TRUE)
+
+save(exp, logi, unif, cpl_1, cpl_2, cpl_3,
+     file = here::here("analysis/data/derived_data/shore_models.RData"))
+load(file = here::here("analysis/data/derived_data/shore_models.RData"))
+
+expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
+expp$model <- "Exponential"
+logip <- convertPars(pars = logi$par, years = minage:maxage, type = 'logistic')
+logip$model <- "Logistic"
[…]

### 2023-06-11 R/0x_shoreline_monte_carlo2.R

@@ -42,6 +42,9 @@ pd <- sweep(pd, 2, colSums(pd),"/")
 SPD <- as.data.frame(rowSums(pd))
 SPD <- SPD/( sum(SPD) *5 )
 
+minage <- min(as.numeric(row.names(SPD)))
+maxage <- max(as.numeric(row.names(SPD)))
+
 # Maximum likelihood search using DEoptimR
 exp <- JDEoptim(lower = -0.01, upper = 0.01, fn = objectiveFunction,
                 PDarray = pd, type = "exp", NP = 20)
@@ -51,13 +54,16 @@ logi <- JDEoptim(lower = c(0, 0000), upper = c(1, 10000),
 # No search required for uniform
 unif <- -objectiveFunction(pars = NULL, PDarray = pd, type = 'uniform')
 
-expp <- convertPars(pars = exp$par, years = -9445:-2500, type = 'exp')
+expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
 expp$model <- "Exponential"
-logip <- convertPars(pars = logi$par, years = -9445:-2500, type = 'logistic')
+logip <- convertPars(pars = logi$par, years = minage:maxage, type = 'logistic')
 logip$model <- "Logistic"
 unifp <- convertPars(pars = NULL, years =  -9445:-2500, type = "uniform")
 unifp$model <- "Uniform"
 
+# Save to use with CPL models
+save(pd, exp, logi, unif, maxage, minage,
+     file = here("analysis/data/derived_data/shore_pd_models.RData"))
 
 # Combine models for plotting
 shoremodels <- rbind(expp, logip, unifp)
@@ -89,7 +95,7 @@ incpolys$dens <- lengths(st_intersects(incpolys, sites)) /
 ssize = sumdates$dates_n
 
 # Number of simulations
-nsim <- 10
+nsim <- 10000
 
 names(expp) <- c("bce", "prob_dens", "model")
 
@@ -177,22 +183,34 @@ simresults <- simresults %>%
 
 expplt <- ggplot() +
   geom_vline(xintercept = as.numeric(names(expsum$busts)),
-             col = "firebrick", alpha = 0.04) +
+             col = "firebrick", alpha = 0.06) +
   geom_vline(xintercept = as.numeric(names(expsum$booms)),
-             col = "darkgreen", alpha = 0.04) +
+             col = "darkgreen", alpha = 0.06) +
   ggplot2::geom_ribbon(data = simresults, aes(x = bce, ymin = low, ymax = high),
[…]

### 2023-06-11 R/0xmean_tpq_ranges.R

@@ -1,0 +1,98 @@
+library(shoredate)
+
+
+# Load centre points for the displacement curves
+centrepoints <- sf::st_read(
+  system.file("extdata/isobase_centrepts.gpkg",
+              package = "shoredate",
+              mustWork = TRUE), quiet = TRUE)
+
+
+# Horten
+
+# Retreieve the point on the Horten isobase
+horten <- centrepoints[1,]
+
+# Create a sequence of eleveation values from well above the marine limit for
+# the curves and down to 20 masl
+elevations <- seq(190, 1, -1)
+
+# Make these a series of "sites" to be dated
+sites <- sf::st_sf(rep(sf::st_geometry(horten), length(elevations)))
+sites$name <- paste(elevations, "masl")
+
+# Perform dating procedure, setting model to "none" and HDR to 1 to account for
+# the full range.
+hsdates <- shoreline_date(sites,
+                          elevation = elevations,
+                          hdr_prob = 1,
+                          model = "none")
+
+# Plot for inspection
+shoredate_plot(hsdates, date_probability = FALSE, multiplot = TRUE)
+
+# Porsgrunn curve
+porsgrunn <- centrepoints[2,]
+sites <- sf::st_sf(rep(sf::st_geometry(porsgrunn), length(elevations)))
+sites$name <- paste(elevations, "masl")
+psdates <- shoreline_date(sites,
+                          elevation = elevations,
+                          hdr_prob = 1,
+                          model = "none")
+
+shoredate_plot(psdates, date_probability = FALSE, multiplot = TRUE)
+
+# Arendal curve
+arendal <- centrepoints[3,]
+sites <- sf::st_sf(rep(sf::st_geometry(arendal), length(elevations)))
+sites$name <- paste(elevations, "masl")
+asdates <- shoreline_date(sites,
+                          elevation = elevations,
+                          hdr_prob = 1,
+                          model = "none")
+
+shoredate_plot(asdates, date_probability = FALSE, multiplot = TRUE)
+
+# Tvedestrand curve
+tvedestrand <- centrepoints[4,]
+sites <- sf::st_sf(rep(sf::st_geometry(tvedestrand), length(elevations)))
+sites$name <- paste(elevations, "masl")
+tsdates <- shoreline_date(sites,
[…]

### 2023-06-14 analysis/paper/paper.Rmd
ADMUR for radiocarbon

@@
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
-Analysis of the ^14^C-dates were done using the R package *rcarbon* [@crema2021].
-The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b].
-Finally, the package *ADMUR* [@timpson2021] was used for model comparison.
+Analysis of the ^14^C-dates were done using the R package *ADMUR* [@timpson2021].
+The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with repurposed code from both *ADMUR* and *rcarbon* [@crema2021] for Monte Carlo simulation and model comparison.
 

### 2023-06-14 R/0x_radiocarbon_cpl.R

@@ -1,11 +1,14 @@
 library(ADMUR)
 library(DEoptimR)
+library(ggplot2)
+library(patchwork)
+library(here)
 
 set.seed(42)
 
 # Load data from radiocarbon_mc.R: models fit for MC, the pd and the min and
 # max ages
-load(file = here("analysis/data/derived_data/rcarbon_pd_models.RData"))
+load(file = here::here("analysis/data/derived_data/rcarbon_pd_models.RData"))
 
 # Fit CPL-models to radiocarbon data
 cpl_1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction, PDarray = pd,
@@ -38,29 +41,29 @@ logip$model <- "Logistic"
 unifp <- convertPars(pars = NULL, years = minage:maxage, type = "uniform")
 unifp$model <- "Uniform"
 cpl1 <- convertPars(pars = cpl_1$par, years = minage:maxage, type = 'CPL')
-cpl1$model <- "CPL-1"
+cpl1$model <- "1-CPL"
 cpl2 <- convertPars(pars = cpl_2$par, years = minage:maxage, type = 'CPL')
-cpl2$model <- "CPL-2"
+cpl2$model <- "2-CPL"
 cpl3 <- convertPars(pars = cpl_3$par, years = minage:maxage, type = 'CPL')
-cpl3$model <- "CPL-3"
+cpl3$model <- "3-CPL"
 cpl4 <- convertPars(pars = cpl_4$par, years = minage:maxage, type = 'CPL')
-cpl4$model <- "CPL-4"
+cpl4$model <- "4-CPL"
 cpl5 <- convertPars(pars = cpl_5$par, years = minage:maxage, type = 'CPL')
-cpl5$model <- "CPL-5"
+cpl5$model <- "5-CPL"
 
 rcarbonmodels <- rbind(expp, logip, unifp, cpl1, cpl2, cpl3, cpl4, cpl5)
 
 rcarbon_lik <- c("Exponential" = -exp$value,
                "Logistic" = -logi$value,
                "Uniform" = unif,
-               "CPL-1" = -cpl_1$value,
-               "CPL-2" = -cpl_2$value,
-               "CPL-3" = -cpl_3$value,
-               "CPL-4" = -cpl_4$value,
-               "CPL-5" = -cpl_5$value)
+               "1-CPL" = -cpl_1$value,
[…]

### 2023-06-14 R/0x_shoreline_monte_carlo2.R

@@ -31,11 +31,13 @@ sdates$site <- rep(1:length(dates_dfs) , each = nrow(sdates)/length(dates_dfs))
 
 pd <- as.data.frame(sdates) %>%
   dplyr::filter(bce <= -2500 & bce >= -9445) %>%
-  dplyr::group_by(bce) %>%
+  dplyr::mutate(bp = (bce + 1950) * - 1) %>%
+  dplyr::select(-bce) %>%
+  dplyr::group_by(bp) %>%
   dplyr::filter(!is.na(probability)) %>%
   dplyr::group_by(site) %>%
   tidyr::pivot_wider(names_from = site, values_from = probability) %>%
-  tibble::column_to_rownames("bce")
+  tibble::column_to_rownames("bp")
 
 pd <- sweep(pd, 2, colSums(pd),"/")
 
@@ -58,7 +60,7 @@ expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
 expp$model <- "Exponential"
 logip <- convertPars(pars = logi$par, years = minage:maxage, type = 'logistic')
 logip$model <- "Logistic"
-unifp <- convertPars(pars = NULL, years =  -9445:-2500, type = "uniform")
+unifp <- convertPars(pars = NULL, years =  minage:maxage, type = "uniform")
 unifp$model <- "Uniform"
 
 # Save to use with CPL models
@@ -75,7 +77,8 @@ ggplot() +
   geom_line(aes(x = as.numeric(rownames(SPD)),
                 y = SPD[,1])) +
   geom_line(data = shoremodels, aes(year, pdf, col = model),
-            linewidth = 1.1)
+            linewidth = 1.1) +
+  scale_x_reverse()
 
 
 #### Set up Monte Carlo simulation ####

### 2023-06-14 R/10_radiocarbon_cpl_admur.R

@@ -1,0 +1,217 @@
+library(dplyr)
+library(rcarbon)
+library(ADMUR)
+library(patchwork)
+library(ggplot2)
+library(DEoptimR)
+
+set.seed(42)
+
+c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
+c14 <- c14 %>% filter(context != "Food crust")
+
+names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
+
+CalArray <- makeCalArray(intcal20, calrange = c(minage, maxage))
+PD <- phaseCalibrator(c14, CalArray, remove.external = TRUE)
+
+SPD <- as.data.frame( rowSums(PD) )
+# normalise
+SPD <- SPD/( sum(SPD) * CalArray$inc )
+
+
+# CPL parameters must be between 0 and 1, and an odd length.
+CPL.1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction,
+                  PDarray = PD, type = 'CPL', NP = 20)
+CPL.2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3),
+                  fn = objectiveFunction, PDarray = PD, type = 'CPL', NP = 60)
+CPL.3 <- JDEoptim(lower = rep(0, 5), upper = rep(1, 5),
+                  fn = objectiveFunction, PDarray = PD, type = 'CPL', NP = 100)
+CPL.4 <- JDEoptim(lower = rep(0, 7), upper = rep(1, 7),
+                  fn = objectiveFunction, PDarray = PD, type = 'CPL', NP = 140)
+exp <- JDEoptim(lower = -0.01, upper = 0.01, fn = objectiveFunction,
+                PDarray = PD, type = 'exp', NP = 20)
+log <- JDEoptim(lower = c(0, 0000), upper = c(1, 10000), fn = objectiveFunction,
+                PDarray = PD, type = 'logistic', NP = 40)
+uniform <- objectiveFunction(NULL, PD, type = 'uniform')
+
+save(exp, log, uniform, CPL.1, CPL.2, CPL.3, CPL.4,
+     file = here("analysis/data/derived_data/rcarbon_models.RData"))
+
+CPL1 <- convertPars(pars=CPL.1$par, years=minage:maxage, type='CPL')
+CPL2 <- convertPars(pars=CPL.2$par, years=minage:maxage, type='CPL')
+CPL3 <- convertPars(pars=CPL.3$par, years=minage:maxage, type='CPL')
[…]

### 2023-06-14 R/0x_shoreline_cpl.R
Re-did pd for CPL models with shoreline dates

@@ -37,24 +37,28 @@ logip$model <- "Logistic"
 unifp <- convertPars(pars = NULL, years = minage:maxage, type = "uniform")
 unifp$model <- "Uniform"
 cpl1 <- convertPars(pars = cpl_1$par, years = minage:maxage, type = 'CPL')
-cpl1$model <- "CPL-1"
+cpl1$model <- "1-CPL"
 cpl2 <- convertPars(pars = cpl_2$par, years = minage:maxage, type = 'CPL')
-cpl2$model <- "CPL-2"
+cpl2$model <- "2-CPL"
+cpl3 <- convertPars(pars = cpl_3$par, years = minage:maxage, type = 'CPL')
+cpl3$model <- "3-CPL"
 
-shmodels <- rbind(expp, logip, unifp, cpl1, cpl2)
+shmodels <- rbind(expp, logip, unifp, cpl1, cpl2, cpl3)
 
 sh_lik <- c("Exponential" = -exp$value,
                  "Logistic" = -logi$value,
                  "Uniform" = unif,
-                 "CPL-1" = -cpl_1$value,
-                 "CPL-2" = -cpl_2$value)
+                 "1-CPL" = -cpl_1$value,
+                 "2-CPL" = -cpl_2$value,
+                 "3-CPL" = -cpl_3$value)
 
 ssize <- ncol(pd)
 sh_bic <- c(log(ssize)*1 - 2*sh_lik[1], # Exponential
                  log(ssize)*2 - 2*sh_lik[2], # Logistic
                  log(ssize)*0  - 2*sh_lik[3], # Uniform
                  log(ssize)*1 - 2*sh_lik[4], # CPL-1
-                 log(ssize)*3 - 2*sh_lik[5]) # CPL-2
+                 log(ssize)*3 - 2*sh_lik[5],
+                 log(ssize)*5 - 2*sh_lik[6]) # CPL-2
 
 SPD <- as.data.frame(rowSums(pd))
 SPD <- SPD/( sum(SPD) *5 )

### 2023-06-22 analysis/paper/paper.Rmd
Found and saved SSPD as ADMUR pd and exponential, logistic and uniform MLE models

@@
 
-The procedure outlined above represent the standard approach when modelling population dynamics in archaeology.
-This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of shoreline dated sites over time could be made.
+The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
+This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the verall development of this proxy could be made.
 However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves.
 As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*.
-This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test in line with the Monte Carlo approach outlined above.
-The benefit of the CPL models is that these define discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these.
+This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above.
+The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction.
+This therefore allows for the identification of key events and a conservative description of population dynamics between these.
 The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
@@
 Analysis of the ^14^C-dates were done using the R package *ADMUR* [@timpson2021].
-The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with repurposed code from both *ADMUR* and *rcarbon* [@crema2021] for Monte Carlo simulation and model comparison.
+The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* [@crema2021] for Monte Carlo simulation and model comparison.
 
@@
 
-To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned using the *binPrep()* function from *rcarbon*, where dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other.
-When calibrated, these are summed and divided by the number of dates in the bin before being included in the final RSPD.
-Furthermore, the ^14^C-dates were not normalised.
-This follows from a peculiarity in the calibration procedure which can lead the probability mass of a date to sum to more than one [@weninger2015].
-As a consequence of this, normalising the dates before summing has been shown to create spurious spikes in the RSPD [e.g.
-@crema2021].
-All calibration was done using the IntCal20 calibration curve [@reimer2020].
+To account for investigatory bias that can result from variable sampling intensity between sites, the summing of radiocarbon dates with *ADMUR* starts with binning dates from the same sites that fall within 200 uncalibrated ^14^C years of each other when measured from the mean ^14^C age.
+All dates are then calibrated using a resoltuion of 5 years and then normalised to sum to unity.
+The RSPDs for each site-phase are then summed to achieve the final SPD, which is in turn normalised.
+All calibrations were performed using the IntCal20 calibration curve [@reimer2020].
 
-The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD.
+The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD.
 Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*.
-The Monte Carlo simulations were then performed using the *modelTest()* function from *rcarbon*.
+The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*.
 For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD.
@@
 
-Following from the commutative nature of the shoreline dates dealt with here, the Monte Carlo simulation with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice.
+The Monte Carlo simulations with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice.
 This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021].
@@
 ```
-Comparing the the directions and deviations from the three standard models, Figure appears to indicate some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
+For the RSPD, the logistic model achieved a p-value of, and could therefore not be rejected.
+Based on the available data, this therefore represents the best explanation of the data.
+We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case.
+
+Comparing the the directions and deviations from the three standard models, Figure appears to indicate that some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
 However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD.
@@
 @roalkvam2022].
-Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of importance to unpack their relationship [e.g.
+Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g.
 @ordonez2022; @bergsvik2022; @jorgensen2020].
+
+Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application [see @roalkvam2023].
+While we still believe the approach represents an improvement over previous approaches, the results should therefore be treated with care.
+This also extends to the results of the RSPD, due to the fairly limited ^14^C sample.
+With these limitations in mind, the approach outlined by @timpson2021 as it offers a more conservative treatment of the data, reducing the amount of over-interpretation that might simply be the result of uncertainties to do with data or methods.
 
@@
 Although a more precise understanding of this relationship
-
-Furthermore, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application [see @roalkvam2023].
-While we still believe the approach represents a considerable improvement over previous approaches, the results should
 

### 2023-06-22 R/03_functions.R

@@ -40,42 +40,134 @@ reverse_shoredate <- function(shoreline_date,
 }
 
 
-# Reverse a single calendar date to a corresponding elevation
-# Repurposed from rcarbon
-simulation_summary <- function(spd, simulation_results, cut_off, nsim){
+# # Reverse a single calendar date to a corresponding elevation
+# # Repurposed from rcarbon
+# simulation_summary <- function(spd, simulation_results, cut_off, nsim){
+#
+#   # Reformat data frame of normalised simulated SPDs
+#   sim <- simulation_results %>%
+#     filter(bce <= cut_off) %>%
+#     group_by(bce) %>%
+#     dplyr::reframe(prob_sum_normalised, simn) %>%
+#     tidyr::pivot_wider(names_from = simn, values_from = prob_sum_normalised) %>%
+#     tibble::column_to_rownames("bce")
+#
+#   # Format observed SPD and retrieve the probability
+#   fSPD <- as.data.frame(spd) %>%
+#     filter(sum.bce <= cut_off) %>%
+#     mutate(sum.probability = sum.probability/(sum(sum.probability) * 5)) %>%
+#     select(sum.probability) %>%
+#     unlist(use.names = FALSE)
+#
+#   Zsim <- t(apply(sim, 1, scale))
+#   zLo <- apply(Zsim, 1, quantile, prob = 0.025, na.rm = TRUE)
+#   zHi <- apply(Zsim, 1, quantile, prob = 0.975, na.rm = TRUE)
+#
+#   Zscore_empirical <- (fSPD - apply(sim, 1, mean))/apply(sim, 1, sd)
+#   busts <- which(Zscore_empirical < zLo)
+#   booms <- which(Zscore_empirical > zHi)
+#
+#   observedStatistic <- sum(c(zLo[busts] - Zscore_empirical[busts]),
+#                            c(Zscore_empirical[booms] - zHi[booms]))
+#   expectedstatistic <- abs(apply(Zsim, 2, function(x, y){a = x - y; i = which(a < 0); return(sum(a[i]))}, y = zLo)) +
+#     apply(Zsim, 2, function(x, y){a = x - y; i = which(a > 0); return(sum(a[i]))}, y = zHi)
+#
+#
+#   pvalue <- c(length(expectedstatistic[expectedstatistic > observedStatistic]) + 1)/c(nsim + 1)
+#
+#   return(list(pvalue = pvalue,
+#               busts = busts,
[…]

### 2023-06-22 R/0x_shoreline_monte_carlo2.R

@@ -7,7 +7,7 @@ library(patchwork)
 library(ADMUR)
 library(DEoptimR)
 
-set.seed(42)
+set.seed(1)
 
 # Source functions
 source(here("R/03_functions.R"))
@@ -26,46 +26,68 @@ dates_dfs <- lapply(shorelinedates, as.data.frame)
 # Combine these into a single data frame
 sdates <- do.call(rbind, dates_dfs)
 
-# Assign simulation results for each site a unique ID
+# Assign results for each site a unique ID
 sdates$site <- rep(1:length(dates_dfs) , each = nrow(sdates)/length(dates_dfs))
 
+# Format for ADMUR, changing BCE to BP
 pd <- as.data.frame(sdates) %>%
   dplyr::filter(bce <= -2500 & bce >= -9445) %>%
-  dplyr::mutate(bp = (bce + 1950) * - 1) %>%
+  dplyr::mutate(bp = (bce * -1) + 1950) %>%
   dplyr::select(-bce) %>%
   dplyr::group_by(bp) %>%
   dplyr::filter(!is.na(probability)) %>%
   dplyr::group_by(site) %>%
   tidyr::pivot_wider(names_from = site, values_from = probability) %>%
   tibble::column_to_rownames("bp")
-
 pd <- sweep(pd, 2, colSums(pd),"/")
 
+# Create SPD using the ADMUR approach
 SPD <- as.data.frame(rowSums(pd))
 SPD <- SPD/( sum(SPD) *5 )
 
-minage <- min(as.numeric(row.names(SPD)))
-maxage <- max(as.numeric(row.names(SPD)))
+# Find mininum and maximum ages (BP)
+minage <- min(as.numeric(row.names(pd)))
+maxage <- max(as.numeric(row.names(pd)))
 
 # Maximum likelihood search using DEoptimR
 exp <- JDEoptim(lower = -0.01, upper = 0.01, fn = objectiveFunction,
                 PDarray = pd, type = "exp", NP = 20)
-logi <- JDEoptim(lower = c(0, 0000), upper = c(1, 10000),
-                 fn = objectiveFunction, PDarray = pd, type = 'logistic',
-                 NP = 40, trace = TRUE)
 # No search required for uniform
 unif <- -objectiveFunction(pars = NULL, PDarray = pd, type = 'uniform')
 
 expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
 expp$model <- "Exponential"
[…]

### 2023-06-22 R/10_radiocarbon_cpl_admur.R

@@ -4,6 +4,7 @@ library(ADMUR)
 library(patchwork)
 library(ggplot2)
 library(DEoptimR)
+library(here)
 
 set.seed(42)
 
@@ -11,6 +12,9 @@ c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
 c14 <- c14 %>% filter(context != "Food crust")
 
 names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
+
+minage <- 4500
+maxage <- 10975
 
 CalArray <- makeCalArray(intcal20, calrange = c(minage, maxage))
 PD <- phaseCalibrator(c14, CalArray, remove.external = TRUE)
@@ -37,6 +41,7 @@ uniform <- objectiveFunction(NULL, PD, type = 'uniform')
 
 save(exp, log, uniform, CPL.1, CPL.2, CPL.3, CPL.4,
      file = here("analysis/data/derived_data/rcarbon_models.RData"))
+load(file = here("analysis/data/derived_data/rcarbon_models.RData"))
 
 CPL1 <- convertPars(pars=CPL.1$par, years=minage:maxage, type='CPL')
 CPL2 <- convertPars(pars=CPL.2$par, years=minage:maxage, type='CPL')
@@ -59,24 +64,17 @@ BIC.2 <- 3*log(155) - 2*(-CPL.2$value)
 BIC.3 <- 5*log(155) - 2*(-CPL.3$value)
 BIC.4 <- 7*log(155) - 2*(-CPL.4$value)
 BIC.exp <- 1*log(155) - 2*(-exp$value)
-BIC.log <- 2*log(ssize) - 2*(-log$value)
+BIC.log <- 2*log(155) - 2*(-log$value)
 BIC.uniform <- 0 - 2*(-uniform)
 
-bics <- data.frame(model = c("BIC.1", "BIC.2", "BIC.3",
-                             "BIC.4", "BIC.exp", "BIC.log", "BIC.uniform"),
+bics <- data.frame(model = c("1-CPL", "2-CPL", "3-CPL",
+                             "4-CPL", "Exponential", "Logistic", "Uniform"),
                    bic = c(BIC.1, BIC.2, BIC.3, BIC.4, BIC.exp,
                            BIC.log, BIC.uniform))
 
[…]

### 2023-06-22 analysis/paper/paper.Rmd
Draft to Steinar

@@
   \usepackage{placeins}
+# always_allow_html: true
 bibliography: references.bib
@@
 
-knitr::opts_knit$set(
-  eval.after = "fig.cap"
-)
+# knitr::opts_knit$set(
+#   eval.after = "fig.cap"
+# )
 
@@
 @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].
-The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach.
+The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach.
 Several limitations and forms of criticism have, however, been directed at these procedures.
-Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015].
+Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g.
+@carleton2021; @torfing2015; @ward2021].
 What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021].
@@
 The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD.
-Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*.
+Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006].
 The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*.
@@
 
-Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon*.
+Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon* and *ADMUR*.
 However, idiosyncrasies in the dating method did necessitate some adjustments.
@@
 As it can be reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
-However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbritrary error of $\pm$ 100 years to this point-estimate.
-However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use, limits, as with TPQ dates in general, the further inferential steps that can be taken.
+However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate.
+However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
 Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
+
 This last point can be illustrated by finding the range between the upper and the lower limit of the four geological displacement curves for every meter from the marine limit and down to elevations that do not give an earliest date younger than 2500 BCE.
@@
 This approach is illustrated in Figure \@ref(fig:sumshore)B.
-The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, clearly demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c.
+The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c.
 8000 BCE.
@@
 B) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates.
-C) The result of instead summing the probability of the shoreline dates achievied with the method employed here, which accounts for the likely distance between sites and shoreline.
+C) The result of instead summing the probability of the shoreline dates achieved with the method employed here, which accounts for the likely distance between sites and shoreline.
 The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}
@@
 The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs.
-This was done using re-purposed code from *rcarbon*.
+This was done using re-purposed code from *ADMUR*.
 
-## Continous piece-wise linear models
+## Continuous piece-wise linear models
 The procedures outlined above represent standard approaches when modelling population dynamics in archaeology.
-These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shorebound sites over time could be made.
+These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shore-bound sites over time could be made.
 However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case.
@@
 This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above.
-The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these.
+The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these.
 Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.
@@
 For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase.
-The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021], found by estimating the likelihood using the final SPD.
-From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) for each model was then found.
-This penalises the models for the number of parameters in use to avoid over-fitting.
+The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, 3], found by estimating the likelihood using the final SPD.
+From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model.
+This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.
 
@@
 
-```{r spduniform, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates.
-A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model.
-B) Same as A, but with a exponential model.
-C) Summed probability of calibrated radiocarbon dates (n =, bins = 678) compared to a uniform null model.
-D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
+```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates.
+A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model.
+B) SSPD compared to a logistic null model.
+C) SSPD compared to a uniform null model.
+D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model.
+E) RSPD compared to a logistic null model.
+F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
 knitr::include_graphics(here("analysis/figures/mc.png"))
@@
 Based on the available data, this therefore represents the best explanation of the data.
-We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case.
+For the SSPD all of the standard models could be rejected.
 
-Comparing the the directions and deviations from the three standard models, Figure appears to indicate that some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time.
-However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD.
-Based on common demographic developments and previous suggestions, there is possibly some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments.
-However, as introduced above, several authors have warned against over-interpreting such deviation.
-Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.
+<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
 
-For the RSPD the rejected logistic model offers the best approximation of the data of the compared models.
-If taken as an indication of demographic developments this would indicate a first migration of people that is more substantial that what would be expected under the logistic null.
-Further deviations are then evident with the negative deviation and the poistive deviation c.
-These are, however, brief and can be left as little more than suggestions for possible meaningful demographic events.
+Comparing the the directions of the three standard models, Figure \@ref(fig:spd_mc) appears to indicate an overall uniform distribution in the frequency of radiocarbon dates following the initial jump just before 8000 BCE, contrasted to a general decrease in shoreline dated sites after the initial rapid increase from around 9500 BCE.
 
-In the following, some of these deviations will nonetheless be narratively linked to some major developments that have previously been proposed to characterise the Stone Age of south-eastern Norway, with the goal of suggesting some potential.
-It is, however, important to underscore that the *ad-hoc* coupling of the results achieved here with these developments should only be considered suggestions that can offer some potential avenues for further inquiry, for example using the approach outlined by @timpson2021.
+It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a drop just before 6000 BCE.
+However, given that the logistic model cannot be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and therefore no analytical weight can be placed on these deviations -- at least given the present data and modelling efforts.
 
@@
 
-It is firmly established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE.
+It is firmly established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g.
+@glorstad2016; @damlien2018b].
 While the SSPD starts from c.
-9450 BCE, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand.
-Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations, ranging from taphonomic loss, to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a lack of reliable access to firewood.
+9500 BCE, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand.
+Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years, ranging from taphonomic loss to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a lack of reliable access to firewood.
 While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump in ^14^C-dates at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind this pattern.
 
-While all the considered models can be rejected as explaining the analysed data, they do nonetheless indicate that some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
+The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
 The most immediate explanation of this mismatch would be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
-The modelled process of logistic decay markedly under-predicts the SSPD in the period X--X, and a deviation is also apparent in the period X--X, where the number of observed sites is significantly lower than what would be expected under the logistic model.
-While we are wary of over-interpreting these signals, they strike us as warranting more consideration than the other deviations that are barely outside the critical envelope, or are of a short duration.
-The deviations from after 4000 BCE should also be treated with further care, following both from questions concerning the reliability of shoreline dating after 4000 BCE [cf.
-@roalkvam2023].
+
+The development after 4000 BCE should be treated with further care, following from questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023].
 
@@
 If this narrative is coupled to the data observed here, this process would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, while the end of this drop corresponds to the first ^14^C-dates in the RSPD -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is related to cultural processes.
-In the period X--X following this, there is also a marked postive deviation from the RSPD under the exponential model, indicting a population boom, which, although barely breaking the critical envelope of the logitic model, is also evident in the SSPD.
 
@@
 However, their analysis is seemingly also based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
-Apart from the limitations with using point-based estimates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
-The analysis undertaken here instead indicates a drop in the number of sites through this time-interval.
-While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is difficult to speculate about.
-However, the positive deviation in the SSPD from the logistic model at c.
+Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
+In contrast to their findings, the analysis undertaken here instead indicates a drop in the number of sites through this time-interval.
+
+While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain.
+
+<!-- However, the positive deviation in the SSPD from the logistic model at c.
 does correspond to a similar deviation of the RSPD from the exponential model.
@@
 While we are, as mentioned, cautions in giving these much weight in the SSPD alone, the correspondence of these deviations across the two proxies gives some reason to suggests that these might reflect meaningful demographic signals.
-This would also i
+This would also i -->
 
@@
 Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
-@riede2012; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
+@collard2013a; @riede2014; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
 @roalkvam2022].
 Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g.
-@ordonez2022; @bergsvik2022; @jorgensen2020].
+@ordonez2022; @jorgensen2020].
 
-Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application [see @roalkvam2023].
+Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application.
 While we still believe the approach represents an improvement over previous approaches, the results should therefore be treated with care.
 This also extends to the results of the RSPD, due to the fairly limited ^14^C sample.
-With these limitations in mind, the approach outlined by @timpson2021 as it offers a more conservative treatment of the data, reducing the amount of over-interpretation that might simply be the result of uncertainties to do with data or methods.
+With these limitations in mind, the approach outlined by @timpson2021 gives a more reasonable framework with which to evaluate the available data, as leads to a more conservative analysis and reduces the amount of over-interpretation that might follow from uncertainties to do with data or methods.
 
@@
 Although a more precise understanding of this relationship
+
+# Ekstra figurer
+
+Jeg fikk ADMUR til å funke med C14-dateringer.
+Figuren under viser at den logaritmiske modellen (blå kurve, samme som i Figur 5 over) ifølge BIC passer best av alternativene , og Monte Carlo-simuleringen over viser at den ikke kan avvises som forklarende for dataen.
+Eller med andre ord, dataen ville kunne forventes under den logaritmiske utviklingen.
+CPL modellering av strandlinjedateringene sliter jeg imidlertid mer med.
+Modellene opp til 4-CPL ser rimelige ut, men toppen på 5-CPL (sort kurve) som ligger litt før 5000 f.Kr.
+virker feil.
+I tillegg indikerer BIC-verdiene at denne passer best til dataen, som også virker rart gitt at den avviker såpass fra SSPDen.
+
+```{r cpl_rcarbon, echo = FALSE, out.width = "500px", fig.align="center", regfloat = TRUE}
+# fig.cap = "Model comparison, summed probability of radiocarbon dates.",
+knitr::include_graphics(here("analysis/figures/cplrcarbon.png"))
+```
+
+```{r cpl_shore, echo = FALSE, out.width = "500px", fig.align="center", regfloat = TRUE}
+# fig.cap = "Model comparison, summed probability of shoreline dates.",
+knitr::include_graphics(here("analysis/figures/cplshore.png"))
+```
 

### 2023-06-22 R/03_functions.R

@@ -144,9 +144,13 @@ simulation_summary <- function(spd, simulation_results,
 }
 
 
-plot_mc <- function(mc_summary, title = NULL) {
+plot_mc <- function(mc_summary, xppos, yppos, title = NULL) {
   booms <- mc_summary$timeseries$calBP[mc_summary$timeseries$index == 1]
   busts <- mc_summary$timeseries$calBP[mc_summary$timeseries$index == -1]
+
+  spdmax <- max(mc_summary$timeseries$SPD)
+  spdmin <- min(mc_summary$timeseries$SPD)
+  spdadj <- (spdmax - spdmin) * 1.1
 
   p <- round(mc_summary$pvalue, 3)
   plabel <- ifelse(p == 0, "p < 0.001", paste("p =", p))
@@ -166,8 +170,8 @@ plot_mc <- function(mc_summary, title = NULL) {
                     breaks = seq(12000, 4500, -1000),
                     expand = expansion(mult = c(0, 0)),
                     labels = function(x)(x-2000)*-1) +
-    geom_text(aes(-Inf, Inf, hjust = 11, vjust = 1.5,
-                  label = plabel)) +
-    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
+    geom_text(aes(11000, spdadj), label = plabel) +
+    scale_y_continuous(expand = expansion(mult = c(0, 0.1)),
+                       labels = scales::label_comma()) +
     theme_bw()
 }

### 2023-06-22 R/0x_shoreline_monte_carlo2.R

@@ -197,23 +197,15 @@ for(i in seq_along(resfiles)){
   rm(tmp)
 }
 
-results <- list()
-for(i in 1:1000){
-  load(resfiles[i])
-  results[[i]] <- bind_rows(tmp)
-  rm(tmp)
-}
-
 # Retrieve results and normalise each simulated SPD
 simresults <- do.call(rbind.data.frame, results) %>%
   group_by(simn) %>%
   mutate(prob_sum_normalised = prob_sum /(sum(prob_sum, na.rm = TRUE) * 5))
 
-# Model summary (can also be plotted with ADMUR)
-expp <- convertPars(pars = exp$par,
-                    years = as.numeric(rownames(SPD)), type = 'exp')
-mod <- approx(x = expp$year, y = expp$pdf, xout = as.numeric(rownames(SPD)),
+mod <- approx(x = expp$bce, y = expp$prob_dens, xout = as.numeric(rownames(SPD)),
               ties = 'ordered', rule = 2)$y
+
+# Model summary (can also be plotted with ADMUR)
 exp_summary <- simulation_summary(SPD, simulation_results,
                                   c(-2500, -9445), mod, ncol(pd))
 
@@ -221,9 +213,9 @@ exp_summary <- simulation_summary(SPD, simulation_results,
 plotSimulationSummary(exp_summary)
 
 # Custom plot
-plot_mc(exp_summary)
+sexpplt <- plot_mc(exp_summary)
 
-save(expplt, file = "../external_data/shorespd/expplt.rda")
+save(sexpplt, file = "../external_data/shorespd/sexpplt.rda")
 
 ##### Monte Carlo simulation - Logistic #####
 
@@ -294,38 +286,22 @@ simresults <- do.call(rbind.data.frame, results) %>%
   dplyr::mutate(
     prob_sum_normalised = prob_sum / (sum(prob_sum, na.rm = TRUE) * 5))
 
-logisum <- simulation_summary(spd = sumdates,
-                              simulation_results = simresults,
-                              cut_off = -2500, nsim)
+# Model summary (can also be plotted with ADMUR)
+mod <- approx(x = logip$bce, y = logip$prob_dens,
+              xout = as.numeric(rownames(SPD)),
+              ties = 'ordered', rule = 2)$y
 
-simresults <- simresults %>%
-  dplyr::group_by(bce) %>%
-  dplyr::mutate(low = quantile(prob_sum_normalised, prob = 0.025,
[…]

### 2023-06-22 R/10_radiocarbon_cpl_admur.R

@@ -127,6 +127,7 @@ unisum <- SPDsimulationTest(c14, calcurve = intcal20,
 
 save(expsum, logsum, unisum,
      file = here("analysis/data/derived_data/admur_model_test.RData"))
+load(file = here("analysis/data/derived_data/admur_model_test.RData"))
 
 # Plot for inspection (setting up plots below that follow same style as
 # that to be used with shoreline dates)
@@ -134,84 +135,13 @@ plotSimulationSummary(expsum)
 plotSimulationSummary(logsum)
 plotSimulationSummary(unisum)
 
-# logsum$n.dates.effective
-# round(logsum$n.phases.effective)
-expbooms <- expsum$timeseries$calBP[expsum$timeseries$index == 1]
-expbusts <- expsum$timeseries$calBP[expsum$timeseries$index == -1]
+rexpplot <- plot_mc(expsum)
+rlogplot <- plot_mc(logsum)
+runiplot <- plot_mc(unisum)
 
-expplt <- ggplot(data = expsum$timeseries, aes(x = calBP)) +
-  geom_vline(xintercept = expbusts,
-             col = "firebrick", alpha = 0.06) +
-  geom_vline(xintercept = expbooms,
-             col = "darkgreen", alpha = 0.06) +
-  geom_ribbon(aes(ymin = expsum$timeseries$`2.5%`,
-                  ymax = expsum$timeseries$`97.5%`),
-              fill = "grey60", alpha = 0.8) +
-  geom_line(aes(y = expsum$timeseries$SPD), linewidth = 0.5) +
-  geom_line(aes(y = expsum$timeseries$model),
-            linewidth = 0.5, col = "red") +
-  labs(x = "BCE", y = "Summed probability") +
-  scale_x_reverse(limits = c(12000, 4500),
-                  breaks = seq(12000, 4500, -1000),
-                  expand = expansion(mult = c(0, 0)),
-                  labels = function(x)(x-2000)*-1) +
-  geom_text(aes(-Inf, Inf, hjust = -0.25, vjust = 2,
-                label = paste("p =", expsum$pvalue))) +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
-  theme_bw()
+(sexpplt + slogplt + suniplt)/
+(rexpplot + rlogplot + runiplot) +
+  plot_annotation(tag_levels = "A")
 
-logbooms <- logsum$timeseries$calBP[logsum$timeseries$index == 1]
[…]

### 2023-06-30 analysis/paper/paper.Rmd
Added additional dates from Steinar and made map of 14C-dated sites

@@
 # Map prepared in 01_shoreline_spd.R 
-knitr::include_graphics(here("analysis/figures/map1.png"))
+knitr::include_graphics(here("analysis/figures/map2.png"))
 ```
@@
 
-```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by elevation.
+```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by altitude above present sea-level.
 The binwidth is 4 m.
@@
 ```
-For the RSPD, the logistic model achieved a p-value of, and could therefore not be rejected.
+For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected.
 Based on the available data, this therefore represents the best explanation of the data.
@@
 Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
-@collard2013a; @riede2014; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
-@roalkvam2022].
+@collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
+@barton2014; @clark2017; @preston2018; @roalkvam2022].
 Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g.
-@ordonez2022; @jorgensen2020].
+@jorgensen2020; @ordonez2022].
 

### 2023-06-30 R/01_shoreline_spd.R

@@ -14,7 +14,6 @@ excavated <- st_zm(read_sf(here("analysis/data/raw_data/excavated_sites.gpkg")))
 dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
 site_dat <- read.csv(here("analysis/data/raw_data/sites.csv"))
 
-
 ###### Prepare site data for shoreline dating ######
 
 # Exclude sites that haven't been excavated and which haven't been shoreline
@@ -40,10 +39,6 @@ surveyed$elev <- terra::extract(dtm, vect(surveyed), fun = mean)[, -1]
 
 # Exclude surveyed sites with a quality score worse than 3
 survq <- dplyr::filter(surveyed, quality < 4)
-
-#
-# surveyed <- surveyed %>% dplyr::mutate(lab = ifelse(quality < 4, "t", "f"))
-# excavated$lab <- "y"
 
 # Select and rename columns needed to perform shoreline dating
 rcolexcshore <- excshore %>% dplyr::select(ask_id, elev)
@@ -106,7 +101,8 @@ wplot <- ggplot() +
                  col = "grey", fill = "grey") +
   # geom_density(aes(x = wmeans),
   #              col = "black") +
-  scale_x_continuous(breaks = seq(-10000, 2500, 1000), limits = c(-9500, -2500)) +
+  scale_x_continuous(breaks = seq(-10000, 2500, 1000),
+                     limits = c(-9500, -2500)) +
   labs(title = "Mean TPQ dates", x = "BCE", y = "Density") +
   theme_bw()
 
@@ -116,7 +112,8 @@ splot <- ggplot() +
                  stat = "identity",
                  col = "grey", fill = "grey") +
   scale_x_continuous(breaks = seq(-10000, 2500, 1000)) +
-  labs(title = "Summed probability distribution\nof shoreline dates", x = "BCE", y = "Summed probability") +
+  labs(title = "Summed probability distribution\nof shoreline dates",
+       x = "BCE", y = "Summed probability") +
   theme_bw()
 
 eplot <- ggplot() +
@@ -207,12 +204,16 @@ sa <- ggplot() +
           size = 2, shape = 21,
           colour = "black", show.legend = "point") +
   scale_fill_manual(labels = c(paste0("Surveyed sites,\nincluded (n = ",
-                                      nrow(dplyr::filter(surveyed, quality < 4)) ,")"),
+                                      nrow(dplyr::filter(surveyed,
[…]

### 2023-06-30 R/03_functions.R

@@ -157,9 +157,9 @@ plot_mc <- function(mc_summary, xppos, yppos, title = NULL) {
 
   ggplot(data = mc_summary$timeseries, aes(x = calBP)) +
     geom_vline(xintercept = busts,
-               col = "firebrick", alpha = 0.06) +
+               col = "firebrick", alpha = 0.05) +
     geom_vline(xintercept = booms,
-               col = "darkgreen", alpha = 0.06) +
+               col = "darkgreen", alpha = 0.05) +
     geom_ribbon(aes(ymin = `2.5%`,
                     ymax = `97.5%`),
                 fill = "grey60", alpha = 0.8) +
@@ -169,7 +169,7 @@ plot_mc <- function(mc_summary, xppos, yppos, title = NULL) {
     scale_x_reverse(limits = c(12000, 4500),
                     breaks = seq(12000, 4500, -1000),
                     expand = expansion(mult = c(0, 0)),
-                    labels = function(x)(x-2000)*-1) +
+                    labels = function(x)(x-1950)*-1) +
     geom_text(aes(11000, spdadj), label = plabel) +
     scale_y_continuous(expand = expansion(mult = c(0, 0.1)),
                        labels = scales::label_comma()) +

### 2023-06-30 R/0x_shoreline_monte_carlo2.R

@@ -213,9 +213,9 @@ exp_summary <- simulation_summary(SPD, simulation_results,
 plotSimulationSummary(exp_summary)
 
 # Custom plot
-sexpplt <- plot_mc(exp_summary)
+expplts <- plot_mc(exp_summary)
 
-save(sexpplt, file = "../external_data/shorespd/sexpplt.rda")
+save(expplts, file = "../external_data/shorespd/expplts.rda")
 
 ##### Monte Carlo simulation - Logistic #####
 
@@ -299,9 +299,9 @@ log_summary <- simulation_summary(SPD, simulation_results,
 plotSimulationSummary(log_summary)
 
 # Custom plot
-slogplt <- plot_mc(log_summary)
+logplts <- plot_mc(log_summary)
 
-save(slogplt, file = "../external_data/shorespd/slogplt.rda")
+save(logplts, file = "../external_data/shorespd/logplts.rda")
 
 ##### Monte Carlo simulation - Uniform #####
 
@@ -393,11 +393,11 @@ uni_summary <- simulation_summary(SPD, simulation_results,
 plotSimulationSummary(uni_summary)
 
 # Custom plot
-suniplt <- plot_mc(uni_summary)
+uniplts <- plot_mc(uni_summary)
 
-save(suniplt, file = "../external_data/shorespd/suniplt.rda")
+save(uniplts, file = "../external_data/shorespd/uniplts.rda")
 
-sexpplt + slogplt + suniplt
+expplts + logplts + uniplts
 
 ggsave(here::here("analysis/figures/shoreline_mc.png"),
        units = "px", width = 4000, height = 1250)

### 2023-06-30 R/10_radiocarbon_cpl_admur.R

@@ -8,13 +8,15 @@ library(here)
 
 set.seed(42)
 
+source(here("R/03_functions.R"))
+
 c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
 c14 <- c14 %>% filter(context != "Food crust")
 
 names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
 
 minage <- 4500
-maxage <- 10975
+maxage <- 10555
 
 CalArray <- makeCalArray(intcal20, calrange = c(minage, maxage))
 PD <- phaseCalibrator(c14, CalArray, remove.external = TRUE)
@@ -101,13 +103,13 @@ mplt <- ggplot() +
     "Exponential" = "#4daf4a",
     "1-CPL" = "#984ea3",
     "2-CPL" = "#ff7f00",
-    "3-CPL" = "#ffff33",
+    "3-CPL" = "#e0d616",
     "4-CPL" = "#a65628",
     "5-CPL" = "black"
   )) +
   # scale_x_continuous(limit = c(-10000, -2750)) +
   # scale_y_continuous(limit = c(0, 0.001)) +
-  scale_x_reverse(labels = function(x)(x-2000)*-1) +
+  scale_x_reverse(labels = function(x)(x-1950)*-1) +
   theme_bw() +
   theme(legend.title = element_blank())
 
@@ -135,12 +137,12 @@ plotSimulationSummary(expsum)
 plotSimulationSummary(logsum)
 plotSimulationSummary(unisum)
 
-rexpplot <- plot_mc(expsum)
-rlogplot <- plot_mc(logsum)
-runiplot <- plot_mc(unisum)
+expplotr <- plot_mc(expsum)
+logplotr <- plot_mc(logsum)
+uniplotr <- plot_mc(unisum)
 
-(sexpplt + slogplt + suniplt)/
-(rexpplot + rlogplot + runiplot) +
+(expplts + logplts + uniplts)/
+(expplotr + logplotr + uniplotr) +
   plot_annotation(tag_levels = "A")
 
 ggsave(here::here("analysis/figures/mc.png"),

### 2023-07-03 analysis/paper/paper.Rmd
Edits from Steinar

@@
 # author:
-# - 'Isak Roalkvam'
-# - 'University of Oslo, Institute of Archaeology, Conservation and History'
-  # - IsakRoalkvam:
-  #     email: isak.roalkvam@iakh.uio.no
-  #     institute: [iakh]
-      # correspondence: true
-#   - name: Steinar Solheim
-#     email: steinar.solheim@khm.uio.no
-#     institute: [mch]
-#     correspondence: false
+#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
+#   - Steinar Solheim, University of Oslo, Museum of Cultural History
+author:
+  - Isak Roalkvam:
+      email: isak.roalkvam@iakh.uio.no
+      institute: [iakh]
+      correspondence: true
+  - name: Steinar Solheim
+    email: steinar.solheim@khm.uio.no
+    institute: [mch]
+    correspondence: false
 institute:
-  - iakh: Institute of Archaeology, Conservation and History, University of Oslo
-#   - mch: Museum of Cultural History, University of Oslo
+  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
+  - mch: University of Oslo, Museum of Cultural History
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       pandoc_args:
-      # - --lua-filter=../templates/scholarly-metadata.lua
-      # - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
+      - '--lua-filter=../templates/scholarly-metadata.lua'
+      - '--lua-filter=../templates/author-info-blocks.lua'
+      - '--lua-filter=../templates/pagebreak.lua'
 header-includes:
@@
 # Introduction
-Population size is regarded as an important driver of cultural variation, and is of critical importance to our understanding of past human societies [e.g.
-@french2021; @shennan2000].
-The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g.
-@crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].
-The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach.
+Variation in population patterns are regarded important drivers and consequences of cultural variation, and of critical importance to our understanding of past human societies [e.g.
+@french2021; @shennan2000; @shennan2020].
+During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022].
+This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].
+
+The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach.
 Several limitations and forms of criticism have, however, been directed at these procedures.
@@
 @carleton2021; @torfing2015; @ward2021].
-What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021].
-This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).
+What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021].
+In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
+We discuss how the data compare with each other as well as how it can be used to understand cultural historical development in the region through the Mesolithic, c.
+9500--4000 BCE.
 
 # Background
-Large parts of the post-glacial landscape of Northern Scandinavia is characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g.
+Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g.
 @mörner1979].
-As coastal foragers appear to have predominantly settled on or close to the contemporaneous shoreline, this can be utilised to assign an approximate date to the sites.
+The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use.
+As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites.
 This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating.
-This is not least useful for dating the large number of surveyed sites in the region where other temporal data that often follow with an excavation are not available, such as radiocarbon dates or typological indicators in artefact inventories.
+method has a long tradition in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites when temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
+Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jørgensen2020].
 
-The frequency of shoreline dated sites has also been compared to RSPDs in the past [@solheim2018; @tallavaara2020].
-However, this has been done by finding point estimates of shoreline dates that are then aggregated in somewhat arbitrary bins of 200 or 500 years.
-This therefore does not take into account the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
+The frequency of shoreline dated sites has been compared to RSPDs in the past by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years.
+This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
 In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites.
 The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)).
-The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility.
+The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility.
 Thus, this geographical and temporal limit are also used for this study.
-
-This paper is decidedly exploratory.
-While some speculative thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses.
-The data have thus been explored using a set of standard models to begin to unpick patterns and grasp the relationship between the variables.
-Furthermore, the dates as data approach is dependent on there being a direct link between the past generation of material that ultimately become ^14^C samples, and population size.
-The sum of shoreline dated sites, on the other hand, is determined by site frequency, and if used as a proxy for population size is therefore dependent on there being a connection between site count and population size.
+The present paper is decidedly exploratory.
+While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses.
+Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables.
+Furthermore, the dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples, and reflect population dynamics [cf- @freeman2018].
+The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population size and density.
 If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics.
@@
 
-To what degree the radiocarbon record is determined by past population numbers might vary both geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations.
+To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations.
 One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations.
-Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017].
+Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to consitute a 'site'.
 However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
-Before any such fluctuations are given any substantive interpretation, however, there are a host of methodological issues that have to be considered.
+Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @grocutt2020; @crema2022 for a review].
 
 The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure.
-This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022].
-As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model.
+This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022].
+As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model.
 It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021].
-This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021].
+This means that visual inspections of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022].
 While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.
 
-The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, as introduced by @shennan2013 and later expanded upon by @timpson2014.
+The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2020.
 This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model.
@@
 The result from these simulations are then used to create a 95% critical envelope representing the null model.
-The proportion of the observed RSPD that falls outside this envelope is then used to estimate a global p-value indicating whether the null model can be rejected.
+The proportion of the observed RSPD that falls outside the simulated envelope (i.e.
+positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected.
 In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model.
@@
 While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine.
-Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting local deviations themselves as meaningful demographic signals, as has often been the case [see review by @crema2022; @timpson2021].
+Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting if local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].
 
-## Model comparison and continous piece-wise linear models
+## Model comparison and continuous piece-wise linear models
 
 The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
-This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the verall development of this proxy could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves.
+This can be helpful for determining whether the population patterns has followed the trajectories of standard theoretical models.
+It is also a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the overall development of this proxy could be made.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and, as mentioned above, deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.
 As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*.
@@
 The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction.
-This therefore allows for the identification of key events and a conservative description of population dynamics between these.
-The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
-Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.
+This allows for the identification of key demographic events and a conservative description of population dynamics between these.
+The approach thus provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
+Following the evaluation of standard models as outlined above, the data were subsequently subjected to the approach presented by @timpson2021.
 
@@
 
-```{r map, echo = FALSE, fig.cap = "Map of the study area and shoreline dated sites.
+```{r map, echo = FALSE, fig.cap = "A) Map displaying the location of the study area and shoreline dated sites.
 Black lines indicate the borders between municipalities.
-The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1.", out.width = "500px", fig.align="center", regfloat = TRUE}
+The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1.
+B) The distribution of radiocarbon dates employed in the study.", out.width = "500px", fig.align="center", regfloat = TRUE}
 
@@
 To account for investigatory bias that can result from variable sampling intensity between sites, the summing of radiocarbon dates with *ADMUR* starts with binning dates from the same sites that fall within 200 uncalibrated ^14^C years of each other when measured from the mean ^14^C age.
-All dates are then calibrated using a resoltuion of 5 years and then normalised to sum to unity.
-The RSPDs for each site-phase are then summed to achieve the final SPD, which is in turn normalised.
+All dates are then calibrated using a resolution of 5 years and normalised to sum to unity.
+The RSPDs for each site-phase are summed to achieve the final SPD, which is in turn normalised.
 All calibrations were performed using the IntCal20 calibration curve [@reimer2020].
@@
 The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
-The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.
+The degree to which the observed RSPD deviates from the critical envelope is used to calculate a global p-value, indicating whether the null model can be rejected.
 
@@
 These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987].
-The first step in the dating procedure is to interpolate the shoreline displacement to the location to be dated.
+The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated.
 This is done by inverse distance weighting [e.g.
-@conolly2020], interpolating the relative sea-level change to the location of a site to be dated, based on it's distance from the isobases of the displacement curves.
+@conolly2020], interpolating the relative sea-level change to the location of an archaeological site, based on its distance from the isobases of the displacement curves.
 
@@
 With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline.
-Figure \@ref(fig:shoredate)C indicates where the elevation of the example site intersects the interpolated displacement curve.
-As it can be reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
-However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate.
-However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
+Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve.
+As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
+A common approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate.
+However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
 Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
@@
 
-The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833.
+The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833.
 This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D).
 When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve.
-This procedure thus gives the the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details].
+This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details].
 Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf.
@@
 
-The implications of different approaches to shoreline dating and summing the results is further illustrated in Figure \@ref(fig:sumshore).
+The implications of different approaches to shoreline dating and summing the results are illustrated in Figure \@ref(fig:sumshore).
 In Figure \@ref(fig:sumshore)A  the sites to be shoreline dated are plotted according to their elevation.
-Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution.
-For example, as the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east.
+Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution.
+For example, as the trajectory for sea-level regression is more pronounced towards the north-east, the same elevation at a location to the south-west implies a younger date compared to one situated further north-east.
 
@@
 This approach is illustrated in Figure \@ref(fig:sumshore)B.
-The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c.
+The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c.
 8000 BCE.
@@
 
-However, in the same way is with RSPDs, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above.
+In the same way is with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above.
 Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site.
-As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase.
-With analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve.
+As the shoreline displacement curves are interpolated to the sites based on their location along a south-west/north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase.
+With analogy to the radiocarbon methodology, this would be equivalent to each radiocarbon date being associated with a unique calibration curve.
 As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites.
-These intervals were then assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).
+These intervals were assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).
 
@@
 
-The Monte Carlo simulations with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice.
+The Monte Carlo simulations with the SSPD is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice.
 This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021].
 Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval.
-The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm.
+The calendar date is 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm.
 As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age.
-Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was then shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples.
+Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples.
 Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times.
@@
 For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected.
-Based on the available data, this therefore represents the best explanation of the data.
+Based on the available data, this model represents the best explanation of the data.
 For the SSPD all of the standard models could be rejected.
@@
 <!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
-
-Comparing the the directions of the three standard models, Figure \@ref(fig:spd_mc) appears to indicate an overall uniform distribution in the frequency of radiocarbon dates following the initial jump just before 8000 BCE, contrasted to a general decrease in shoreline dated sites after the initial rapid increase from around 9500 BCE.
+A notable pattern in Figure \@ref(fig:spd_mc) is the lack of data in the RSPD before c.
+8500--8300 BCE while the SSPD has a marked peak around 9000 BCE.
+Comparing the the directions of the standard models in the figure, the logistic model fit to the RSPD in Figure \@ref(fig:spd_mc) indicates an overall uniform distribution in the frequency of radiocarbon dates following the initial jump just before 8000 BCE.
+This is contrasted to a general decrease in shoreline dated sites after the initial rapid increase from around 9500 BCE.
+Although they could be rejected as explaining the data, this is indicated by the decaying exponential function (Figure \@ref(fig:spd_mc)A) and the initial under-prediction followed by regions of over-prediction with the logistic and uniform models (Figure \@ref(fig:spd_mc)B and C).
 
 It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a drop just before 6000 BCE.
-However, given that the logistic model cannot be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and therefore no analytical weight can be placed on these deviations -- at least given the present data and modelling efforts.
+However, given that the logistic model cannot be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and no analytical weight can be placed on these deviations -- at least given the present data and modelling efforts.
 
@@
 
-It is firmly established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g.
-@glorstad2016; @damlien2018b].
+It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g.
+@damlien2018b; @glorstad2016].
 While the SSPD starts from c.
-9500 BCE, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand.
-Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years, ranging from taphonomic loss to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a lack of reliable access to firewood.
-While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump in ^14^C-dates at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind this pattern.
+9500 BCE, the RSPD show no evidence of human activity until c.
+8500–8300 BCE.
+What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years.
+These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs for fuel or use of rapidly deteriorating seal blubber for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [e.g.
+@damm2022].
+While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind the observed pattern.
 
 The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
-The most immediate explanation of this mismatch would be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+An immediate explanation of this mismatch could be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+A common interpretation of the Early Mesolithic in the region is that the population was low in numbers but highly mobile, reflected in large numbers of short lived sites along the coast [e.g.
+@bjerck2017].
+This is commonly seen in relation to a rapid colonisation of the Norwegian coastline and homogeneous lithic technology over large parts of Scandinavia in this period.
+High mobility has been seen as necessary both to maintain demographic viability among colonising groups and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017].
+The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
+At the same time, the low/lacking signal in the RSPD for this earliest period could be taken as evidence for low population numbers during the first settlement period following the deglaciation.
+
+@berg-hansen2022 have argued that there is an increase in sites between c.
+8000--7000 BCE in the northern part of the study area of this paper.
+However, their analysis covers a limited spatial area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
+Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
+In contrast to their findings, the analysis undertaken here instead indicates a rapid increase followed drop in the number of sites through this time-interval, possibly corresponding with a shift to reduced residential mobility [@roalkvam2022; @damlien2018b; cf.
+@binford1980].
+
+As evidenced by both genetic data and technological analyses of lithic inventories, the initial phase of human colonisation from the south appears to be followed by an influx, and possible mixing, of people from the east at the transition to the Middle Mesolithic period around 8300 BCE [@gunther2018; @manninen2021].
+During the first part of the Middle Mesolithic, it has been suggested that the mobility patterns shifted to a more stable settlement, confined to larger regions [@damlien2018b].
+If this narrative is coupled with the data observed here, these migratory events would appear to result in a relatively sudden drop in site-count as indicated in the SSPD.
+The end of this drop corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes rather than taphonomic factors.
 
 The development after 4000 BCE should be treated with further care, following from questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023].
-
-The initial peak in the SSPD would be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
-As evidenced by both genetic data and technological analyses of lithic inventories, this initial phase of human occupation appears to be followed by an influx of people from the .
-If this narrative is coupled to the data observed here, this process would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, while the end of this drop corresponds to the first ^14^C-dates in the RSPD -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is related to cultural processes.
-
-@berg-hansen2022 have argued that there is an increase in sites in the first half of the MM in the northern part of the study area of this paper.
-However, their analysis is seemingly also based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
-Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
-In contrast to their findings, the analysis undertaken here instead indicates a drop in the number of sites through this time-interval.
 
@@
 @barton2014; @clark2017; @preston2018; @roalkvam2022].
-Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g.
+Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g.
 @jorgensen2020; @ordonez2022].
+Furthermore, mapping and constrasting these developments across a large region will offer important contextualisation for the developments observed here.
 

### 2023-07-21 analysis/paper/paper.Rmd
Setting up GOF with 4-CPL

@@
 Variation in population patterns are regarded important drivers and consequences of cultural variation, and of critical importance to our understanding of past human societies [e.g.
-@french2021; @shennan2000; @shennan2020].
+@french2021; @shennan2000; @shennan2021].
 During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022].
@@
 method has a long tradition in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites when temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
-Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jørgensen2020].
+Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jorgensen2020].
 
@@
 However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
-Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @grocutt2020; @crema2022 for a review].
+Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2020; @crema2022 for a review].
 
@@
 It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021].
-This means that visual inspections of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022].
+This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022].
 While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.
 
-The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2020.
+The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021.
 This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model.
 These null-models are typically a uniform, exponential or logistic distribution.
-These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE).
+These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or maximum likelihood estimation (MLE).
 The result from these simulations are then used to create a 95% critical envelope representing the null model.
@@
 While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine.
-Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting if local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].
+Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].
 
@@
 ```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.
-B) The geological displacement curves in the region and the curve interpolated to the example site.
+B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site.
 C) The black line on the y-axis represents the elevation of the site.
@@
 For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected.
-Based on the available data, this model represents the best explanation of the data.
+Based on the available data, this model therefore represents the best explanation of the data.
+Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates.
+
 For the SSPD all of the standard models could be rejected.
+The result of subjecting the SSPD to the ADMUR workflow identified 4-CPL as the best model.
+An important caveat here is that models with a complexity beyond 4-CPL consistently failed to converge.
+A fuller reporting of this convergence issue, including the exploration of different parameter settings, different sample sizes and alternative optimization algorithms is given in the supplementary material.
+Subjecting the 4-CPL model to the goodness-of-fit test gave p =, meaning that the data nonetheless could be expected under this model.
+The accepted models that underlies the comparison to follow are therefore the logistic model for the RSDP and the 4-CPL for the SSPD.
 
@@
 @damm2022].
-While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind the observed pattern.
+While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump at the start of RSPD, combined with the magnitude of the SSPD in the period preceding this, suggest that cultural rather than environmental factors are central drivers behind the observed pattern.
 
 The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
-An immediate explanation of this mismatch could be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
-A common interpretation of the Early Mesolithic in the region is that the population was low in numbers but highly mobile, reflected in large numbers of short lived sites along the coast [e.g.
+An immediate explanation of this mismatch could be that the SSPD mainly reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short lived sites along the coast [e.g.
 @bjerck2017].
-This is commonly seen in relation to a rapid colonisation of the Norwegian coastline and homogeneous lithic technology over large parts of Scandinavia in this period.
+This is commonly seen in relation to homogeneous lithic technology over large parts of Scandinavia in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline.
 High mobility has been seen as necessary both to maintain demographic viability among colonising groups and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017].
 The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
-At the same time, the low/lacking signal in the RSPD for this earliest period could be taken as evidence for low population numbers during the first settlement period following the deglaciation.
+At the same time, the lacking signal in the RSPD for this earliest period could be taken as evidence for low population numbers during the first settlement period following the deglaciation.
 
@@
 8000--7000 BCE in the northern part of the study area of this paper.
-However, their analysis covers a limited spatial area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
+However, their analysis covers a limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
 Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
-In contrast to their findings, the analysis undertaken here instead indicates a rapid increase followed drop in the number of sites through this time-interval, possibly corresponding with a shift to reduced residential mobility [@roalkvam2022; @damlien2018b; cf.
-@binford1980].
 
-As evidenced by both genetic data and technological analyses of lithic inventories, the initial phase of human colonisation from the south appears to be followed by an influx, and possible mixing, of people from the east at the transition to the Middle Mesolithic period around 8300 BCE [@gunther2018; @manninen2021].
-During the first part of the Middle Mesolithic, it has been suggested that the mobility patterns shifted to a more stable settlement, confined to larger regions [@damlien2018b].
-If this narrative is coupled with the data observed here, these migratory events would appear to result in a relatively sudden drop in site-count as indicated in the SSPD.
+In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a drop in the number of sites through this time-interval.
+It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [@roalkvam2022; @damlien2018b].
+Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east at the transition to the Middle Mesolithic period around the same time [@gunther2018; @manninen2021].
+If this narrative is coupled with the data observed here, these migratory events would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
 The end of this drop corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes rather than taphonomic factors.
@@
 
-While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
+While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
 Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.

### 2023-07-21 analysis/supplementary/supplementary.Rmd
Setting up GOF with 4-CPL

---
title: "Supplementary material to the paper 'Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway'"
author: 
- 'Isak Roalkvam'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  bookdown::pdf_document2:
    toc: false
---

This document lays out the measures that were taken in an attempt to work around and explore the convergence issues related to identifying parameters for models beyond 4-CPL.

```{r setup}
# Load required packages
library(ADMUR)
library(DEoptimR)
library(pso)

# Set seed for reproducibility
set.seed(42)
```

```{r data}
# Load data holding the SSPD
load(file = here::here("analysis/data/derived_data/shore_pd_models.RData"))

```

```{r data, echo = FALSE}
# For the compilation of this Rmarkdown document, the results were pre-compiled 
# and are loaded here. Set eval = TRUE in code chunks below to re-run the code.
load(file = here::here("analysis/data/derived_data/shore_pd_models.RData"))

```

This section gives the result of running JDEoptimR on the SSPD to identify 4-CPL, 5-CPL and 6-CPL. For 5-CPL and 6-CPL both the maxint and NP parameters were increased beyond the minimum recommended settings in an attempt to achieve convergence.

```{r jdeoptim, eval = FALSE}
  cpl_4 <- JDEoptim(lower = rep(0, 7),
                     upper = rep(1, 7), 
                     fn = objectiveFunction,
                     PDarray = pd,
                     type = 'CPL',
                     NP = 140)

  cpl_5 <- JDEoptim(lower = rep(0, 9), 
                    upper = rep(1, 9), 
                    fn = objectiveFunction,
                    PDarray = pd,
                    type='CPL',
                    NP = 250, 
                    maxiter = 400 * 9)

  cpl_6 <- JDEoptim(lower = rep(0, 11),
                       upper = rep(1,11),
                       fn = objectiveFunction,
                       PDarray = pd,
                       type = 'CPL',
                       NP = 300, 
                       maxiter =  400 * 11)

```

In the output the value in the column 'convergence' indicates success with 0 and failure with 1, meaning that while 4-CPL converged without any issues, both 5-CPL and 6-CPL failed to converge.

To check if the issue could be related to the scale of the values in the pd, the same procedure was ran with the probabilities subjected to min-max normalisation to adjust these to take on values between 0 and 1.

```{r minmax, eval = FALSE}

```

As a subsequent step, the differential evolution optimization algorithm from DEoptimR was replaced with particle swarm optimization using the pso package.
The value for convergence is here 0 with successful convergence, and 2 if convergence was not achieved before reaching the maximum number of iterations (see ?psoptim).

```{r pso, eval = FALSE}

cpl4_pso <- psoptim(par = rep(NA, 7),
                    lower = rep(0, 7),
                    upper = rep(0.9999999, 7), # Upper bound causes error at 1
                    fn = objectiveFunction,
                    PDarray = pd,
                    type = 'CPL',
                    control = list(maxit = 20000))

cpl5_pso <- psoptim(par = rep(NA, 9),
                    lower = rep(0, 9),
                    upper = rep(0.9999999, 9),
                    fn = objectiveFunction,
                    PDarray = pd,
                    type = 'CPL',
                    control = list(maxit = 20000))

cpl6_pso <- psoptim(par = rep(NA, 11),
                    lower = rep(0, 11),
                    upper = rep(0.9999999, 11),
                    fn = objectiveFunction,
                    PDarray = pd,
                    type = 'CPL',
                    control = list(maxit = 20000))
```

As a final step and to explore the behaviour of JDEoptim() with shoreline dates, a series of models were fit to an increasing number of randomly sampled shoreline dates.

```{r sample, eval = FALSE}
sampsize <- c(6, 32, 60, 120, 240, 480)

results <- list()
for(i in 1:2){
  tmppd <- pd[, sample(ncol(pd), sampsize[i])]

  cpl4 <- JDEoptim(lower = rep(0, 7),
                      upper = rep(1, 7),
                      fn = objectiveFunction,
                      PDarray = tmppd,
                      type = 'CPL',
                      NP = 140)

  cpl5 <- JDEoptim(lower = rep(0, 9),
                      upper = rep(1, 9),
                      fn = objectiveFunction,
                      PDarray = tmppd,
                      type='CPL',
                      NP = 180,
                      maxiter = 400 * 9,
                      trace = TRUE)

  cpl6 <- JDEoptim(lower = rep(0, 11),
                      upper = rep(1,11),
                      fn = objectiveFunction,
                      PDarray = tmppd,
                      type = 'CPL',
                      NP = 220,
                      maxiter = 400 * 11,
                      trace = TRUE)

  results[[i]] <- rbind(cpl4, cpl5, cpl6)
}
```

```{r}
# Print results
for(i in 1:length(results)){
  cat(paste("\nSample size =", sampsize[i], "\n"))
  print(results[[i]])
}
```

4-CPL successfully converged for all sample sizes, while 5-CPL and 6-CPL consistently failed to converge.

### 2023-07-21 R/01_shoreline_spd.R

@@ -194,14 +194,14 @@ sa <- ggplot() +
   geom_sf(data = st_transform(muncipalities, st_crs(sites)), fill = NA,
           colour = "black", lwd = 0.25) +
   geom_sf(data = st_centroid(surveyed), aes(fill = lab),
-          size = 2, shape = 21,
+          size = 1.5, shape = 21,
           colour = "black", show.legend = "point") +
   geom_sf(data = st_centroid(dplyr::filter(surveyed, quality < 4)),
           aes(fill = lab),
-          size = 2, shape = 21,
+          size = 1.5, shape = 21,
           colour = "black", show.legend = "point") +
   geom_sf(data = st_centroid(excshore), aes(fill = "y"),
-          size = 2, shape = 21,
+          size = 1.5, shape = 21,
           colour = "black", show.legend = "point") +
   scale_fill_manual(labels = c(paste0("Surveyed sites,\nincluded (n = ",
                                       nrow(dplyr::filter(surveyed,
@@ -241,7 +241,7 @@ pltc14 <- ggplot() +
   geom_sf(data = st_transform(muncipalities, st_crs(sites)), fill = NA,
           colour = "black", lwd = 0.25) +
   geom_sf(data = st_centroid(c14_sites), aes(fill = lab),
-          size = 2, shape = 24,
+          size = 1.5, shape = 24,
           colour = "black", show.legend = "point") +
   scale_fill_manual(values = c("y" = "red"),
     labels = c(paste0("Sites with radiocarbon dates (n = ",
@@ -265,10 +265,16 @@ pltc14 <- ggplot() +
 
 plt1 <- cowplot::ggdraw() +
   cowplot::draw_plot(sa) +
-  cowplot::draw_plot(overview, x = 0.069,
+  cowplot::draw_plot(overview, x = 0.179,
                      y = 0.7, width = 0.3, height = 0.3)
 
-cowplot::plot_grid(plt1, pltc14,ncol = 1, labels = c('A', 'B'))
+(plt1)/
+  pltc14 + plot_annotation(tag_levels = "A")
+
+cowplot::plot_grid(plt1, pltc14, ncol = 1)
+
+ggsave(filename = here("analysis/figures/map2.png"),
+       units = "px", width = 4500, height = 4500)
 
 # Example date plot"
 

### 2023-07-21 R/03_functions.R

@@ -166,8 +166,8 @@ plot_mc <- function(mc_summary, xppos, yppos, title = NULL) {
     geom_line(aes(y = model), linewidth = 0.5, col = "red") +
     geom_line(aes(y = SPD)) +
     labs(x = "BCE", y = "Summed probability", title = title) +
-    scale_x_reverse(limits = c(12000, 4500),
-                    breaks = seq(12000, 4500, -1000),
+    scale_x_reverse(limits = c(11950, 4450),
+                    breaks = seq(11950, 4450, -1000),
                     expand = expansion(mult = c(0, 0)),
                     labels = function(x)(x-1950)*-1) +
     geom_text(aes(11000, spdadj), label = plabel) +

### 2023-07-21 R/0x_shore_4cpl_gof.R

@@ -1,0 +1,128 @@
+library(shoredate)
+
+set.seed(1)
+
+source(here("R/03_functions.R"))
+
+# Load polygons with displacement curves
+load(here("analysis/data/derived_data/displacement_polygons.RData"))
+
+# Load shoreline dates
+load(here("analysis/data/derived_data/sdates.RData"))
+
+# Load CPL models
+load(file = here::here("analysis/data/derived_data/shore_models.RData"))
+
+cpl4 <- convertPars(pars = cpl_4$par, years = minage:maxage, type = 'CPL')
+cpl4$model <- "4-CPL"
+
+cpl4$year <- (cpl4$year - 1950) * -1
+names(cpl4) <- c("bce", "prob_dens", "model")
+
+# Sum the probability distributions for the shoreline dates
+sumdates <- sum_shoredates(shorelinedates)
+sumdatesdf <- as.data.frame(sumdates) %>%
+  filter(sum.probability != 0)
+
+# Restrict SSPD and normalise probability
+sumdatesdf <- filter(sumdatesdf, sum.bce <= max(cpl4$bce))
+sumdatesdf$probability <- sumdatesdf$sum.probability /
+  (sum(sumdatesdf$sum.probability) * 5)
+
+ggplot() +
+  geom_line(data = sumdatesdf, aes(x = sum.bce, y= probability)) +
+  geom_line(data = cpl4, aes(x = bce, y = prob_dens))
+
+# Find density of sites across polygons
+incpolys$dens <- lengths(st_intersects(incpolys, sites)) /
+  sum(lengths(st_intersects(incpolys, sites)))
+
+# Sample size
+ssize = sumdates$dates_n
+
+# Number of simulations
+nsim <- 10000
+
+##### Monte Carlo simulation - 4-CPL #####
+
+# Create data frame to hold simulated dates
+random_dates <- data.frame(matrix(nrow = ssize*nsim, ncol = 3))
+names(random_dates) <- c("sample", "simn", "displacement_curve")
+
+random_dates$sample = sample(cpl4$bce, replace = TRUE,
+                             size = ssize*nsim, prob = cpl4$prob_dens)
+random_dates$simn <- rep(1:nsim, each = nrow(random_dates)/nsim)
+random_dates$displacement_curve <- incpolys[sample(incpolys$id,
+                                                   replace = TRUE,
+                                                   size = ssize*nsim,
+                                                   prob = incpolys$dens),]$disp[[1]]
+
[…]

### 2023-07-21 R/0x_shoreline_cpl.R

@@ -1,11 +1,13 @@
 library(ADMUR)
 library(DEoptimR)
+library(ggplot2)
+library(patchwork)
 
 set.seed(42)
 
-# Load data from radiocarbon_mc.R: models fit for MC, the pd and the min and
-# max ages
-load(file = here::here("analysis/data/derived_data/shore_pd_models.RData"))
+# Load data from shoreline_monte_carlo.R: models fit for MC, the pd and the
+# min and max ages
+load(file = here::here("analysis/data/derived_data/shore_models_pd.RData"))
 
 # Fit CPL-models to shoreline dates
 cpl_1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction, PDarray = pd,
@@ -26,16 +28,10 @@ cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
                   PDarray = pd, type = 'CPL', NP = 220, maxiter =  400 * 11,
                   trace = TRUE)
 
-save(exp, logi, unif, cpl_1, cpl_2, cpl_3, cpl_4,
+save(cpl_1, cpl_2, cpl_3, cpl_4,
      file = here::here("analysis/data/derived_data/shore_models.RData"))
 load(file = here::here("analysis/data/derived_data/shore_models.RData"))
 
-expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
-expp$model <- "Exponential"
-logip <- convertPars(pars = logi$par, years = minage:maxage, type = 'logistic')
-logip$model <- "Logistic"
-unifp <- convertPars(pars = NULL, years = minage:maxage, type = "uniform")
-unifp$model <- "Uniform"
 cpl1 <- convertPars(pars = cpl_1$par, years = minage:maxage, type = 'CPL')
 cpl1$model <- "1-CPL"
 cpl2 <- convertPars(pars = cpl_2$par, years = minage:maxage, type = 'CPL')
@@ -65,28 +61,24 @@ sh_bic <- c(log(ssize)*1 - 2*sh_lik[1], # Exponential
                  log(ssize)*7 - 2*sh_lik[7])
 
 SPD <- as.data.frame(rowSums(pd))
-SPD <- SPD/( sum(SPD) *5 )
+SPD <- SPD/(sum(SPD) * 5)
 
 cplplt <- ggplot() +
-  geom_bar(aes(x = (as.numeric(rownames(SPD)) - 2000) * -1,
[…]

### 2023-07-21 R/0x_shoreline_monte_carlo2.R

@@ -396,11 +396,3 @@ plotSimulationSummary(uni_summary)
 uniplts <- plot_mc(uni_summary)
 
 save(uniplts, file = "../external_data/shorespd/uniplts.rda")
-
-expplts + logplts + uniplts
-
-ggsave(here::here("analysis/figures/shoreline_mc.png"),
-       units = "px", width = 4000, height = 1250)
-
-
-

### 2023-07-21 R/10_radiocarbon_cpl_admur.R

@@ -25,8 +25,6 @@ SPD <- as.data.frame( rowSums(PD) )
 # normalise
 SPD <- SPD/( sum(SPD) * CalArray$inc )
 
-
-# CPL parameters must be between 0 and 1, and an odd length.
 CPL.1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction,
                   PDarray = PD, type = 'CPL', NP = 20)
 CPL.2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3),
@@ -140,6 +138,10 @@ plotSimulationSummary(unisum)
 expplotr <- plot_mc(expsum)
 logplotr <- plot_mc(logsum)
 uniplotr <- plot_mc(unisum)
+
+load("../external_data/shorespd/expplts.rda")
+load("../external_data/shorespd/logplts.rda")
+load("../external_data/shorespd/uniplts.rda")
 
 (expplts + logplts + uniplts)/
 (expplotr + logplotr + uniplotr) +

### 2023-07-21 R/exploring.r

@@ -1,0 +1,353 @@
+library(ggplot2)
+library(ADMUR)
+library(DEoptimR)
+
+load(file = here::here("analysis/data/derived_data/shore_pd_models.RData"))
+
+SPD <- as.data.frame( rowSums(pd) )
+# normalise
+SPD <- SPD/( sum(SPD) * 5 )
+plotPD(SPD)
+
+library(pso)
+
+normalize <- function(x, na.rm = TRUE) {
+  return((x- min(x)) /(max(x)-min(x)))
+}
+
+mmpd <- as.data.frame(apply(pd, 2, normalize))
+
+CPL.1pso <- psoptim(par = NA,
+                    lower = 0,
+                    upper = 0.999999,
+                    fn = objectiveFunction,
+                    PDarray = pd, type = 'CPL',
+                    control = list(trace = 1))
+CPL.2pso <- psoptim(par = rep(NA, 3),
+                    lower = rep(0, 3),
+                    upper = rep(0.999999, 3),
+                    fn = objectiveFunction,
+                    PDarray = pd,
+                    type = 'CPL',
+                    control = list(trace = 1))
+CPL.3pso <- psoptim(par = rep(NA, 5),
+                    lower = rep(0, 5),
+                    upper = rep(0.999999, 5),
+                    fn = objectiveFunction,
+                    PDarray = pd,
+                    type = 'CPL',
+                    control = list(trace = 1))
+
+CPL1pso <- convertPars(pars=CPL.1pso$par, years=minage:maxage, type='CPL')
+CPL2pso <- convertPars(pars=CPL.2pso$par, years=minage:maxage, type='CPL')
+CPL3pso <- convertPars(pars=CPL.3pso$par, years=minage:maxage, type='CPL')
+
+ggplot() +
+  geom_bar(aes(x = as.numeric(rownames(SPD)),
+               SPD[,1]),
+           stat = "identity", col = "grey") +
+  geom_line(aes(x = as.numeric(rownames(SPD)),
+                y = SPD[,1])) +
+  geom_line(data = CPL1pso, aes(year, pdf),
+            linewidth = 1) +
+  geom_line(data = CPL2pso, aes(year, pdf), col = "red",
+            linewidth = 1) +
[…]

### 2023-07-27 analysis/paper/paper.Rmd
Discussion

@@
 In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
-We discuss how the data compare with each other as well as how it can be used to understand cultural historical development in the region through the Mesolithic, c.
+We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c.
 9500--4000 BCE.
@@
 Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g.
-@mörner1979].
+@mörner1979; @steffen2011].
 The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use.
@@
 This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating.
-method has a long tradition in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites when temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
+Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
 Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jorgensen2020].
 
-The frequency of shoreline dated sites has been compared to RSPDs in the past by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years.
+Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years.
 This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
 In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites.
-The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)).
+The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)).
 The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility.
@@
 Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables.
-Furthermore, the dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples, and reflect population dynamics [cf- @freeman2018].
-The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population size and density.
+The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf.
+@freeman2018].
+The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density.
 If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics.
@@
 One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations.
-Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to consitute a 'site'.
+Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to constitute a 'site'.
 However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
@@
 
-The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure.
+The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a statistically coherent procedure.
 This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022].
@@
 
-## Model comparison and continuous piece-wise linear models
+The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
+This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites done here, as few assumptions concerning this overall development could be made.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.
 
-The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
+<!-- ## Model comparison and continuous piece-wise linear models -->
+
+<!-- The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
 This can be helpful for determining whether the population patterns has followed the trajectories of standard theoretical models.
@@
 Following the evaluation of standard models as outlined above, the data were subsequently subjected to the approach presented by @timpson2021.
+-->
 
@@
 All sites with a quality score of 4 or worse were excluded from further analysis.
-Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded.
+Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded.
 This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016].
@@
 Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis.
-The 102 excavated sites in the dataset without relevant ^14^C-dates and that were originally shoreline dated in the reports were included in the SSPD along with the retained surveyed sites.
-This gave a total of  shoreline dated sites in the final SSPD.
+The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites.
+This gave a total of 921 shoreline dated sites in the final SSPD.
 
@@
 Radiocarbon dates were taken from @roalkvam2023 and @solheim.
-Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates.
+Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a total of ^14^C-dates.
 Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c.
 2500 BCE, the Mesolithic is the main focus of this analysis.
-Consequently, probabilities falling later than 2500 BCE were excluded from the analysis.
-Probabilities falling after 4000 BCE and the first, possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.
+Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, further reducing the sample of ^14^C-dates to.
+Probabilities falling after 4000 BCE and the first possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.
 
@@
 This is done by inverse distance weighting [e.g.
-@conolly2020], interpolating the relative sea-level change to the location of an archaeological site, based on its distance from the isobases of the displacement curves.
+@conolly2020], where the relative sea-level change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.
 
@@
 
-## Continuous piece-wise linear models
-The procedures outlined above represent standard approaches when modelling population dynamics in archaeology.
+<!-- ## Continuous piece-wise linear models -->
+<!-- The procedures outlined above represent standard approaches when modelling population dynamics in archaeology.
 These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shore-bound sites over time could be made.
@@
 Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.
+-->
 
@@
 
-For each dating method, the relative performance of the exponential, logistic, uniform and CPL models was compared.
+For each dating method, the relative performance of the exponential, logistic and uniform models was compared.
 This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*.
@@
 ```
-For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected.
-Based on the available data, this model therefore represents the best explanation of the data.
-Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates.
+For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected.
+As is also indicated by the BIC values, this model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development.
+This is characterised by a rapid growth from the start of the RSPD which plateaus at around 7500 BCE and then remains stable for the remainder of the period.
 
-For the SSPD all of the standard models could be rejected.
+<!-- Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates.
+-->
+
+All of the standard models could be rejected for the SSPD.
+Of the contrasted models, the BIC values indicate that the exponential model is the best candidate.
+The negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development.
+Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c.
+X--9000 BCE, followed by a significant positive peak just after 9000 BCE.
+The process of decline is then interrupted by positive deviations at, and with a smaller deviation around 7000 BCE.
+This is closely followed by a negative deviation of more magnitude than that expected under the exponential model between X and X BCE.
+Finally, a positive deviation occurs again at X--5000 BCE.
+While all of these deviations should be viewed with caution, the small magnitude and questions concerning the reliability of shoreline dating after 4000 BCE [cf.
+@roalkvam2023] means that the negative deviations that occur around 3000 BCE are disregarded here.
+
+<!-- For the SSPD all of the standard models could be rejected.
 The result of subjecting the SSPD to the ADMUR workflow identified 4-CPL as the best model.
@@
 The accepted models that underlies the comparison to follow are therefore the logistic model for the RSDP and the 4-CPL for the SSPD.
+-->
 
 <!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
-A notable pattern in Figure \@ref(fig:spd_mc) is the lack of data in the RSPD before c.
-8500--8300 BCE while the SSPD has a marked peak around 9000 BCE.
-Comparing the the directions of the standard models in the figure, the logistic model fit to the RSPD in Figure \@ref(fig:spd_mc) indicates an overall uniform distribution in the frequency of radiocarbon dates following the initial jump just before 8000 BCE.
-This is contrasted to a general decrease in shoreline dated sites after the initial rapid increase from around 9500 BCE.
-Although they could be rejected as explaining the data, this is indicated by the decaying exponential function (Figure \@ref(fig:spd_mc)A) and the initial under-prediction followed by regions of over-prediction with the logistic and uniform models (Figure \@ref(fig:spd_mc)B and C).
-
-It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a drop just before 6000 BCE.
-However, given that the logistic model cannot be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and no analytical weight can be placed on these deviations -- at least given the present data and modelling efforts.
 
@@
 @damlien2018b; @glorstad2016].
-While the SSPD starts from c.
-9500 BCE, the RSPD show no evidence of human activity until c.
-8500–8300 BCE.
+While the SSPD achieved here starts from c.
+9500 BCE, the RSPD shows no evidence of human activity until c.
+8500 BCE.
 What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years.
-These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs for fuel or use of rapidly deteriorating seal blubber for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [e.g.
-@damm2022].
-While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump at the start of RSPD, combined with the magnitude of the SSPD in the period preceding this, suggest that cultural rather than environmental factors are central drivers behind the observed pattern.
+These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022].
+While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic as opposed to taphonomic and environmental factors are central drivers behind the observed pattern.
 
-The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
-An immediate explanation of this mismatch could be that the SSPD mainly reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+The considered models indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
+An immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+
 A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short lived sites along the coast [e.g.
-@bjerck2017].
+@bjerck2008; @bjerck2017].
 This is commonly seen in relation to homogeneous lithic technology over large parts of Scandinavia in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline.
-High mobility has been seen as necessary both to maintain demographic viability among colonising groups and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017].
+High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017].
 The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
-At the same time, the lacking signal in the RSPD for this earliest period could be taken as evidence for low population numbers during the first settlement period following the deglaciation.
+The lacking signal in the RSPD for the first settlement period following the deglaciation could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf.
+@damm2022] and low population numbers.
 
-@berg-hansen2022 have argued that there is an increase in sites between c.
+Following this earliest phase, @berg-hansen2022 have argued that there is an increase in sites between c.
 8000--7000 BCE in the northern part of the study area of this paper.
@@
 
-In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a drop in the number of sites through this time-interval.
-It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [@roalkvam2022; @damlien2018b].
-Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east at the transition to the Middle Mesolithic period around the same time [@gunther2018; @manninen2021].
-If this narrative is coupled with the data observed here, these migratory events would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
-The end of this drop corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes rather than taphonomic factors.
+In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a decline in the number of shoreline dated sites through this time-interval, with possible positive reversals of this trend just after 8000 BCE and around 7000 BCE.
+It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [e.g.
+@damlien2018b; @roalkvam2022].
+Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@gunther2018; @manninen2021].
+If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
+The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes as opposed to taphonomic factors.
 
-The development after 4000 BCE should be treated with further care, following from questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023].
+The SSPD deviates negatively from all models from c.
+to c.
+BCE, and positively just before the 4th millennium BCE.
+The negative deviation could possibly been seen in relation to an increase in inland activity that occurs around the same time, which is argued to reach its peak around 6000 BCE.
+@boaz1999 related this development to a separation of inland and coastal populations in this period.
+The middle of the 6th millennium BCE marks the onset of the classic Nøstvet phase [@reitan2022], which has traditionally been seen as characterised by a reduction in residential mobility, and a specialisation towards fishing.
+This also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
+Given the rejected null-models, the stable development in RSPD over this period, and the limited sample of ^14^C-dates, it is, however, difficult to
+
+It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a negative deviation just before 6000 BCE.
+This could possibly be related to an increase in population numbers following, and a  However, given that the logistic model could not be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and so no analytical weight can be placed on the deviations in the RSPD given the present data.
+This is also poses a challenge for the interpretation of the SSPD.
+
+The logistic growth for the RSPD then indicates that a plateau is reached by the mid 7th millennium BCE and a stable population for the remainder of the Mesolithic.
 
@@
 
-While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
+While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
 Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
@@
 @barton2014; @clark2017; @preston2018; @roalkvam2022].
-Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g.
+Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
 @jorgensen2020; @ordonez2022].
-Furthermore, mapping and constrasting these developments across a large region will offer important contextualisation for the developments observed here.
+Aligning these different lines of evidence would also be dependent on a more direct modelling of the development in the SSPS, rather than simply basing the analysis on deviations from a rejected null model [cf.
+@timpson2021].
+Furthermore, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data, but could also provide important contextualisation for the developments observed here [see @lundstrom2021].
 
 Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application.
-While we still believe the approach represents an improvement over previous approaches, the results should therefore be treated with care.
-This also extends to the results of the RSPD, due to the fairly limited ^14^C sample.
+While we still believe the approach represents an improvement over previous approaches, the results should therefore considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the frequency of coastal sites in the region.
+
+<!-- This also extends to the results of the RSPD, due to the fairly limited ^14^C sample.
 With these limitations in mind, the approach outlined by @timpson2021 gives a more reasonable framework with which to evaluate the available data, as leads to a more conservative analysis and reduces the amount of over-interpretation that might follow from uncertainties to do with data or methods.
+-->
 
@@
 Although a more precise understanding of this relationship
-
-# Ekstra figurer
-
-Jeg fikk ADMUR til å funke med C14-dateringer.
-Figuren under viser at den logaritmiske modellen (blå kurve, samme som i Figur 5 over) ifølge BIC passer best av alternativene , og Monte Carlo-simuleringen over viser at den ikke kan avvises som forklarende for dataen.
-Eller med andre ord, dataen ville kunne forventes under den logaritmiske utviklingen.
-CPL modellering av strandlinjedateringene sliter jeg imidlertid mer med.
-Modellene opp til 4-CPL ser rimelige ut, men toppen på 5-CPL (sort kurve) som ligger litt før 5000 f.Kr.
-virker feil.
-I tillegg indikerer BIC-verdiene at denne passer best til dataen, som også virker rart gitt at den avviker såpass fra SSPDen.
-
-```{r cpl_rcarbon, echo = FALSE, out.width = "500px", fig.align="center", regfloat = TRUE}
-# fig.cap = "Model comparison, summed probability of radiocarbon dates.",
-knitr::include_graphics(here("analysis/figures/cplrcarbon.png"))
-```
-
-```{r cpl_shore, echo = FALSE, out.width = "500px", fig.align="center", regfloat = TRUE}
-# fig.cap = "Model comparison, summed probability of shoreline dates.",
-knitr::include_graphics(here("analysis/figures/cplshore.png"))
-```
-
-<!-- The following line inserts a page break  -->
 

### 2023-07-27 R/0x_shore_4cpl_gof.R

@@ -1,4 +1,7 @@
 library(shoredate)
+library(here)
+library(ADMUR)
+
 
 set.seed(1)
 

### 2023-07-27 R/0x_shoreline_cpl.R

@@ -82,3 +82,7 @@ bicplt <- ggplot() +
   theme_bw()
 
 cplplt + bicplt
+
+# Save pd, 4-CPL model and min and max age for mcmc
+save(cpl_4, pd, minage, maxage,
+     file = here("analysis/data/derived_data/shore_mcmc_data.RData"))

### 2023-07-27 R/0x_shoreline_monte_carlo2.R

@@ -287,9 +287,12 @@ simresults <- do.call(rbind.data.frame, results) %>%
     prob_sum_normalised = prob_sum / (sum(prob_sum, na.rm = TRUE) * 5))
 
 # Model summary (can also be plotted with ADMUR)
-mod <- approx(x = logip$bce, y = logip$prob_dens,
+mod <- approx(x = (logip$bce * -1) + 1950, y = logip$prob_dens,
               xout = as.numeric(rownames(SPD)),
               ties = 'ordered', rule = 2)$y
+
+ggplot() + geom_line(data = logip, aes(bce, prob_dens)) +
+  geom_line(aes(x = as.numeric(rownames(SPD)), y = SPD[,1]))
 
 # Summary results
 log_summary <- simulation_summary(SPD, simulation_results,
@@ -302,6 +305,7 @@ plotSimulationSummary(log_summary)
 logplts <- plot_mc(log_summary)
 
 save(logplts, file = "../external_data/shorespd/logplts.rda")
+save(simresults, file = "../external_data/shorespd/logi_simresults.rda")
 
 ##### Monte Carlo simulation - Uniform #####
 

### 2023-07-27 R/0xmcmc.R

@@ -1,0 +1,44 @@
+library(ADMUR)
+
+# Load pd, logistic model, and in and max age for RSPD
+load(file = here::here("analysis/data/derived_data/rcarbon_mcmc_data.RData"))
+
+# Rename
+minage_rcarbon <- minage
+maxage_rcarbon <- maxage
+pd_rcarbon <- PD
+log_rcarbon <- log
+
+# Load pd, 4-CPL model, and in and max age for SSPD
+load(file = here::here("analysis/data/derived_data/shore_mcmc_data.RData"))
+
+# Rename
+minage_shore <- minage
+maxage_shore <- maxage
+pd_shore <- pd
+cpl4_shore <- cpl_4
+
+chain_rcarbon <- mcmc(PDarray = pd_rcarbon,
+                      startPars = log_rcarbon$par,
+                      type = 'logistic',
+                      N = 30000,
+                      burn = 2000,
+                      thin = 5,
+                      jumps = 0.025)
+
+chain_shore <- mcmc(PDarray = pd_shore,
+                      startPars = cpl4_shore$par,
+                      type = 'CPL',
+                      N = 30000,
+                      burn = 2000,
+                      thin = 5,
+                      jumps = 0.025)
+
+save(chain_rcarbon, chain_shore,
+     file = here("analysis/data/derived_data/mcmc_chains.RData"))
+
+chain_rcarbon$acceptance.ratio
+chain_shore$acceptance.ratio
+
+chain_rcarbon$all.pars
+

### 2023-07-27 R/0xmean_tpq_ranges.R

@@ -1,12 +1,13 @@
-library(shoredate)
+# Find the average range of TPQ ranges for the geological displacement curves,
+# to be reported in the main text.
 
+library(shoredate)
 
 # Load centre points for the displacement curves
 centrepoints <- sf::st_read(
   system.file("extdata/isobase_centrepts.gpkg",
               package = "shoredate",
               mustWork = TRUE), quiet = TRUE)
-
 
 # Horten
 

### 2023-07-27 R/10_radiocarbon_cpl_admur.R

@@ -21,9 +21,9 @@ maxage <- 10555
 CalArray <- makeCalArray(intcal20, calrange = c(minage, maxage))
 PD <- phaseCalibrator(c14, CalArray, remove.external = TRUE)
 
-SPD <- as.data.frame( rowSums(PD) )
+SPD <- as.data.frame(rowSums(PD))
 # normalise
-SPD <- SPD/( sum(SPD) * CalArray$inc )
+SPD <- SPD/(sum(SPD) * CalArray$inc)
 
 CPL.1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction,
                   PDarray = PD, type = 'CPL', NP = 20)
@@ -149,3 +149,7 @@ load("../external_data/shorespd/uniplts.rda")
 
 ggsave(here::here("analysis/figures/mc.png"),
        units = "px", width = 4000, height = 1250*2)
+
+# Save pd, logistic model and min and max age for mcmc
+save(log, PD, minage, maxage,
+     file = here("analysis/data/derived_data/rcarbon_mcmc_data.RData"))

### 2023-07-28 analysis/paper/paper.Rmd
Discussion

@@
 If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
-The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes as opposed to taphonomic factors.
+The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes.
 
@@
 This is also poses a challenge for the interpretation of the SSPD.
+
+Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this early period, neither proxy is devoid of influence from population density or mobility patterns.
+Properly adjudicating this relationship will thus depend on independent data on mobility patterns.
 

### 2023-07-28 R/0x_radiocarbon_cpl.R

@@ -48,10 +48,8 @@ cpl3 <- convertPars(pars = cpl_3$par, years = minage:maxage, type = 'CPL')
 cpl3$model <- "3-CPL"
 cpl4 <- convertPars(pars = cpl_4$par, years = minage:maxage, type = 'CPL')
 cpl4$model <- "4-CPL"
-cpl5 <- convertPars(pars = cpl_5$par, years = minage:maxage, type = 'CPL')
-cpl5$model <- "5-CPL"
 
-rcarbonmodels <- rbind(expp, logip, unifp, cpl1, cpl2, cpl3, cpl4, cpl5)
+rcarbonmodels <- rbind(expp, logip, unifp, cpl1, cpl2, cpl3, cpl4)
 
 rcarbon_lik <- c("Exponential" = -exp$value,
                "Logistic" = -logi$value,

### 2023-07-31 analysis/paper/paper.Rmd
Supp2

@@
   \usepackage{placeins}
+  \usepackage[left]{lineno}
+  \linenumbers
 # always_allow_html: true
@@
 Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
-Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jorgensen2020].
+Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].
 
@@
 Thus, this geographical and temporal limit are also used for this study.
-The present paper is decidedly exploratory.
+
+The present study is decidedly exploratory.
 While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses.
@@
 This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*.
-For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase.
-The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, 3], found by estimating the likelihood using the final SPD.
+For any given model, this involves finding the scalar product for each shoreline date against the probability of the null-model, and in the case of the ^14^C-dates, each site-phase.
+The overall model likelihood is the product of these individual probability and is not, critically [@timpson2021, 3], found by using the final SPD.
 From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model.
@@
 Of the contrasted models, the BIC values indicate that the exponential model is the best candidate.
-The negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development.
+Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development.
 Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c.
@@
 Finally, a positive deviation occurs again at X--5000 BCE.
-While all of these deviations should be viewed with caution, the small magnitude and questions concerning the reliability of shoreline dating after 4000 BCE [cf.
-@roalkvam2023] means that the negative deviations that occur around 3000 BCE are disregarded here.
+While all of the above deviations should also be considered with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf.
+@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that are disregarded here.
 
@@
 <!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
+```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the standard models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
+knitr::include_graphics(here("analysis/figures/bic.png"))
+```
 
@@
 9500 BCE, the RSPD shows no evidence of human activity until c.
-8500 BCE.
+8600 BCE.
 What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years.
-These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022].
-While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic as opposed to taphonomic and environmental factors are central drivers behind the observed pattern.
+These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022].
+While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind the observed pattern.
 
-The considered models indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
+The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
 An immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
 
-A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short lived sites along the coast [e.g.
+A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g.
 @bjerck2008; @bjerck2017].
-This is commonly seen in relation to homogeneous lithic technology over large parts of Scandinavia in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline.
-High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017].
+This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g.
+@bang-andersen2012; @bjerck2009; @fuglestvedt2012].
+High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018].
 The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
@@
 
-Following this earliest phase, @berg-hansen2022 have argued that there is an increase in sites between c.
-8000--7000 BCE in the northern part of the study area of this paper.
-However, their analysis covers a limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
-Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
-
-In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a decline in the number of shoreline dated sites through this time-interval, with possible positive reversals of this trend just after 8000 BCE and around 7000 BCE.
+Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak starts around 8800 BCE and concludes around 8000 BCE.
 It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [e.g.
 @damlien2018b; @roalkvam2022].
-Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@gunther2018; @manninen2021].
+Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021].
 If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
-The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes.
+The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE.
+This logistic growth in the RSPD has a corresponding positive deviation in the SSPD, which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.
 
-The SSPD deviates negatively from all models from c.
-to c.
-BCE, and positively just before the 4th millennium BCE.
-The negative deviation could possibly been seen in relation to an increase in inland activity that occurs around the same time, which is argued to reach its peak around 6000 BCE.
-@boaz1999 related this development to a separation of inland and coastal populations in this period.
-The middle of the 6th millennium BCE marks the onset of the classic Nøstvet phase [@reitan2022], which has traditionally been seen as characterised by a reduction in residential mobility, and a specialisation towards fishing.
+@berg-hansen2022 argued that there is an increase in sites between c.
+8000--7000 BCE in the northern part of the study area of this paper.
+Their analysis covers a more limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
+Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint discrete time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
+However, in line with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and with a smaller magnitude around 7100--6900 BCE.
+
+Subsequently, the SSPD deviates negatively from the exponential decay from c.
+6800 to c.
+5800 BCE, a decrease that was also noted by @berghansen2022.
+This is followed by a positive deviation just before the 4th millennium BCE.
+The negative deviation could possibly been seen in relation to an increase in inland activity in southern Norway that occurs around the same time, which is argued to reach its peak around 6000 BCE.
+It is also worth noting that there appears to be a corresponding negative deviation from the logistic null model in the RSPD just before 6000 BCE.
+However, given that the model could not be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and so no analytical weight can be placed on the deviations in the RSPD, given the present data.
+
+@boaz1999 related the development in inland activity to a separation of inland and coastal populations in this period.
+This is followed by the onset of the classic Nøstvet phase in the middle of the 6th millennium BCE  [@reitan2022], which has traditionally been seen as characterised by a reduction in residential mobility, and a specialisation towards fishing.
 This also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
+The developments that might be reflected in the positive deviation in the SSPD.
+
 Given the rejected null-models, the stable development in RSPD over this period, and the limited sample of ^14^C-dates, it is, however, difficult to
 
-It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a negative deviation just before 6000 BCE.
-This could possibly be related to an increase in population numbers following, and a  However, given that the logistic model could not be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and so no analytical weight can be placed on the deviations in the RSPD given the present data.
-This is also poses a challenge for the interpretation of the SSPD.
-
 Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this early period, neither proxy is devoid of influence from population density or mobility patterns.
-Properly adjudicating this relationship will thus depend on independent data on mobility patterns.
+Furthermore, the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD could suggest that after However, this has to be left as very tenuous suggestions, as properly adjudicating this relationship will depend on independent data that tracks the developments of the other factors that hvae been suggested to be of relevance to the behaviour of the two proxies.
 
@@
 While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain.
-
 <!-- However, the positive deviation in the SSPD from the logistic model at c.
@@
 
-While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
+Thus, while it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
 Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
-@collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
+@berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
 @barton2014; @clark2017; @preston2018; @roalkvam2022].
@@
 @jorgensen2020; @ordonez2022].
-Aligning these different lines of evidence would also be dependent on a more direct modelling of the development in the SSPS, rather than simply basing the analysis on deviations from a rejected null model [cf.
+Aligning these different lines of evidence would also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
 @timpson2021].
@@
 # Conclusion
-The finding that the frequency of radiocarbon dates and frequency of shoreline dated sites, in general, follow inverse trajectories is a valuable insight.
+
+This paper has presented a first exploration of summing the results of using the probabilistic method for shoreline dating presented in @roalkvam2023a for assessing the development of the frequency of shoreline dated sites over time.
+The approach demonstrates that it
+
+The finding that the frequency of radiocarbon dates and frequency of shoreline dated sites, in general, follow trajectories is a valuable insight.
 Although a more precise understanding of this relationship

### 2023-07-31 analysis/supplementary/supplementary.Rmd
Supp2

@@
 
-```{r data, echo = FALSE}
+```{r precompiled, echo = FALSE}
 # For the compilation of this Rmarkdown document, the results were pre-compiled 
 # and are loaded here. Set eval = TRUE in code chunks below to re-run the code.
-load(file = here::here("analysis/data/derived_data/shore_pd_models.RData"))
+load(file = here::here("analysis/data/derived_data/supplementary.RData"))
 

### 2023-07-31 analysis/supplementary/supplementary2.Rmd
Supp2

---
title: "Exploring convergence issues beyond 4-CPL for SPD of shoreline dates"
author: 
- 'Isak Roalkvam'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup}
# Load required packages
library(ggplot2)
library(patchwork)
library(ADMUR)
library(DEoptimR)

# Set seed for reproducibility
set.seed(42)
```

```{r data}
# Load data holding the PD array of shoreline dates
load(file = here::here("analysis/data/derived_data/shore_models_pd.RData"))

# Create SPD
SPD <- as.data.frame(rowSums(pd))
SPD <- SPD/(sum(SPD) * 5)

# Plot PD array and SPD for inspection
plotPD(pd)
plotPD(SPD)
```

```{r precompiled, echo = FALSE}
# For the compilation of this Rmarkdown document, the results were pre-compiled 
# and are loaded here. Set eval = TRUE in code chunks below to re-run the code.
load(file = here::here("analysis/data/derived_data/sampsize.RData"))

```

```{r sampsize, eval = FALSE}

# Define sample sizes
sampsize <- c(6, 32, 60, 120, 240, 480)

results <- list()
temppds <- list()

# Fit 4-6 CPL to randomly drawn number of shoreline dates, and save the results
for(i in 1:length(sampsize)){
  tmppd <- pd[, sample(ncol(pd), sampsize[i])]
  temppds[[i]] <- tmppd

  cpl4 <- JDEoptim(lower = rep(0, 7),
                   upper = rep(1, 7),
                   fn = objectiveFunction,
                   PDarray = tmppd,
                   type = 'CPL',
                   NP = 140)

  cpl5 <- JDEoptim(lower = rep(0, 9),
                   upper = rep(1, 9),
                   fn = objectiveFunction,
                   PDarray = tmppd,
                   type='CPL',
                   NP = 180)

  cpl6 <- JDEoptim(lower = rep(0, 11),
                   upper = rep(1,11),
                   fn = objectiveFunction,
                   PDarray = tmppd,
                   type = 'CPL',
                   NP = 220)

  results[[i]] <- rbind(cpl4, cpl5, cpl6)
  save(results, temppds, sampsize,
       file = here::here("analysis/data/derived_data/sampsize.RData"))
}
```

```{r, fig.dim = c(8, 5)}
# Plot the results
for(i in 1:length(temppds)){
  tmpspd <- as.data.frame(rowSums(temppds[[i]]) )
  tmpspd <- tmpspd/( sum(tmpspd) * 5 )

  tmpmodels <- as.data.frame(results[[i]])

  tmpcpl4 <- convertPars(pars = tmpmodels$par[[1]],
                         years = minage:maxage, type = 'CPL')
  tmpcpl4$model <- "4-CPL"
  tmpcpl4$convergence <- tmpmodels$convergence[[1]]

  tmpcpl5 <- convertPars(pars = tmpmodels$par[[2]],
                         years = minage:maxage, type = 'CPL')
  tmpcpl5$model <- "5-CPL"
  tmpcpl5$convergence <- tmpmodels$convergence[[2]]

  tmpcpl6 <- convertPars(pars = tmpmodels$par[[3]],
                         years = minage:maxage, type = 'CPL')
  tmpcpl6$model <- "6-CPL"
  tmpcpl6$convergence <- tmpmodels$convergence[[3]]

  tmpcpls <- rbind(tmpcpl4, tmpcpl5, tmpcpl6)

  print(ggplot() +
    geom_bar(aes(x = as.numeric(rownames(tmpspd)), y =  tmpspd[,1]), 
             stat = "identity", col = "grey") +
    geom_line(aes(x = as.numeric(rownames(tmpspd)), y =  tmpspd[,1])) +
    geom_line(data = tmpcpls, aes(x = year, y = pdf, col = model),
              linewidth = 1.1) +
    labs(x = "calBP", y = "PD", title = paste("Sample size = ", sampsize[i])) +
    scale_colour_manual(values = c("4-CPL" = "black",
                                    "5-CPL" = "goldenrod2",
                                    "6-CPL" = "cornflowerblue")) +
    scale_x_reverse() +
    theme_bw() +
    theme(legend.position = "bottom",
          legend.title=element_blank()) +
  ggplot() +
    geom_point(data = tmpcpls,
               aes(x = model,
                   y = as.factor(convergence),
                   col = as.factor(convergence)),
               size = 3) +
    scale_colour_manual(values = c("1" = "red",
                                   "0" = "chartreuse3")) +
    labs(x = "Model", y = "Convergence") +
    theme_bw() +
    theme(legend.position = "none"))
}
```

### 2023-07-31 R/0x_shoreline_cpl.R

@@ -76,11 +76,13 @@ cplplt <- ggplot() +
   theme_bw() +
   theme(legend.title = element_blank())
 
-bicplt <- ggplot() +
-  geom_point(aes(sh_bic, names(sh_bic)), size = 3) +
-  labs(x = "BIC", y = "Model") +
-  theme_bw()
+bicplt_shore <- ggplot() +
+  geom_point(aes(sh_bic, names(sh_bic)), size = 2) +
+  labs(x = "BIC", y = "", subtitle = "Models fit to SSPD") +
+  theme_bw() +
+  theme(panel.border = element_rect(linewidth = 0.5))
 
+bicplt_shore + bicplt
 cplplt + bicplt
 
 # Save pd, 4-CPL model and min and max age for mcmc

### 2023-07-31 R/10_radiocarbon_cpl_admur.R

@@ -72,9 +72,15 @@ bics <- data.frame(model = c("1-CPL", "2-CPL", "3-CPL",
                    bic = c(BIC.1, BIC.2, BIC.3, BIC.4, BIC.exp,
                            BIC.log, BIC.uniform))
 
+bics <- data.frame(model = c("1-CPL", "2-CPL", "3-CPL",
+                             "4-CPL", "Exponential", "Logistic", "Uniform"),
+                   bic = c(BIC.1, BIC.2, BIC.3, BIC.4, BIC.exp,
+                           BIC.log, BIC.uniform))
+
 bicplt <- ggplot(data = bics, aes(x = bic, y = model)) + geom_point(size = 2) +
-  labs(x = "BIC", y = "Model") +
+  labs(x = "BIC", y = "", subtitle = "Models fit to RSPD") +
   theme_bw()
+
 
 CPL1$name <- "1-CPL"
 CPL2$name <- "2-CPL"

### 2023-08-01 analysis/paper/paper.Rmd
Discussion

@@
 # Introduction
-Variation in population patterns are regarded important drivers and consequences of cultural variation, and of critical importance to our understanding of past human societies [e.g.
+Variation in population patterns are regarded as important drivers and potential consequences of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g.
 @french2021; @shennan2000; @shennan2021].
@@
 What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021].
-In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
+In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
 We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c.
@@
 # Background
-Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g.
+Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g.
 @mörner1979; @steffen2011].
-The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use.
+The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at their time of use.
 As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites.
@@
 
-Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years.
-This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
-In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites.
+Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years, or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021].
+The point-based estimates do therefore not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
+In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites.
 The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)).
@@
 One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations.
-Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to constitute a 'site'.
+Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a 'site' [@palmisano2017].
 However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
-Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2020; @crema2022 for a review].
+Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2021; @crema2022 for reviews].
 
@@
 It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021].
-This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022].
+This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, and that they cannot be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022].
 While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.
@@
 The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021.
-This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model.
-These null-models are typically a uniform, exponential or logistic distribution.
+This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model.
+These null models are typically a uniform, exponential or logistic distribution.
 These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or maximum likelihood estimation (MLE).
@@
 While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine.
-Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].
+Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].
 
@@
 This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites done here, as few assumptions concerning this overall development could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.
 
@@
 Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded.
-This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016].
+This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to be older than around 9300 BCE in the region [@glorstad2016].
 Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019.
@@
 Radiocarbon dates were taken from @roalkvam2023 and @solheim.
-Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a total of ^14^C-dates.
+Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83].
 Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c.
 2500 BCE, the Mesolithic is the main focus of this analysis.
-Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, further reducing the sample of ^14^C-dates to.
+Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to the final sample of 310 ^14^C-dates.
 Probabilities falling after 4000 BCE and the first possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.
@@
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
-Analysis of the ^14^C-dates were done using the R package *ADMUR* [@timpson2021].
-The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* [@crema2021] for Monte Carlo simulation and model comparison.
+Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021].
+The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for Monte Carlo simulation and model comparison.
 
@@
 This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*.
-For any given model, this involves finding the scalar product for each shoreline date against the probability of the null-model, and in the case of the ^14^C-dates, each site-phase.
+For any given model, this involves finding the scalar product for each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase.
 The overall model likelihood is the product of these individual probability and is not, critically [@timpson2021, 3], found by using the final SPD.
@@
 For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected.
-As is also indicated by the BIC values, this model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development.
+As is also indicated by the BIC values, this model represents the best explanation of the data of the compared models, while the failure to reject the model means that the data could be expected under this logistic development.
 This is characterised by a rapid growth from the start of the RSPD which plateaus at around 7500 BCE and then remains stable for the remainder of the period.
@@
 Of the contrasted models, the BIC values indicate that the exponential model is the best candidate.
-Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development.
+Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development.
 Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c.
-X--9000 BCE, followed by a significant positive peak just after 9000 BCE.
-The process of decline is then interrupted by positive deviations at, and with a smaller deviation around 7000 BCE.
-This is closely followed by a negative deviation of more magnitude than that expected under the exponential model between X and X BCE.
-Finally, a positive deviation occurs again at X--5000 BCE.
-While all of the above deviations should also be considered with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf.
-@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that are disregarded here.
+9200--9000 BCE, followed by a significant positive peak just after 9000 BCE until around 8500 BCE.
+The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE.
+This is closely followed by a negative deviation between c.
+6800 and 5800 BCE.
+Finally, a positive deviation occurs again around 5300--5000 BCE.
+While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf.
+@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded here.
 
@@
 <!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
-```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the standard models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
+```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
 knitr::include_graphics(here("analysis/figures/bic.png"))
@@
 
-It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g.
+Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g.
+@åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g.
 @damlien2018b; @glorstad2016].
@@
 9500 BCE, the RSPD shows no evidence of human activity until c.
-8600 BCE.
+8500 BCE.
 What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years.
-These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022].
-While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind the observed pattern.
+These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022].
+A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of a more forested landscape, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf.
+@damm2022; @sørensen2014].
+While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.
 
 The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
-An immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic in the region [e.g.
+@lindblom1984; @fuglestvedt2018, 16--17], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
 
-A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g.
-@bjerck2008; @bjerck2017].
+A common interpretation of the Early Mesolithic in Norway is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g.
+@bjerck2017; @bjerck2008; @fuglestvedt2012].
 This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g.
@@
 The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
-The lacking signal in the RSPD for the first settlement period following the deglaciation could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf.
+The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf.
 @damm2022] and low population numbers.
 
-Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak starts around 8800 BCE and concludes around 8000 BCE.
-It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [e.g.
+Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak, starting from around 8800 BCE and concluding around 8000 BCE.
+It has previously been suggested that a change in mobility patterns occurs from some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g.
 @damlien2018b; @roalkvam2022].
@@
 If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
-The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE.
-This logistic growth in the RSPD has a corresponding positive deviation in the SSPD, which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.
+The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that has plateaued in the first centuries after 8000 BCE.
+The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.
 
-@berg-hansen2022 argued that there is an increase in sites between c.
+<!-- @berg-hansen2022 found an increase in sites between c.
 8000--7000 BCE in the northern part of the study area of this paper.
-Their analysis covers a more limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length.
+Their analysis covers a more limited temporal span and geographical area, and is based on the aggregation of point-based estimates of shoreline dates within periods of variable length.
 Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint discrete time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
-However, in line with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and with a smaller magnitude around 7100--6900 BCE.
+However, consistent with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and, although with a smaller magnitude, around 7100--6900 BCE.
+-->
 
-Subsequently, the SSPD deviates negatively from the exponential decay from c.
+Subsequently, the SSPD deviates negatively from the exponential null from c.
 6800 to c.
-5800 BCE, a decrease that was also noted by @berghansen2022.
-This is followed by a positive deviation just before the 4th millennium BCE.
-The negative deviation could possibly been seen in relation to an increase in inland activity in southern Norway that occurs around the same time, which is argued to reach its peak around 6000 BCE.
-It is also worth noting that there appears to be a corresponding negative deviation from the logistic null model in the RSPD just before 6000 BCE.
-However, given that the model could not be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and so no analytical weight can be placed on the deviations in the RSPD, given the present data.
+5800 BCE (a decrease also noted by @berg-hansen2022).
+This is followed by a positive deviation in the last centuries before 5000 BCE.
+The negative deviation could possibly been seen in relation to an increase in inland activity in southern Norway, which is to take place in the period from c.
+6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018].
+This period is followed by a marked reduction in inland activity, which thus corresponds to the positive deviation in the SSPD, until a new increase occurs around 4500--4000 BCE [@persson2018, 204].
+@boaz1999 related the first increase inland activity to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in coastal adaptation.
+The subsequent decrease in inland activity has been related to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022].
+The Nøstvet period is argued to be characterised by a specialised coastal adaptation that also coincide with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway, starting from around [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
+The onset of the classic Nøstvet phase thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.
 
-@boaz1999 related the development in inland activity to a separation of inland and coastal populations in this period.
-This is followed by the onset of the classic Nøstvet phase in the middle of the 6th millennium BCE  [@reitan2022], which has traditionally been seen as characterised by a reduction in residential mobility, and a specialisation towards fishing.
-This also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
-The developments that might be reflected in the positive deviation in the SSPD.
+Relating these patterns to demographic developments or land-use and mobility patterns is challenging given the data available here.
+Given the stationarity of the logistic model fit to the RSPD after c.
+7500 BCE, is it possible that fluctuatons in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments.
+There appears to be a negative deviation from the logistic null model in the RSPD which corresponds to the one in the SSPD just before 6000 BCE.
+However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, given the present data.
+While it is possible that the fairly limited sample of ^14^C-dates might lead to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
 
-Given the rejected null-models, the stable development in RSPD over this period, and the limited sample of ^14^C-dates, it is, however, difficult to
+Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns.
+This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE.
+However, properly adjudicating this relationship and the degree to which it has been the same throughout the period, will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
 
-Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this early period, neither proxy is devoid of influence from population density or mobility patterns.
-Furthermore, the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD could suggest that after However, this has to be left as very tenuous suggestions, as properly adjudicating this relationship will depend on independent data that tracks the developments of the other factors that hvae been suggested to be of relevance to the behaviour of the two proxies.
+Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations.
+The critical envelopes resulting from the simulations therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the trajectory of relative sea-level change and a relationship between sites and shoreline that can constantly be described by the employed gamma distribution in the dating procedure.
+Beyond uncertainties to do with the dating procedure itself, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models.
+An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that expected under a constant adjustment to the shoreline.
+If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
+Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the cultural historical implications of variation in this proxy.
 
-The logistic growth for the RSPD then indicates that a plateau is reached by the mid 7th millennium BCE and a stable population for the remainder of the Mesolithic.
+<!-- The logistic growth for the RSPD then indicates that a plateau is reached by the mid 7th millennium BCE and a stable population for the remainder of the Mesolithic.  -->
 
-While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain.
+<!-- While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain.
+-->
 <!-- However, the positive deviation in the SSPD from the logistic model at c.
@@
 This would also i -->
-
-Thus, while it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
-Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
-@berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
-@barton2014; @clark2017; @preston2018; @roalkvam2022].
-Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
-@jorgensen2020; @ordonez2022].
-Aligning these different lines of evidence would also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
-@timpson2021].
-Furthermore, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data, but could also provide important contextualisation for the developments observed here [see @lundstrom2021].
-
-Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application.
-While we still believe the approach represents an improvement over previous approaches, the results should therefore considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the frequency of coastal sites in the region.
 
@@
 
-This paper has presented a first exploration of summing the results of using the probabilistic method for shoreline dating presented in @roalkvam2023a for assessing the development of the frequency of shoreline dated sites over time.
-The approach demonstrates that it
+This paper has presented a new methodology for assembling and assessing the result of summing a large number of shoreline dates, drawing on the probabilistic method for shoreline dating presented in @roalkvam2023, and analogous procedures for the treatment of radiocarbon dates [@crema2022; @timpson2014; @crema2021].
+The result of this procedure was compared with the SPD of radiocarbon dates from within the same area to begin to assess the relationship between the two measures and their potential association with overall developments in population density and mobility patterns.
 
-The finding that the frequency of radiocarbon dates and frequency of shoreline dated sites, in general, follow trajectories is a valuable insight.
-Although a more precise understanding of this relationship
+While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
+Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
+@berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
+@barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change.
+In terms of mobility patterns, focus has been directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility could  also prove to be important.
+In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
+@jorgensen2020; @ordonez2022].
+Furthermore, aligning these different lines of evidence will also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
+@timpson2021].
+Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data, but could also provide important contextualisation for the developments observed here [see e.g.
+@lundstrom2021; @manninen2023].
+
+Seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application.
+While we believe the approach represents an improvement over previous approaches, the results should therefore considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.
 

### 2023-08-01 R/0xmean_tpq_ranges.R

@@ -97,3 +97,27 @@ tpqsd <- sd(c(prange, hrange, arange, trange), na.rm = TRUE)
 # save results
 save(tpqmean, tpqsd,
      file = here("analysis/data/derived_data/disp_tpq_ranges.RData"))
+
+library(dplyr)
+
+dispcurves <- get(load(system.file("extdata/displacement_curves.rda",
+              package = "shoredate")))
+
+rocdisp <- dispcurves %>%
+  group_by(name) %>%
+  arrange(bce, .by_group = TRUE) %>%
+  mutate(rate_upper = upperelev - lag(upperelev),
+         rate_lower = lowerelev - lag(lowerelev)) %>%
+  rowwise() %>%
+  mutate(roc_mean = mean(c(rate_upper, rate_lower), na.rm = TRUE))
+
+rocdisp <- rocdisp %>%
+  group_by(name) %>%
+  mutate( roc_mean = rowMeans(select(rocdisp,
+                  c(rate_upper, rate_lower)), na.rm = TRUE))
+
+ggplot(data = rocdisp) +
+  # geom_line(aes(x = bce, y = rate_upper, col = name)) +
+  # geom_line(aes(x = bce, y = rate_lower, col = name)) +
+  geom_line(aes(x = bce, y = roc_mean, col = name)) +
+  scale_x_continuous(breaks = seq(-10000, 0, 1000))

### 2023-08-03 analysis/paper/paper.Rmd
Extended, results, discussion and conclusion

@@
     bookdown::pdf_document2:
-      fig_caption: yes
+      fig_caption: true
       toc: false
@@
 @carleton2021; @torfing2015; @ward2021].
-What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021].
+What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012].
 In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
@@
 This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date.
-D) The shoreline dating approach used for this study, where the gamma distrbution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
+D) The shoreline dating approach used for this study, where the gamma distrbiution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
 
@@
 ```
-For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected.
+The results of the Monte Carlo simulations are given in Figure \@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values.
+The comparison of the models based on BIC scores is given in Figure \@ref(fig:bic).
+
+For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E).
 As is also indicated by the BIC values, this model represents the best explanation of the data of the compared models, while the failure to reject the model means that the data could be expected under this logistic development.
-This is characterised by a rapid growth from the start of the RSPD which plateaus at around 7500 BCE and then remains stable for the remainder of the period.
+This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.
 
@@
 Of the contrasted models, the BIC values indicate that the exponential model is the best candidate.
-Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development.
+Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A).
 Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c.
@@
 <!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
-```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
+```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs. Lower values indicate a better model.", out.width = "500px", fig.align="center", regfloat = TRUE}
 knitr::include_graphics(here("analysis/figures/bic.png"))
@@
 These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022].
-A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of a more forested landscape, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf.
+A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscape, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf.
 @damm2022; @sørensen2014].
@@
 The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
-Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic in the region [e.g.
-@lindblom1984; @fuglestvedt2018, 16--17], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic [e.g.
+@bang-andersen2005; @fuglestvedt2018, 16--17; @lindblom1984; @bjerck2008, 103--105], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
 
@@
 
-Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak, starting from around 8800 BCE and concluding around 8000 BCE.
+Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c.
+7900 BCE.
 It has previously been suggested that a change in mobility patterns occurs from some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g.
@@
 If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
-The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that has plateaued in the first centuries after 8000 BCE.
+The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE.
 The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.
@@
 6800 to c.
-5800 BCE (a decrease also noted by @berg-hansen2022).
+5800 BCE.
 This is followed by a positive deviation in the last centuries before 5000 BCE.
-The negative deviation could possibly been seen in relation to an increase in inland activity in southern Norway, which is to take place in the period from c.
+These developments should possibly been seen in relation to simultaneous developments in the inland areas of southern Norway.
+The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c.
 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018].
-This period is followed by a marked reduction in inland activity, which thus corresponds to the positive deviation in the SSPD, until a new increase occurs around 4500--4000 BCE [@persson2018, 204].
-@boaz1999 related the first increase inland activity to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in coastal adaptation.
-The subsequent decrease in inland activity has been related to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022].
-The Nøstvet period is argued to be characterised by a specialised coastal adaptation that also coincide with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway, starting from around [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
-The onset of the classic Nøstvet phase thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.
+@boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in coastal adaptation.
+This period is followed by a marked reduction in inland activity, lasting until c.
+4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c.
+5800 BCE and the positive deviation from c.
+5300--5000 BCE.
+@boaz1999 associated the period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022].
+The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincide with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
+The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.
 
-Relating these patterns to demographic developments or land-use and mobility patterns is challenging given the data available here.
-Given the stationarity of the logistic model fit to the RSPD after c.
-7500 BCE, is it possible that fluctuatons in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments.
-There appears to be a negative deviation from the logistic null model in the RSPD which corresponds to the one in the SSPD just before 6000 BCE.
-However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, given the present data.
-While it is possible that the fairly limited sample of ^14^C-dates might lead to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
+Directly relating these patterns to demographic developments or land-use and mobility patterns is challenging given the data available here.
+Given the stationarity of the logistic model of the RSPD after c.
+7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments.
+There appears to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE.
+However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data.
+While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
 
@@
 This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE.
-However, properly adjudicating this relationship and the degree to which it has been the same throughout the period, will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
+However, properly adjudicating this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
 
-Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations.
-The critical envelopes resulting from the simulations therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the trajectory of relative sea-level change and a relationship between sites and shoreline that can constantly be described by the employed gamma distribution in the dating procedure.
-Beyond uncertainties to do with the dating procedure itself, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models.
-An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that expected under a constant adjustment to the shoreline.
-If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
-Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the cultural historical implications of variation in this proxy.
+Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g.
+@crema2022].
+The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma distribution employed for the dating procedure (Figure \@ref(fig:shoredate)D).
+Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models.
+An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline.
+If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
+Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of variation in this proxy.
 
@@
 
-This paper has presented a new methodology for assembling and assessing the result of summing a large number of shoreline dates, drawing on the probabilistic method for shoreline dating presented in @roalkvam2023, and analogous procedures for the treatment of radiocarbon dates [@crema2022; @timpson2014; @crema2021].
-The result of this procedure was compared with the SPD of radiocarbon dates from within the same area to begin to assess the relationship between the two measures and their potential association with overall developments in population density and mobility patterns.
+This paper has presented a new methodology for assembling and assessing the sum of a large number of shoreline dates.
+This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021].
+The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types and their potential association with overall developments in population density and mobility patterns.
+
+The main results were that the logistic model.
+This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE.
+This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay.
+While this indicates that some process of decline in the number of shoreline dated has occurred over time, the model was rejected as an adequate description of the data.
+Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations, but given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions for future lines of interrogation.
 

### 2023-08-04 analysis/paper/paper.Rmd
Minor edits

@@
 @french2021; @shennan2000; @shennan2021].
-During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022].
+During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022].
 This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].
@@
 Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g.
-@carleton2021; @torfing2015; @ward2021].
-What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012].
+@attenbrow2015; @carleton2021; @torfing2015; @ward2021].
+What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012].
 In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
@@
 This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*.
-For any given model, this involves finding the scalar product for each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase.
-The overall model likelihood is the product of these individual probability and is not, critically [@timpson2021, 3], found by using the final SPD.
+For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase.
+The overall model likelihood is the product of these individual probabilities and is not, critically [@crema2022, @timpson2021], found by using the values of the final SPD.
 From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model.
@@
 While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf.
-@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded here.
+@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.
 
@@
 
-Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c.
+Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c.
 7900 BCE.
-It has previously been suggested that a change in mobility patterns occurs from some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g.
+It has previously been suggested that a change in mobility patterns occurs some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g.
 @damlien2018b; @roalkvam2022].
-Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021].
+Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021].
 If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
@@
 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018].
-@boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in coastal adaptation.
+@boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in the utilisation of coastal areas.
 This period is followed by a marked reduction in inland activity, lasting until c.
 4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c.
-5800 BCE and the positive deviation from c.
+5800 BCE, and the positive deviation from c.
 5300--5000 BCE.
 @boaz1999 associated the period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022].
-The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincide with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
+The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
 The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.
@@
 While this indicates that some process of decline in the number of shoreline dated has occurred over time, the model was rejected as an adequate description of the data.
-Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations, but given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions for future lines of interrogation.
+Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations, but given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions representing potential lines for further interrogation.
 

### 2023-08-08 analysis/paper/paper.Rmd
Added a few references

@@
 This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating.
-Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
+Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
 Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].
 
-Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years, or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021].
+Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g.
+@solheim2018; @fossum2020; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021].
 The point-based estimates do therefore not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
@@
 As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
-A common approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate.
-However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
+A common approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve [e.g.
+@breivik2014], and sometimes adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate.
+However, TPQ dates that do not account for the likely distance between site and sea when the site was in use, and limits the further inferential steps that can be taken.
 Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
@@
 When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
-These point-estimates are then aggregated in bins of 200 or 500 years [e.g.
-@breivik2014; @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020].
+As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years.
 This approach is illustrated in Figure \@ref(fig:sumshore)B.
@@
 For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase.
-The overall model likelihood is the product of these individual probabilities and is not, critically [@crema2022, @timpson2021], found by using the values of the final SPD.
+The overall model likelihood is the product of these individual probabilities and is not, critically [@crema2022; @timpson2021], found by using the values of the final SPD.
 From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model.
@@
 These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022].
-A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscape, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf.
+A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf.
 @damm2022; @sørensen2014].
@@
 
-The main results were that the logistic model.
+The main results was that the RSPD is consistent with the logistic model fit to the data.
 This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE.
 This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay.
-While this indicates that some process of decline in the number of shoreline dated has occurred over time, the model was rejected as an adequate description of the data.
-Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations, but given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions representing potential lines for further interrogation.
+While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD.
+Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations.
+Given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions representing potential lines for further interrogation.
 
 While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
-Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
+This is also relevant for the treatment of RSPDs, which in in Norwegian archaeology have usually been treated as constantly and uniformly impacted by population density through the period.
+We suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a determining contributor to the absence of radiocarbon dates from this earliest period.
+The degree to which mobility patterns might have also and variably impacted the radiocarbon record from later periods remains unclear, and could prove important for isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@hiscock2016, 219].
+
+Some lines of inquiry in the future could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
 @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
 @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change.
-In terms of mobility patterns, focus has been directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility could  also prove to be important.
+In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility could also prove to be important.
 In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
@@
 
-Seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application.
+Seeing as the implemented method for shoreline dating was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application.
 While we believe the approach represents an improvement over previous approaches, the results should therefore considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.

### 2023-08-08 analysis/supplementary/supplementary2.Rmd
Added a few references

@@
 ```
+
+The figure below displays the results of attempting to adjust parameter settings for improving the parameter search for the 5-CPL model.
+There only appears to minute variation in the 'value', and none of the models reach convergence.
+The parameters were altered one at a time, using the following code as the starting point:
+```{r, eval = FALSE}
+JDEoptim(lower = rep(0, 9),
+                   upper = rep(1, 9),
+                   fn = objectiveFunction,
+                   PDarray = pd,
+                   type='CPL',
+                   NP = 180)
+```

### 2023-08-08 R/0x_shoreline_monte_carlo2.R

@@ -202,7 +202,8 @@ simresults <- do.call(rbind.data.frame, results) %>%
   group_by(simn) %>%
   mutate(prob_sum_normalised = prob_sum /(sum(prob_sum, na.rm = TRUE) * 5))
 
-mod <- approx(x = expp$bce, y = expp$prob_dens, xout = as.numeric(rownames(SPD)),
+mod <- approx(x = (expp$bce * -1) + 1950, y = expp$prob_dens,
+              xout = as.numeric(rownames(SPD)),
               ties = 'ordered', rule = 2)$y
 
 # Model summary (can also be plotted with ADMUR)
@@ -291,9 +292,6 @@ mod <- approx(x = (logip$bce * -1) + 1950, y = logip$prob_dens,
               xout = as.numeric(rownames(SPD)),
               ties = 'ordered', rule = 2)$y
 
-ggplot() + geom_line(data = logip, aes(bce, prob_dens)) +
-  geom_line(aes(x = as.numeric(rownames(SPD)), y = SPD[,1]))
-
 # Summary results
 log_summary <- simulation_summary(SPD, simulation_results,
                                   c(-2500, -9445), mod, ncol(pd))
@@ -305,7 +303,7 @@ plotSimulationSummary(log_summary)
 logplts <- plot_mc(log_summary)
 
 save(logplts, file = "../external_data/shorespd/logplts.rda")
-save(simresults, file = "../external_data/shorespd/logi_simresults.rda")
+load(file = "../external_data/shorespd/logi_simresults.rda")
 
 ##### Monte Carlo simulation - Uniform #####
 

### 2023-08-10 analysis/paper/paper.Rmd
Minor edits

@@
 During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022].
-This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].
+This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2023; see also @persson2009].
 
@@
 Several limitations and forms of criticism have, however, been directed at these procedures.
-Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g.
+Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g.
 @attenbrow2015; @carleton2021; @torfing2015; @ward2021].
@@
 In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
-We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c.
+We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c.
 9500--4000 BCE.
@@
 Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
-Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].
+Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].
 
@@
 The present study is decidedly exploratory.
-While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses.
+While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses.
 Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables.
@@
 
-The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a statistically coherent procedure.
+The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a directly coherent statistical procedure.
 This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022].
@@
 These null models are typically a uniform, exponential or logistic distribution.
-These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or maximum likelihood estimation (MLE).
+These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments, maximum likelihood estimation (MLE).
 The result from these simulations are then used to create a 95% critical envelope representing the null model.
@@
 The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
-This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites done here, as few assumptions concerning this overall development could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.
+This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and, it is worth underscoring, that deviations from these rejected null models do not in themselves give statistical justification of the interpretation of any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves -- statistical or otherwise.
 
@@
 Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded.
-This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to be older than around 9300 BCE in the region [@glorstad2016].
+This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to hvae an older date than around 9300 BCE in the region [@glorstad2016].
 Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019.
@@
 2500 BCE, the Mesolithic is the main focus of this analysis.
-Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to the final sample of 310 ^14^C-dates.
-Probabilities falling after 4000 BCE and the first possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.
+Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates.
+Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.
 
@@
 As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
-A common approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve [e.g.
-@breivik2014], and sometimes adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate.
-However, TPQ dates that do not account for the likely distance between site and sea when the site was in use, and limits the further inferential steps that can be taken.
-Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
+One approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve [e.g.
+@breivik2014].
+However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
+Furthermore, the common practice of adding a constant error to a point-estimate, as mentioned above, does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
 
@@
 From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model.
-This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.
+This penalises the models for the number of parameters in use to avoid over-fitting, and provides a measure for the relative performance of the models.
 
@@
 For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E).
-As is also indicated by the BIC values, this model represents the best explanation of the data of the compared models, while the failure to reject the model means that the data could be expected under this logistic development.
+As is also indicated by the BIC values, of the compared models, the logistic model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development.
 This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.
@@
 Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c.
-9200--9000 BCE, followed by a significant positive peak just after 9000 BCE until around 8500 BCE.
+9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE.
 The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE.
@@
 
-The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material.
+The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the Mesolithic, contrasted by a general stationary frequency of ^14^C-dated material.
 Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic [e.g.
@@
 @damlien2018b; @roalkvam2022].
-Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021].
+Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021].
 If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns.
@@
 This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE.
-However, properly adjudicating this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
+However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
 
@@
 While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD.
-Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations.
-Given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions representing potential lines for further interrogation.
+Some of the deviations from the model were then narratively explored to suggest some possible reasons for the reasons and nature of these deviations.
+Given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.
 
@@
 We suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a determining contributor to the absence of radiocarbon dates from this earliest period.
-The degree to which mobility patterns might have also and variably impacted the radiocarbon record from later periods remains unclear, and could prove important for isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@hiscock2016, 219].
+The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods remains unclear, and could prove important for isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@hiscock2016, 219].
 
-Some lines of inquiry in the future could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
+Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g.
 @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
@@
 In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
-@jorgensen2020; @ordonez2022].
-Furthermore, aligning these different lines of evidence will also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
+@jorgensen2020; @ordonez2022; @lundstrom2023].
+Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
 @timpson2021].
-Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data, but could also provide important contextualisation for the developments observed here [see e.g.
-@lundstrom2021; @manninen2023].
+Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g.
+@lundstrom2021; @manninen2023; @solheim2018].
 
 Seeing as the implemented method for shoreline dating was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application.
-While we believe the approach represents an improvement over previous approaches, the results should therefore considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.
+While we believe the approach represents an improvement over previous approaches, the results should therefore be considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.
 

### 2023-08-11 analysis/paper/paper.Rmd
Minor edits

@@
 One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations.
-Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a 'site' [@palmisano2017].
+Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a 'site' [@crema2020; @palmisano2017].
 However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.

### 2023-08-15 analysis/paper/paper.Rmd
Setting up supp 2

--- 2023-08-10 analysis/paper/paper.Rmd
@@
 # Introduction
-Variation in population patterns are regarded as important drivers and potential consequences of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g.
+Variation in population patterns are regarded as important drivers and potential outcomes of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g.
 @french2021; @shennan2000; @shennan2021].
@@
 One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations.
-Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a 'site' [@palmisano2017].
+Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a site [@drennan2015, 16--20; @palmisano2017].
 However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
@@
 This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, and that they cannot be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022].
-While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.
+While this problem cannot be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.
 
@@
 This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and, it is worth underscoring, that deviations from these rejected null models do not in themselves give statistical justification of the interpretation of any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves -- statistical or otherwise.
+However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and, it is worth underscoring, deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.
 
@@
 The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample.
-Radiocarbon dates were taken from @roalkvam2023 and @solheim.
+Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden.
 Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83].
@@
 As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
-One approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve [e.g.
-@breivik2014].
 However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
-Furthermore, the common practice of adding a constant error to a point-estimate, as mentioned above, does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
+Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
 
@@
 C) The black line on the y-axis represents the elevation of the site.
-This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date.
+This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea and represents a \\textit{terminus post quem} date for the site.
 D) The shoreline dating approach used for this study, where the gamma distrbiution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
@@
 This approach is illustrated in Figure \@ref(fig:sumshore)B.
-The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c.
+The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c.
 8000 BCE.
@@
 8500 BCE.
-What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years.
+What has caused the lack of ^14^C-dates in south-eastern Norway from this earliest period has been given many possible explanations through the years.
 These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022].
@@
 7900 BCE.
-It has previously been suggested that a change in mobility patterns occurs some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g.
+It has previously been suggested that a change in mobility patterns occurs some time around 8500--8000 BCE, involving a reduction in residential mobility [e.g.
 @damlien2018b; @roalkvam2022].
@@
 While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
+A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf.
+@timpson2021; @tallavaara2021], and which impacts the comparison of the two proxies following both from the resolution of the dating methods and from the different sample sizes underlying the SPDs.
 

### 2023-08-15 analysis/supplementary/supplementary2.Rmd
Setting up supp 2

--- 2023-07-31 analysis/supplementary/supplementary2.Rmd
@@
 # Load data holding the PD array of shoreline dates
-load(file = here::here("analysis/data/derived_data/shore_models_pd.RData"))
+load(file = here::here("analysis/data/derived_data/shore_pd.RData"))
 
@@
 ```{r precompiled, echo = FALSE}
-# For the compilation of this Rmarkdown document, the results were pre-compiled 
-# and are loaded here. Set eval = TRUE in code chunks below to re-run the code.
+# For the compilation of this Rmarkdown document, the results of the code chunk 
+# below  were pre-compiled and are loaded here. Set eval = TRUE below to re-run
+# the code.
 load(file = here::here("analysis/data/derived_data/sampsize.RData"))
@@
   results[[i]] <- rbind(cpl4, cpl5, cpl6)
-  save(results, temppds, sampsize,
-       file = here::here("analysis/data/derived_data/sampsize.RData"))
 }
@@
 ```
+
+# Different parameter settings for 5-CPL
+
+The figure below displays the results of attempting to adjust parameter settings for improving the parameter search for the 5-CPL model.
+The identified model parameters are identical across the runs, the variation in the returned values is minute, and none of the searches reached convergence.
+The parameters used with `JDEoptim()` were altered one at a time, resulting in 11 separate searches.
+The following code was used as the starting point from where parameters were adjusted:
+```{r, eval = FALSE}
+# Base code from where the parameters were adjusted individually
+JDEoptim(lower = rep(0, 9),
+                   upper = rep(1, 9),
+                   fn = objectiveFunction,
+                   PDarray = pd,
+                   type = 'CPL',
+                   NP = 180)
+```
+
+```{r, echo = FALSE}
+
+load(file = here::here("analysis/data/derived_data/exploring_cpl5.RData"))
+
+cpl5iter <- convertPars(pars = cpl_5$par, years = minage:maxage, type = 'CPL')
+cpl5iter$parameter <- "maxiter = 3600"
+cpl5iter$value <- cpl_5$value
+cpl5iter$convergence <- cpl_5$convergence
+
+cpl5iter2 <- convertPars(pars = cpl_5iter2$par, years = minage:maxage, type = 'CPL')
+cpl5iter2$parameter <- "maxiter = 7200"
+cpl5iter2$value <- cpl_5iter2$value
+cpl5iter2$convergence <- cpl_5iter2$convergence
+
+cpl5np <- convertPars(pars = cpl_5np$par, years = minage:maxage, type = 'CPL')
+cpl5np$parameter <- "NP = 360"
+cpl5np$value <- cpl_5np$value
+cpl5np$convergence <- cpl_5np$convergence
+
+cpl5np2 <- convertPars(pars = cpl_5np2$par, years = minage:maxage, type = 'CPL')
+cpl5np2$parameter <- "NP = 720"
+cpl5np2$value <- cpl_5np2$value
+cpl5np2$convergence <- cpl_5np2$convergence
+
+cpl5jitter <- convertPars(pars = cpl_5jitter$par, years = minage:maxage, type = 'CPL')
+cpl5jitter$parameter <- "jitter_factor = NULL"
+cpl5jitter$value <- cpl_5jitter$value
+cpl5jitter$convergence <- cpl_5jitter$convergence
+
+cpl5jitter2 <- convertPars(pars = cpl_5jitter2$par, years = minage:maxage, type = 'CPL')
+cpl5jitter2$parameter <- "jitter_factor = 0.01"
+cpl5jitter2$value <- cpl_5jitter2$value
+cpl5jitter2$convergence <- cpl_5jitter2$convergence
+
+cpl5jitter3 <- convertPars(pars = cpl_5jitter3$par, years = minage:maxage, type = 'CPL')
+cpl5jitter3$parameter <- "jitter_factor = 0.1"
+cpl5jitter3$value <- cpl_5jitter3$value
+cpl5jitter3$convergence <- cpl_5jitter3$convergence
+
+cpl5jitter4 <- convertPars(pars = cpl_5jitter4$par, years = minage:maxage, type = 'CPL')
+cpl5jitter4$parameter <- "jitter_factor = 0.0001"
+cpl5jitter4$value <- cpl_5jitter4$value
+cpl5jitter4$convergence <- cpl_5jitter4$convergence
+
+cpl5mutprob1 <- convertPars(pars = cpl_5mutprob1$par, years = minage:maxage, type = 'CPL')
+cpl5mutprob1$parameter <- "tau_pF = 0.5"
+cpl5mutprob1$value <- cpl_5mutprob1$value
+cpl5mutprob1$convergence <- cpl_5mutprob1$convergence
+
+cpl5mutprob2 <- convertPars(pars = cpl_5mutprob2$par, years = minage:maxage, type = 'CPL')
+cpl5mutprob2$parameter <- "tau_pF = 0.2"
+cpl5mutprob2$value <- cpl_5mutprob2$value
+cpl5mutprob2$convergence <- cpl_5mutprob2$convergence
+
+cpl5mutprob3 <- convertPars(pars = cpl_5mutprob3$par, years = minage:maxage, type = 'CPL')
+cpl5mutprob3$parameter <- "tau_pF = 1"
+cpl5mutprob3$value <- cpl_5mutprob3$value
+cpl5mutprob3$convergence <- cpl_5mutprob3$convergence
+
+cpl5models <- rbind(cpl5iter, cpl5iter2, cpl5np, cpl5np2, cpl5jitter,
+                    cpl5jitter2, cpl5jitter3, cpl5jitter4, cpl5mutprob1,
+                    cpl5mutprob2, cpl5mutprob3)
+
+plt1 <- ggplot() + geom_line(aes(x = as.numeric(rownames(SPD)), y = SPD[,1])) + geom_line(data = cpl5models, aes(x = year, y = pdf, col = parameter)) +
+  labs(title = "Different parameter settings for 5-CPL search") + scale_x_reverse() +
+  theme_bw()
+plt2 <- ggplot() + geom_point(data = cpl5models, aes(x = value, y = parameter)) + geom_vline(xintercept = cpl_5$value, col = "red") +
+  ylab("") +
+  theme_bw()
+plt3 <- ggplot() + geom_point(data = cpl5models, aes(x = as.factor(convergence), y = parameter, col = as.factor(convergence))) +
+  ylab("") +
+  theme_bw() + theme(legend.position =  "none", axis.text.y=element_blank(), axis.ticks.y=element_blank())
+
+plt1
+
+```
+
+```{r, echo = FALSE}
+plt2 + plt3
+```

### 2023-08-15 R/0x_shoreline_cpl.R

@@ -13,22 +13,17 @@ load(file = here::here("analysis/data/derived_data/shore_models_pd.RData"))
 cpl_1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction, PDarray = pd,
                   type = 'CPL', NP = 20, trace = TRUE)
 cpl_2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 60,  maxiter = 400 * 3,
-                  trace = TRUE)
+                  PDarray = pd, type = 'CPL', NP = 60, trace = TRUE)
 cpl_3 <- JDEoptim(lower = rep(0, 5), upper = rep(1, 5), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 100, maxiter = 400 * 10,
-                  trace = TRUE)
+                  PDarray = pd, type = 'CPL', NP = 100, trace = TRUE)
 cpl_4 <- JDEoptim(lower = rep(0, 7), upper = rep(1, 7), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 140, maxiter = 400 * 10,
-                  trace = TRUE)
+                  PDarray = pd, type = 'CPL', NP = 140, trace = TRUE)
 cpl_5 <- JDEoptim(lower = rep(0, 9), upper = rep(1, 9), fn = objectiveFunction,
-                  PDarray = pd, type='CPL', NP = 180, maxiter = 400 * 10,
-                  trace = TRUE)
+                  PDarray = pd, type='CPL', NP = 180, trace = TRUE)
 cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 220, maxiter =  400 * 11,
-                  trace = TRUE)
+                  PDarray = pd, type = 'CPL', NP = 220, trace = TRUE)
 
-save(cpl_1, cpl_2, cpl_3, cpl_4,
+save(cpl_1, cpl_2, cpl_3, cpl_4, cpl_5, cpl_6,
      file = here::here("analysis/data/derived_data/shore_models.RData"))
[…]

### 2023-08-19 analysis/paper/paper.Rmd
Minor edits

@@
 @attenbrow2015; @carleton2021; @torfing2015; @ward2021].
-What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012].
+What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012].
 In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
@@
 @mörner1979; @steffen2011].
-The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at their time of use.
-As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites.
+The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or located close to the contemporaneous shoreline at their time of use.
+Thus, as coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites.
 This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating.
 Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.
-Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].
+A series of studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].
 
-Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g.
+Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g.
 @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021].
-The point-based estimates do therefore not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
+However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
 In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites.
-The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)).
+The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)).
 The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility.
@@
 Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records.
-The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020].
+The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented by the geometries available in the database [see also @roalkvam2020].
 All sites with a quality score of 4 or worse were excluded from further analysis.
 Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded.
-This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to hvae an older date than around 9300 BCE in the region [@glorstad2016].
+This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to have an older date than around 9300 BCE in the region [@glorstad2016].
 Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019.
@@
 All analyses done in this study were performed using the R programming language [@rcoreteam].
-Underlying data and programming code used for the paper is available in a version-controlled online repository at X.
+Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5.
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
@@
 The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
-The degree to which the observed RSPD deviates from the critical envelope is used to calculate a global p-value, indicating whether the null model can be rejected.
+The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model can be rejected.
 
@@
 
-```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.
+```{r shoredate, echo = FALSE, fig.cap = "A) The location of an example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.
 B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site.
@@
 This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea and represents a \\textit{terminus post quem} date for the site.
-D) The shoreline dating approach used for this study, where the gamma distrbiution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
+D) The shoreline dating approach used for this study, where the gamma function on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
 
@@
 
-When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
+When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
 As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years.
@@
 For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E).
-As is also indicated by the BIC values, of the compared models, the logistic model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development.
+As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development.
 This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.
@@
 4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c.
-5800 BCE, and the positive deviation from c.
+5800 BCE and the positive deviation from c.
 5300--5000 BCE.
-@boaz1999 associated the period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022].
+@boaz1999 associated this period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022].
 The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
@@
 
-Directly relating these patterns to demographic developments or land-use and mobility patterns is challenging given the data available here.
-Given the stationarity of the logistic model of the RSPD after c.
+A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c.
+6000 BCE.
+However, after this the population is found to have been characterised by a cessation of growth, rather than the clear decrease identified by the regional studies in southern Norway.
+Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [cf.
+@crema2017].
+
+Assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the two employed proxies is challenging given the divergence between them without invoking additional variables that can .
+Following from the stationarity of the logistic model of the RSPD after c.
 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments.
-There appears to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE.
+There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE.
 However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data.
@@
 A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf.
-@timpson2021; @tallavaara2021], and which impacts the comparison of the two proxies following both from the resolution of the dating methods and from the different sample sizes underlying the SPDs.
+@tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.
 
 Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns.
-This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE.
+This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE.
 However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
@@
 @crema2022].
-The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma distribution employed for the dating procedure (Figure \@ref(fig:shoredate)D).
+The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D).
 Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models.
@@
 If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
-Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of variation in this proxy.
+Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.
 
@@
 While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD.
-Some of the deviations from the model were then narratively explored to suggest some possible reasons for the reasons and nature of these deviations.
-Given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.
+Some of the deviations from the model were then narratively explored to suggest some possible reasons for these deviations.
+However, given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.
 
 While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
-This is also relevant for the treatment of RSPDs, which in in Norwegian archaeology have usually been treated as constantly and uniformly impacted by population density through the period.
-We suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a determining contributor to the absence of radiocarbon dates from this earliest period.
-The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods remains unclear, and could prove important for isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@hiscock2016, 219].
+This is also relevant for the treatment of RSPDs in the region, as we suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a contributor to the absence of radiocarbon dates from this earliest period.
+The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, potentially as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].
 
@@
 @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change.
-In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility could also prove to be important.
+In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important.
 In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
 @jorgensen2020; @ordonez2022; @lundstrom2023].
-Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
+Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
 @timpson2021].

### 2023-08-19 analysis/supplementary/supplementary2.Rmd
Minor edits

@@
 ```{r data}
-# Load data holding the PD array of shoreline dates
-load(file = here::here("analysis/data/derived_data/shore_pd.RData"))
+# Load data holding the PD array of shoreline dates and min and max age
+load(file = here::here("analysis/data/derived_data/shore_pd.Rdata"))
 
 # Create SPD
-SPD <- as.data.frame(rowSums(pd))
-SPD <- SPD/(sum(SPD) * 5)
+spd <- as.data.frame(rowSums(pd))
+spd <- spd/(sum(spd) * 5)
 
@@
 plotPD(pd)
-plotPD(SPD)
+plotPD(spd)
 ```
@@
 ```{r precompiled, echo = FALSE}
-# For the compilation of this Rmarkdown document, the results of the code chunk 
-# below  were pre-compiled and are loaded here. Set eval = TRUE below to re-run
-# the code.
+# For the compilation of this Rmarkdown document, the results of the two code 
+# chunks below were pre-compiled and are loaded here. Set eval = TRUE below to 
+# re-run the code.
 load(file = here::here("analysis/data/derived_data/sampsize.RData"))
+load(file = here::here("analysis/data/derived_data/shore_models1.RData"))
 
 ```
+
+## Default approach (5-CPL and 6-CPL do not convergence)
+
+```{r, eval = FALSE}
+cpl_1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction, PDarray = pd,
+                  type = 'CPL', NP = 20)
+cpl_2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 60)
+cpl_3 <- JDEoptim(lower = rep(0, 5), upper = rep(1, 5), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 100)
+cpl_4 <- JDEoptim(lower = rep(0, 7), upper = rep(1, 7), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 140)
+cpl_5 <- JDEoptim(lower = rep(0, 9), upper = rep(1, 9), fn = objectiveFunction,
+                  PDarray = pd, type='CPL', NP = 180)
+cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
+                  PDarray = pd, type = 'CPL', NP = 220)
+```
+
+```{r}
+cpl1 <- convertPars(pars = cpl_1$par, years = minage:maxage, type = 'CPL')
+cpl1$model <- "1-CPL"
+cpl2 <- convertPars(pars = cpl_2$par, years = minage:maxage, type = 'CPL')
+cpl2$model <- "2-CPL"
+cpl3 <- convertPars(pars = cpl_3$par, years = minage:maxage, type = 'CPL')
+cpl3$model <- "3-CPL"
+cpl4 <- convertPars(pars = cpl_4$par, years = minage:maxage, type = 'CPL')
+cpl4$model <- "4-CPL"
+cpl5 <- convertPars(pars = cpl_5$par, years = minage:maxage, type = 'CPL')
+cpl5$model <- "5-CPL"
+cpl6 <- convertPars(pars = cpl_6$par, years = minage:maxage, type = 'CPL')
+cpl6$model <- "6-CPL"
+
+shmodels <- rbind(cpl1, cpl2, cpl3, cpl4, cpl5, cpl6)
+
+sh_lik <- c("1-CPL" = -cpl_1$value,
+            "2-CPL" = -cpl_2$value,
+            "3-CPL" = -cpl_3$value,
+            "4-CPL" = -cpl_4$value,
+            "5-CPL" = -cpl_5$value,
+            "6-CPL" = -cpl_6$value)
+
+ssize <- ncol(pd)
+sh_bic <- c(log(ssize)*1 - 2*sh_lik[1], # 1-CPL
+            log(ssize)*3 - 2*sh_lik[2], # 2-CPL
+            log(ssize)*5 - 2*sh_lik[3], # 3-CPL
+            log(ssize)*7 - 2*sh_lik[4], # 4-CPL
+            log(ssize)*9 - 2*sh_lik[5], # 5-CPL
+            log(ssize)*11 - 2*sh_lik[6]) # 6-CPL
+
+```
+
+```{r}
+modelplt <- ggplot() +
+  geom_bar(aes(x = as.numeric(rownames(spd)),
+               spd[,1]),
+           stat = "identity", col = "grey") +
+  geom_line(aes(x = as.numeric(rownames(spd)),
+                y = spd[,1])) +
+  geom_line(data = shmodels, aes(year, pdf, col = model),
+            linewidth = 1.1) +
+  labs(x = "calBP", y = "PD") +
+  scale_colour_manual(values = c(
+    "1-CPL" = "#e41a1c",
+    "2-CPL" = "#4daf4a",
+    "3-CPL" = "#e0d616",
+    "4-CPL" = "#377eb8",
+    "5-CPL" = "black",
+    "6-CPL" = "#984ea3"
+  )) +
+  scale_x_reverse() +
+  theme_bw() +
+  theme(legend.title = element_blank())
+
+bicplt <- ggplot() +
+  geom_point(aes(sh_bic, names(sh_bic)), size = 2) +
+  labs(x = "BIC", y = "") +
+  scale_y_discrete(limits = rev) +
+  theme_bw()
+
+modelplt
+bicplt
+```
+
+## Increasing number of randomly sampled dates
 
@@
 
-# Fit 4-6 CPL to randomly drawn number of shoreline dates, and save the results
+# Fit 4-6 CPL to randomly drawn number of shoreline dates using the default
+# parameters
 for(i in 1:length(sampsize)){
@@
               linewidth = 1.1) +
-    labs(x = "calBP", y = "PD", title = paste("Sample size = ", sampsize[i])) +
+    labs(x = "calBP", y = "PD", title = paste("Randomly sampled dates = ",
+                                              sampsize[i])) +
     scale_colour_manual(values = c("4-CPL" = "black",
@@
 
-# Different parameter settings for 5-CPL
+## Exploring parameter settings for 5-CPL search
 
-The figure below displays the results of attempting to adjust parameter settings for improving the parameter search for the 5-CPL model.
+The figures below displays the results of attempting to adjust parameter settings for improving the parameter search for the 5-CPL model.
 The identified model parameters are identical across the runs, the variation in the returned values is minute, and none of the searches reached convergence.
 The parameters used with `JDEoptim()` were altered one at a time, resulting in 11 separate searches.
-The following code was used as the starting point from where parameters were adjusted:
+I have not included the code for each individual run and the plotting of the results, but the following code was used as the starting point from where parameters were adjusted:
 ```{r, eval = FALSE}
@@
 
-plt1 <- ggplot() + geom_line(aes(x = as.numeric(rownames(SPD)), y = SPD[,1])) + geom_line(data = cpl5models, aes(x = year, y = pdf, col = parameter)) +
-  labs(title = "Different parameter settings for 5-CPL search") + scale_x_reverse() +
+plt1 <- ggplot() + geom_line(aes(x = as.numeric(rownames(spd)), y = spd[,1])) + geom_line(data = cpl5models, aes(x = year, y = pdf, col = parameter)) +
+  scale_x_reverse() +
   theme_bw()

### 2023-08-19 R/0x_shoreline_cpl.R

@@ -25,7 +25,7 @@ cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
 
 save(cpl_1, cpl_2, cpl_3, cpl_4, cpl_5, cpl_6,
      file = here::here("analysis/data/derived_data/shore_models.RData"))
-load(file = here::here("analysis/data/derived_data/shore_models.RData"))
+load(file = here::here("analysis/data/derived_data/shore_models1.RData"))
 
 cpl1 <- convertPars(pars = cpl_1$par, years = minage:maxage, type = 'CPL')
 cpl1$model <- "1-CPL"
@@ -35,6 +35,10 @@ cpl3 <- convertPars(pars = cpl_3$par, years = minage:maxage, type = 'CPL')
 cpl3$model <- "3-CPL"
 cpl4 <- convertPars(pars = cpl_4$par, years = minage:maxage, type = 'CPL')
 cpl4$model <- "4-CPL"
+cpl5 <- convertPars(pars = cpl_5$par, years = minage:maxage, type = 'CPL')
+cpl5$model <- "5-CPL"
+cpl6 <- convertPars(pars = cpl_6$par, years = minage:maxage, type = 'CPL')
+cpl6$model <- "6-CPL"
 
 shmodels <- rbind(expp, logip, unifp, cpl1, cpl2, cpl3, cpl4, cpl5, cpl6)
 

### 2023-08-22 analysis/paper/paper.Rmd
Minor changes before sending off to David

@@
 
-The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach.
+The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach.
 Several limitations and forms of criticism have, however, been directed at these procedures.
@@
 @attenbrow2015; @carleton2021; @torfing2015; @ward2021].
-What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012].
+What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012; @rick1987].
 In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
@@
 Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables.
-The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf.
-@freeman2018].
+The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics.
 The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density.
@@
 However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
-Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2021; @crema2022 for reviews].
+Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @crema2022 for a recent review].
 
@@
 These null models are typically a uniform, exponential or logistic distribution.
-These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments, maximum likelihood estimation (MLE).
+These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE).
 The result from these simulations are then used to create a 95% critical envelope representing the null model.
@@
 While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine.
-Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].
+Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].
 
@@
 This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and, it is worth underscoring, deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.
+However, it is worth underscoring that deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.
 
@@
 Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded.
-This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to have an older date than around 9300 BCE in the region [@glorstad2016].
+This marks the latest start-date among the employed displacement curves, and no sites in the region have yet been conclusively verified to have an older date than around 9300 BCE [@glorstad2016].
 Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019.
@@
 Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021].
-The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for Monte Carlo simulation and model comparison.
+The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], and re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.
 
@@
 Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf.
-@weninger2015, 545], each shoreline date is normalised to sum to unity.
+@weninger2015, 545], each shoreline date is normalised to sum to unity before being summed.
 To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.
@@
 8500 BCE.
-What has caused the lack of ^14^C-dates in south-eastern Norway from this earliest period has been given many possible explanations through the years.
+What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years.
 These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022].
@@
 The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE.
-The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.
+The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.
 
@@
 
-A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c.
+A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies, found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c.
 6000 BCE.
-However, after this the population is found to have been characterised by a cessation of growth, rather than the clear decrease identified by the regional studies in southern Norway.
+However, after this the population dynamics indicated by the radiocarbon record is found to have been characterised by a cessation of growth, rather than the clear decrease suggested by the regional studies in southern Norway.
 Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [cf.
@@
 
-Assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the two employed proxies is challenging given the divergence between them without invoking additional variables that can .
+Following from the divergence between the proxies, assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the measures will remain speculative without drawing on additional variables that can control for such confounding effects.
 Following from the stationarity of the logistic model of the RSPD after c.
@@
 This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE.
-However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
+However, properly assessing this relationship and the degree to which it has held constant throughout the period will, as mentioned, depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
 
@@
 This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021].
-The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types and their potential association with overall developments in population density and mobility patterns.
+The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types.
+<!-- and their potential association with overall developments in population density and mobility patterns. -->
 
 The main results was that the RSPD is consistent with the logistic model fit to the data.
-This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE.
+This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from the first centuries after 8000 BCE.
 This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay.

### 2023-09-06 analysis/paper/paper.Rmd
Minor edits following comments from David

@@
 # Introduction
-Variation in population patterns are regarded as important drivers and potential outcomes of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g.
+Variations in population patterns are regarded as important drivers and potential outcomes of cultural variation, and are thus of critical importance to our understanding of past human societies [e.g.
 @french2021; @shennan2000; @shennan2021].
-During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022].
+During the last decade, the frequency distribution of radiocarbon dates has become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022].
 This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2023; see also @persson2009].
@@
 What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012; @rick1987].
-In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
+In this paper, we report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
 We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c.
@@
 Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g.
-@solheim2018; @fossum2020; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021].
+@fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021].
 However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
@@
 Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a site [@drennan2015, 16--20; @palmisano2017].
-However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria.
+However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently by factors such as those mentioned above.
 Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @crema2022 for a recent review].
 
-The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a directly coherent statistical procedure.
+The most critical of these follows from the fact that the summation of the probabilities associated with the dates for SPDs is not a directly coherent statistical procedure.
 This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022].
@@
 These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE).
-The result from these simulations are then used to create a 95% critical envelope representing the null model.
+The result from these simulations is then used to create a 95% critical envelope representing the null model.
 The proportion of the observed RSPD that falls outside the simulated envelope (i.e.
@@
 First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to.
-Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used.
+Secondly, for example, an exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used.
 While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine.
-Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].
+Finally, the p-value only indicates whether the null model as a whole can be rejected as an explanation of the data, and it does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].
 
 The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
-This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made.
+This can be helpful for getting an impression of the overall developments of population proxies, and it is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made.
 However, it is worth underscoring that deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.
@@
 The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD.
-Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006].
+Following @timpson2021, the model parameters were identified by MLE search with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006].
 The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*.
@@
 Here, this procedure was repeated 10000 times for each null model.
-The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
-The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model can be rejected.
+The 2.5 and 97.5 quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
+The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.
 
@@
 Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B).
-These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987].
+These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987; and map in Figure \@ref(fig:shoredate)A].
 The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated.
@@
 However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
-Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
+Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
 
@@
 
-When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
+When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
 As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years.
@@
 
-The Monte Carlo simulations with the SSPD is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice.
+The Monte Carlo simulations with the SSPD is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline-dated sites, where the probability of drawing any 5-year interval is determined by the null model of choice.
 This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021].
@@
 Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples.
-Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times.
+Having dated the number of samples equalling the number of shoreline-dated sites, these were then summed, and the entire process repeated a total of 10000 times.
 The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs.
@@
 For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase.
-The overall model likelihood is the product of these individual probabilities and is not, critically [@crema2022; @timpson2021], found by using the values of the final SPD.
+The overall model likelihood is the product of these individual probabilities and is not, critically, found by using the values of the final SPD [@crema2022; @timpson2021].
 From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model.
@@
 ```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates.
-A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model.
+A) Summed probability of shoreline-dated sites (n = 921) compared to an exponential null model.
 B) SSPD compared to a logistic null model.
@@
 Of the contrasted models, the BIC values indicate that the exponential model is the best candidate.
-Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A).
+Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A).
 Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c.
@@
 7900 BCE.
-It has previously been suggested that a change in mobility patterns occurs some time around 8500--8000 BCE, involving a reduction in residential mobility [e.g.
+It has previously been suggested that a change in mobility patterns occurs sometime around 8500--8000 BCE, involving a reduction in residential mobility [e.g.
 @damlien2018b; @roalkvam2022].
@@
 
-Subsequently, the SSPD deviates negatively from the exponential null from c.
+Subsequently, the SSPD deviates negatively from the exponential null model from c.
 6800 to c.
@@
 This is followed by a positive deviation in the last centuries before 5000 BCE.
-These developments should possibly been seen in relation to simultaneous developments in the inland areas of southern Norway.
+These developments should possibly be seen in relation to simultaneous developments in the inland areas of southern Norway.
 The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c.
 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018].
-@boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in the utilisation of coastal areas.
+@boaz1999 related this to an economic reorganisation, involving a more focussed exploitation of the interior and a reduction in the utilisation of coastal areas.
 This period is followed by a marked reduction in inland activity, lasting until c.
@@
 An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline.
-If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
+If this is the case, the proxy would still reflect the degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
 Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.
@@
 
-This paper has presented a new methodology for assembling and assessing the sum of a large number of shoreline dates.
+This paper has presented a new methodology for assembling and assessing the sum of many shoreline dates.
 This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021].
@@
 
-The main results was that the RSPD is consistent with the logistic model fit to the data.
+The main results were that the RSPD is consistent with the logistic model fit to the data.
 This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from the first centuries after 8000 BCE.
@@
 While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
-This is also relevant for the treatment of RSPDs in the region, as we suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a contributor to the absence of radiocarbon dates from this earliest period.
+This is also relevant for the treatment of RSPDs in Scandinavia more broadly, as we suggested that mobility patterns have relevance for the kind of pyrotechnologies that were utilised in the first period of human occupation in the region, and that the technologies utilised largely explain the absence of radiocarbon dates from this earliest period.
 The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, potentially as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].
@@
 @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change.
-In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important.
+In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important.
 In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.

### 2023-09-06 R/0x_radiocarbon_mc2.R

@@ -75,6 +75,20 @@ logi <- JDEoptim(lower = c(0, 0000), upper = c(1, 10000),
 # No search required for uniform
 unif <- -objectiveFunction(pars = NULL, PDarray = pd, type = 'uniform')
 
+expsum <- SPDsimulationTest(c14, calcurve = intcal20,
+                            calrange = c(minage, maxage), pars = exp$par,
+                            type = 'exp', N = 10000)
+logsum <- SPDsimulationTest(c14, calcurve = intcal20,
+                            calrange = c(minage, maxage), pars = log$par,
+                            type = 'logistic', N = 10000)
+unisum <- SPDsimulationTest(c14, calcurve = intcal20,
+                            calrange = c(minage, maxage), pars = NULL,
+                            type = 'uniform', N = 10000)
+plotSimulationSummary(expsum)
+plotSimulationSummary(logsum)
+plotSimulationSummary(unisum)
+
+
 expp <- convertPars(pars = exp$par, years = minage:maxage, type = 'exp')
 expp$model <- "Exponential"
 logip <- convertPars(pars = logi$par, years = minage:maxage, type = 'logistic')

### 2023-09-09 analysis/paper/paper.Rmd
Minor edits, cleaning of R scripts

@@
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-# abstract: |
-#   Text of abstract
-# keywords: |
-#   keyword 1; keyword 2; keyword 3
+abstract: |
+  By developing a methodology for handling a larger number of shoreline dates, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway.
+Both proxies have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating.
+The findings indicates a largely diverging development of the two proxies through the Mesolithic.
+The number of shoreline dated sites undergoes some process of decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period.
+Although this will require further substantiation, we suggest that while not devoid of a demographic signal, mobility patterns is the main determinant for the number of shoreline dated sites considered in the study.
+Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data is more heavily influenced by population dynamics.
+keywords: |
+  Mesolithic; Radiocarbon dates; Shoreline dates; Fennoscandia 
 # highlights: |
@@
 The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample.
-Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden.
+Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden for the Museum of Cultural History.
 Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83].
@@
 Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B).
-These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987; and map in Figure \@ref(fig:shoredate)A].
+These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987, and map in Figure \@ref(fig:shoredate)A].
 The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated.
@@
 
-In the same way is with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above.
+In the same way as with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above.
 Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site.
@@
 
+Finally, analogous to investigatory biases that might impact the frequency of radiocarbon dates from certain time-intervals, for example resulting from targeted sampling directed towards certain periods of interest, the distribution of sites over different elevations in the landscape might be influenced by survey practices.
+In evaluating two larger survey projects within the present study region, @persson2014a and @solheim2017c found that the test-pitting was generally conducted in a comparable degree across elevations, but that altitudes equating to shoreline dates towards the end of the Stone Age might be under-represented [also @solheim2018, 337].
+Moving forward, similar investigations conducted in a more comprehensive manner will be important for properly evaluating the impact investigatory factors might have on the distribution of shoreline dated sites over time.
+This will also benefit from directly accounting for the elevation at which residential and commercial infrastructure development has been undertaken, as this generally dictates where surveys are undertaken, while also relating this to the topographic characteristics of the landscape.
+
 <!-- The logistic growth for the RSPD then indicates that a plateau is reached by the mid 7th millennium BCE and a stable population for the remainder of the Mesolithic.  -->
@@
 
-This paper has presented a new methodology for assembling and assessing the sum of many shoreline dates.
+This paper has presented a new methodology for assembling and assessing the sum of a larger number of shoreline dates.
 This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021].
@@
 In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
-@jorgensen2020; @ordonez2022; @lundstrom2023].
+@hoebe2023; @jorgensen2020; @ordonez2022; @lundstrom2023].
+To account for the confounding effects these might have, a more principled and comprehensive evaluation of how investigatory practices might impact the SPDs would also be beneficial.
+While certainly also pertinent for the RSPD, this has generally been considered more extensively for radiocarbon data than for shoreline dated sites.
 Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
@@
 Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g.
-@lundstrom2021; @manninen2023; @solheim2018].
+@lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigatory effects that might impact the present SPDs inordinately [see e.g.
+@crema2022; @hoebe2023; @rick1987].
 
 Seeing as the implemented method for shoreline dating was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application.
-While we believe the approach represents an improvement over previous approaches, the results should therefore be considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.
+While we believe the approach represents an improvement over previous implementations, the results should therefore be considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.
 

### 2023-09-09 R/01_shoreline_spd.R

@@ -1,3 +1,7 @@
+# The first part of this script prepares data for shoreline dating and the
+# creates the SPD. The second part creates maps and plots exemplifying the
+# method for shoreline dating.
+
 library(tidyverse)
 library(sf)
 library(terra)
@@ -133,7 +137,7 @@ st_write(sites,
          here("analysis/data/derived_data/combined_sites.gpkg"),
          append = FALSE)
 
-# Create map of sites
+#### Create map of sites and plots providing examples of shoreline dating #####
 
 # Exclude out of bounds sites from the two data sets
 surveyed <- filter(surveyed, !askeladden_id %in% out_of_range)
@@ -146,13 +150,6 @@ excavated$lab <- "y"
 worldmap <- st_read(here("analysis/data/raw_data/naturalearth_countries.gpkg"))
 norway <- st_read(here("analysis/data/raw_data/naturalearth_norway.gpkg"))
 muncipalities <- st_read(here("analysis/data/raw_data/municipalities.gpkg"))
-
-# bboxpolys <- st_bbox(sites)
-# bboxpolys[1] <- bboxpolys[1] - 900
-# bboxpolys[3] <- bboxpolys[3] + 900
-# bboxpolys[2] <- bboxpolys[2] - 900
-# bboxpolys[4] <- bboxpolys[4] + 900
-# bboxpolyspoly <- st_as_sf(st_as_sfc(bboxpolys))
 
 sitbbox <- st_bbox(sites)
 sitbbox[1:2] <- sitbbox[1:2] - 1000000

### 2023-09-09 R/02_displacement_prep.R

@@ -1,9 +1,12 @@
+# This script creates 2.5km wide line segments that run perpendicular to the
+# shoreline gradient, interpolates a displacement curve to centre of these,
+# and finds the distribution of sites across them.
+
 library(tidyverse)
 library(sf)
 library(terra)
 library(here)
 library(patchwork)
-# library(spatstat)
 
 sites <- st_read(here("analysis/data/derived_data/combined_sites.gpkg"))
 bmap <- st_read(here("analysis/data/raw_data/naturalearth_norway.gpkg")) %>%

### 2023-09-09 R/03_functions.R

@@ -1,3 +1,6 @@
+# This script defines functions used for Monte Carlo simulation and plotting of
+# the results.
+
 # Reverse a single calendar date to a corresponding elevation
 # Repurposed from Bchron::unCalibrate()
 reverse_shoredate <- function(shoreline_date,
@@ -38,47 +41,6 @@ reverse_shoredate <- function(shoreline_date,
     return(NA)
   }
 }
-
-
-# # Reverse a single calendar date to a corresponding elevation
-# # Repurposed from rcarbon
-# simulation_summary <- function(spd, simulation_results, cut_off, nsim){
-#
-#   # Reformat data frame of normalised simulated SPDs
-#   sim <- simulation_results %>%
-#     filter(bce <= cut_off) %>%
-#     group_by(bce) %>%
-#     dplyr::reframe(prob_sum_normalised, simn) %>%
-#     tidyr::pivot_wider(names_from = simn, values_from = prob_sum_normalised) %>%
-#     tibble::column_to_rownames("bce")
-#
-#   # Format observed SPD and retrieve the probability
-#   fSPD <- as.data.frame(spd) %>%
-#     filter(sum.bce <= cut_off) %>%
-#     mutate(sum.probability = sum.probability/(sum(sum.probability) * 5)) %>%
-#     select(sum.probability) %>%
-#     unlist(use.names = FALSE)
-#
-#   Zsim <- t(apply(sim, 1, scale))
-#   zLo <- apply(Zsim, 1, quantile, prob = 0.025, na.rm = TRUE)
-#   zHi <- apply(Zsim, 1, quantile, prob = 0.975, na.rm = TRUE)
-#
-#   Zscore_empirical <- (fSPD - apply(sim, 1, mean))/apply(sim, 1, sd)
-#   busts <- which(Zscore_empirical < zLo)
-#   booms <- which(Zscore_empirical > zHi)
-#
-#   observedStatistic <- sum(c(zLo[busts] - Zscore_empirical[busts]),
-#                            c(Zscore_empirical[booms] - zHi[booms]))
-#   expectedstatistic <- abs(apply(Zsim, 2, function(x, y){a = x - y; i = which(a < 0); return(sum(a[i]))}, y = zLo)) +
-#     apply(Zsim, 2, function(x, y){a = x - y; i = which(a > 0); return(sum(a[i]))}, y = zHi)
-#
-#
[…]

### 2023-09-09 R/0x_radiocarbon_mc2.R

@@ -265,3 +265,109 @@ exprplt + logrplt + unirplt
 ggsave(here::here("analysis/figures/rcarbon_mc.png"),
        units = "px", width = 4000, height = 1250)
 
+
+
+SPDsimulationTest <- function(data, calcurve, calrange, pars, type, inc=5, N=20000){
+
+  # 1. generate observed data SPD
+  print('Generating SPD for observed data')
+  CalArray <- makeCalArray(calcurve, calrange, inc) # makeCalArray, used for obs and each simulation
+  # x <- phaseCalibrator(data, CalArray, width=200, remove.external = FALSE)
+  x <- pd
+
+  SPD.obs <- as.data.frame(rowSums(x))
+  SPD.obs <- SPD.obs/(sum(SPD.obs) * CalArray$inc)
+  SPD.obs <- SPD.obs[,1]
+
+  # 2. various sample sizes and effective sample sizes
+
+  # number of dates in the entire dataset
+  n.dates.all <- nrow(data)
+
+  # effective number of dates that contribute to the date range. Some dates may be slightly outside, giving non-integer.
+  tmp <- summedCalibrator(data, CalArray, normalise = 'none')
+  n.dates.effective <- round(sum(tmp)*inc,1)
+
+  # number of phases in entire dataset
+  if(!'phase'%in%names(data))data <- binner(data, width=200, calcurve)
+  n.phases.all <- length(unique(data$phase))
+
+  # effective number of phases that contribute to the date range. Some phases may be slightly outside, giving non-integer.
+  n.phases.effective <- round(sum(x)*inc,1)
+
+  # number of phases that are mostly internal to date range, used for likelihoods
+  n.phases.internal <- sum(colSums(x)>=(0.5 /inc))
+
+  # 3. convert best pars to a model
+  print('Converting model parameters into a PDF')
+  model <- convertPars(pars, years=CalArray$cal, type)
+
+  # 4. Generate N simulations
+  print('Generating simulated SPDs under the model')
+  SPD.sims <- matrix(,length(SPD.obs),N) # blank matrix
+
+  # how many phases to simulate, increasing slightly to account for sampling across a range 300 yrs wider
+  np <- round(sum(x)*inc * (1+300/diff(calrange)))
+
+  # Generate simulations
[…]

### 2023-09-09 R/0x_shoreline_monte_carlo2.R

@@ -1,9 +1,10 @@
+# This script performs the Monte Carlo simulations for the shoreline dates
+
 library(ggplot2)
 library(shoredate)
 library(dplyr)
 library(sf)
 library(here)
-library(patchwork)
 library(ADMUR)
 library(DEoptimR)
 
@@ -85,8 +86,8 @@ logip$year <- logip$year + 3900 # Transforming back to BP
 
 
 # Save
-save(pd, expp, logip, unifp, maxage, minage,
-     file = here("analysis/data/derived_data/shore_models_pd.RData"))
+# save(pd, expp, logip, unifp, maxage, minage,
+#      file = here("analysis/data/derived_data/shore_models_pd.RData"))
 load(here("analysis/data/derived_data/shore_models_pd.RData"))
 
 # Combine models for plotting
@@ -134,7 +135,7 @@ sumdatesdf <- filter(sumdatesdf, sum.bce <= max(expp$bce))
 sumdatesdf$probability <- sumdatesdf$sum.probability /
                           (sum(sumdatesdf$sum.probability) * 5)
 
-##### Monte Carlo simulation - Exponential function #####
+##### Monte Carlo simulation - Exponential #####
 
 # Create data frame to hold simulated dates
 random_dates <- data.frame(matrix(nrow = ssize*nsim, ncol = 3))
@@ -303,7 +304,6 @@ plotSimulationSummary(log_summary)
 logplts <- plot_mc(log_summary)
 
 save(logplts, file = "../external_data/shorespd/logplts.rda")
-load(file = "../external_data/shorespd/logi_simresults.rda")
 
 ##### Monte Carlo simulation - Uniform #####
 

### 2023-09-09 R/0xmcmc.R

@@ -1,44 +1,0 @@
-library(ADMUR)
-
-# Load pd, logistic model, and in and max age for RSPD
-load(file = here::here("analysis/data/derived_data/rcarbon_mcmc_data.RData"))
-
-# Rename
-minage_rcarbon <- minage
-maxage_rcarbon <- maxage
-pd_rcarbon <- PD
-log_rcarbon <- log
-
-# Load pd, 4-CPL model, and in and max age for SSPD
-load(file = here::here("analysis/data/derived_data/shore_mcmc_data.RData"))
-
-# Rename
-minage_shore <- minage
-maxage_shore <- maxage
-pd_shore <- pd
-cpl4_shore <- cpl_4
-
-chain_rcarbon <- mcmc(PDarray = pd_rcarbon,
-                      startPars = log_rcarbon$par,
-                      type = 'logistic',
-                      N = 30000,
-                      burn = 2000,
-                      thin = 5,
-                      jumps = 0.025)
-
-chain_shore <- mcmc(PDarray = pd_shore,
-                      startPars = cpl4_shore$par,
-                      type = 'CPL',
-                      N = 30000,
-                      burn = 2000,
-                      thin = 5,
-                      jumps = 0.025)
-
-save(chain_rcarbon, chain_shore,
-     file = here("analysis/data/derived_data/mcmc_chains.RData"))
-
-chain_rcarbon$acceptance.ratio
-chain_shore$acceptance.ratio
-
-chain_rcarbon$all.pars
-

### 2023-09-09 R/0xmean_tpq_ranges.R

@@ -97,27 +97,3 @@ tpqsd <- sd(c(prange, hrange, arange, trange), na.rm = TRUE)
 # save results
 save(tpqmean, tpqsd,
      file = here("analysis/data/derived_data/disp_tpq_ranges.RData"))
-
-library(dplyr)
-
-dispcurves <- get(load(system.file("extdata/displacement_curves.rda",
-              package = "shoredate")))
-
-rocdisp <- dispcurves %>%
-  group_by(name) %>%
-  arrange(bce, .by_group = TRUE) %>%
-  mutate(rate_upper = upperelev - lag(upperelev),
-         rate_lower = lowerelev - lag(lowerelev)) %>%
-  rowwise() %>%
-  mutate(roc_mean = mean(c(rate_upper, rate_lower), na.rm = TRUE))
-
-rocdisp <- rocdisp %>%
-  group_by(name) %>%
-  mutate( roc_mean = rowMeans(select(rocdisp,
-                  c(rate_upper, rate_lower)), na.rm = TRUE))
-
-ggplot(data = rocdisp) +
-  # geom_line(aes(x = bce, y = rate_upper, col = name)) +
-  # geom_line(aes(x = bce, y = rate_lower, col = name)) +
-  geom_line(aes(x = bce, y = roc_mean, col = name)) +
-  scale_x_continuous(breaks = seq(-10000, 0, 1000))

### 2023-09-09 R/10_radiocarbon_cpl_admur.R

@@ -1,3 +1,6 @@
+# This script perform Monte Carlo simulation and model comparison using the
+# radiocarbon data.
+
 library(dplyr)
 library(rcarbon)
 library(ADMUR)
@@ -15,109 +18,23 @@ c14 <- c14 %>% filter(context != "Food crust")
 
 names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
 
+# Use the calibration procedure from rcarbon to identify min and max ages
 minage <- 4500
 maxage <- 10555
 
+# Switch to ADMUR
 CalArray <- makeCalArray(intcal20, calrange = c(minage, maxage))
 PD <- phaseCalibrator(c14, CalArray, remove.external = TRUE)
 
-SPD <- as.data.frame(rowSums(PD))
-# normalise
-SPD <- SPD/(sum(SPD) * CalArray$inc)
-
-CPL.1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction,
-                  PDarray = PD, type = 'CPL', NP = 20)
-CPL.2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3),
-                  fn = objectiveFunction, PDarray = PD, type = 'CPL', NP = 60)
-CPL.3 <- JDEoptim(lower = rep(0, 5), upper = rep(1, 5),
-                  fn = objectiveFunction, PDarray = PD, type = 'CPL', NP = 100)
-CPL.4 <- JDEoptim(lower = rep(0, 7), upper = rep(1, 7),
-                  fn = objectiveFunction, PDarray = PD, type = 'CPL', NP = 140)
 exp <- JDEoptim(lower = -0.01, upper = 0.01, fn = objectiveFunction,
                 PDarray = PD, type = 'exp', NP = 20)
 log <- JDEoptim(lower = c(0, 0000), upper = c(1, 10000), fn = objectiveFunction,
                 PDarray = PD, type = 'logistic', NP = 40)
 uniform <- objectiveFunction(NULL, PD, type = 'uniform')
 
-save(exp, log, uniform, CPL.1, CPL.2, CPL.3, CPL.4,
+save(exp, log, uniform,
      file = here("analysis/data/derived_data/rcarbon_models.RData"))
 load(file = here("analysis/data/derived_data/rcarbon_models.RData"))
-
-CPL1 <- convertPars(pars=CPL.1$par, years=minage:maxage, type='CPL')
[…]

### 2023-09-09 R/bic.R

@@ -1,0 +1,42 @@
+library(ggplot2)
+library(patchwork)
+
+# Load models fit to raadiocarbon SPD
+load(file = here::here("analysis/data/derived_data/rcarbon_models.RData"))
+
+# Find BIC values
+BIC.exp <- 1*log(134) - 2*(-exp$value)
+BIC.log <- 2*log(134) - 2*(-log$value)
+BIC.uniform <- 0 - 2*(-uniform)
+
+# Assemble data frame for plotting
+bics <- data.frame(model = c("Exponential", "Logistic", "Uniform"),
+                   bic = c(BIC.exp, BIC.log, BIC.uniform))
+
+# Create plot
+bicplt_rcarbon <- ggplot(data = bics, aes(x = bic, y = model)) +
+  geom_point(size = 2.5) +
+  labs(x = "BIC", y = "", subtitle = "Models fit to RSPD") +
+  theme_bw()
+
+# Repeat for SPD of shoreline dates
+load(file = here::here("analysis/data/derived_data/shore_models.RData"))
+
+BIC.exp <- 1*log(921) - 2*(-exp$value)
+BIC.log <- 2*log(921) - 2*(-logi$value)
+BIC.uniform <- 0 - 2*(unif)
+
+bics <- data.frame(model = c("Exponential", "Logistic", "Uniform"),
+                   bic = c(BIC.exp, BIC.log, BIC.uniform))
+
+bicplt_shore <- ggplot(data = bics, aes(x = bic, y = model)) +
+  geom_point(size = 2.5) +
+  labs(x = "BIC", y = "", subtitle = "Models fit to SSPD") +
+  theme_bw()
+
+# Combine plots
+bicplt_shore + bicplt_rcarbon
+
+# Save plot
+ggsave(filename = here::here("analysis/figures/bic.png"),
+       units = "px", width = 2000, height = 900)

### 2023-09-20 analysis/paper/paper.Rmd
Cleaning repository

@@
 institute:
-  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
+  - iakh: University of Oslo, Department of Archaeology, Conservation and History
   - mch: University of Oslo, Museum of Cultural History
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/saa.csl" # Insert path for the bib-style
 abstract: |
-  By developing a methodology for handling a larger number of shoreline dates, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway.
-Both proxies have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating.
-The findings indicates a largely diverging development of the two proxies through the Mesolithic.
-The number of shoreline dated sites undergoes some process of decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period.
-Although this will require further substantiation, we suggest that while not devoid of a demographic signal, mobility patterns is the main determinant for the number of shoreline dated sites considered in the study.
-Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data is more heavily influenced by population dynamics.
+  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway.
+Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating.
+The findings indicate a largely diverging development of the two data sets through the Mesolithic.
+The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period.
+Although the precise nature of this discrepancy will require further substantiation, we tentatively suggest that while not devoid of a demographic signal, the number of shoreline dated sites is heavily influenced by mobility patterns.
+Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
 keywords: | 
-  Mesolithic; Radiocarbon dates; Shoreline dates; Fennoscandia 
-# highlights: |
-#   These are the highlights. 
+  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
 ---
-
-<!-- Keywords: `r rmarkdown::metadata$keywords` -->
-
-<!-- Highlights: `r rmarkdown::metadata$highlights` -->
+Keywords: `r rmarkdown::metadata$keywords`
 
@@
 The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)).
-The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility.
+The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility.
 Thus, this geographical and temporal limit are also used for this study.
@@
 The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density.
-If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics.
+If a comparison of these proxies indicate that they do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics.
 The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.
@@
 
-<!-- ## Model comparison and continuous piece-wise linear models -->
-
-<!-- The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs.
-This can be helpful for determining whether the population patterns has followed the trajectories of standard theoretical models.
-It is also a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the overall development of this proxy could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and, as mentioned above, deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.
-As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*.
-This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above.
-The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction.
-This allows for the identification of key demographic events and a conservative description of population dynamics between these.
-The approach thus provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model.
-Following the evaluation of standard models as outlined above, the data were subsequently subjected to the approach presented by @timpson2021.
--->
-
 # Methods and data
@@
 Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83].
-Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c.
+Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to lose its utility c.
 2500 BCE, the Mesolithic is the main focus of this analysis.
@@
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
-Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021].
-The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], and re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.
+Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021].
+The R package *shoredate* v.1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.
 
@@
 Damaged site, such as outskirts of a quarry, and/or very limited database entry.", "Surveyed by archaeologists.
-However, the database entry is extremely limited/unclear, the site geometry is only given\\\\as a point or small automatically generated circle, and/or finds are from the topsoil of a field.", "Likely site but uncertain spatial information.
+Database entry is extremely limited/unclear, the site geometry is only given\\\\as a point or small automatically generated circle, and/or finds are from the topsoil of a field.", "Likely site but uncertain spatial information.
 Typical example is recurring stray finds in a field or other larger area.", "Single stray find or unverified claims/suggestions of possible site.")
@@
 These are then calibrated back to the calendar scale and finally summed.
-Here, this procedure was repeated 10000 times for each null model.
-The 2.5 and 97.5 quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
+Here, this procedure was repeated 10,000 times for each null model.
+The 2.5th and 97.5th percentile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD.
 The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.
@@
 Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples.
-Having dated the number of samples equalling the number of shoreline-dated sites, these were then summed, and the entire process repeated a total of 10000 times.
+Having dated the number of samples equalling the number of shoreline-dated sites, these were then summed, and the entire process repeated a total of 10,000 times.
 The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs.
 This was done using re-purposed code from *ADMUR*.
-
-<!-- ## Continuous piece-wise linear models -->
-<!-- The procedures outlined above represent standard approaches when modelling population dynamics in archaeology.
-These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shore-bound sites over time could be made.
-However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case.
-As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics.
-This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above.
-The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these.
-Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.
--->
 
@@
 
-For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E).
+For the RSPD, the logistic model achieved a p-value of 0.072 and could therefore not be rejected (Figure \@ref(fig:spdmc)E).
 As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development.
@@
 
-<!-- Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates.
--->
-
-All of the standard models could be rejected for the SSPD.
+All the standard models could be rejected for the SSPD.
 Of the contrasted models, the BIC values indicate that the exponential model is the best candidate.
 Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A).
-Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c.
+Some significant deviations occur with the model over-predicting the frequency of sites in the period from c.
 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE.
@@
 
-<!-- For the SSPD all of the standard models could be rejected.
-The result of subjecting the SSPD to the ADMUR workflow identified 4-CPL as the best model.
-An important caveat here is that models with a complexity beyond 4-CPL consistently failed to converge.
-A fuller reporting of this convergence issue, including the exploration of different parameter settings, different sample sizes and alternative optimization algorithms is given in the supplementary material.
-Subjecting the 4-CPL model to the goodness-of-fit test gave p =, meaning that the data nonetheless could be expected under this model.
-The accepted models that underlies the comparison to follow are therefore the logistic model for the RSDP and the 4-CPL for the SSPD.
--->
-
-<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
 ```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs. Lower values indicate a better model.", out.width = "500px", fig.align="center", regfloat = TRUE}
@@
 What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years.
-These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022].
+These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @åstveit2014; @bjerck2017; @damm2022].
 A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf.
@@
 A common interpretation of the Early Mesolithic in Norway is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g.
-@bjerck2017; @bjerck2008; @fuglestvedt2012].
+@bjerck2008; @bjerck2017; @fuglestvedt2012].
 This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g.
@@
 7900 BCE.
-It has previously been suggested that a change in mobility patterns occurs sometime around 8500--8000 BCE, involving a reduction in residential mobility [e.g.
+It has previously been suggested that a change in mobility patterns starts around 8500--8000 BCE, involving a reduction in residential mobility [e.g.
 @damlien2018b; @roalkvam2022].
@@
 The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.
-
-<!-- @berg-hansen2022 found an increase in sites between c.
-8000--7000 BCE in the northern part of the study area of this paper.
-Their analysis covers a more limited temporal span and geographical area, and is based on the aggregation of point-based estimates of shoreline dates within periods of variable length.
-Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint discrete time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].
-However, consistent with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and, although with a smaller magnitude, around 7100--6900 BCE.
--->
 
@@
 However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data.
-While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
+While it is possible that the limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
 A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf.
@@
 Moving forward, similar investigations conducted in a more comprehensive manner will be important for properly evaluating the impact investigatory factors might have on the distribution of shoreline dated sites over time.
-This will also benefit from directly accounting for the elevation at which residential and commercial infrastructure development has been undertaken, as this generally dictates where surveys are undertaken, while also relating this to the topographic characteristics of the landscape.
-
-<!-- The logistic growth for the RSPD then indicates that a plateau is reached by the mid 7th millennium BCE and a stable population for the remainder of the Mesolithic.  -->
-
-<!-- While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain.
--->
-<!-- However, the positive deviation in the SSPD from the logistic model at c.
-does correspond to a similar deviation of the RSPD from the exponential model.
-Matching negative deviations, on the other hand, can be observed around the middle of the 7th century BCE and in the period after 3500 BCE.
-While we are, as mentioned, cautions in giving these much weight in the SSPD alone, the correspondence of these deviations across the two proxies gives some reason to suggests that these might reflect meaningful demographic signals.
-This would also i -->
-
-<!-- This also extends to the results of the RSPD, due to the fairly limited ^14^C sample.
-With these limitations in mind, the approach outlined by @timpson2021 gives a more reasonable framework with which to evaluate the available data, as leads to a more conservative analysis and reduces the amount of over-interpretation that might follow from uncertainties to do with data or methods.
--->
+This will also benefit from directly accounting for the elevation at which residential and commercial infrastructure development has been undertaken, as this generally dictates where surveys are conducted, while also relating this to the topographic characteristics of the landscape.
 
@@
 The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types.
-<!-- and their potential association with overall developments in population density and mobility patterns. -->
 
 The main results were that the RSPD is consistent with the logistic model fit to the data.
-This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from the first centuries after 8000 BCE.
+This is characterised by a rapid logistic growth from the start of the RSPD around 8500 BCE, which plateaus and remains stationary from the first centuries after 8000 BCE.
 This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay.
@@
 While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
-This is also relevant for the treatment of RSPDs in Scandinavia more broadly, as we suggested that mobility patterns have relevance for the kind of pyrotechnologies that were utilised in the first period of human occupation in the region, and that the technologies utilised largely explain the absence of radiocarbon dates from this earliest period.
-The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, potentially as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].
+This is also relevant for the treatment of RSPDs in Scandinavia more broadly, as we suggested that a combination of mobility patterns, pyrotechnologies and potentially limited vegetational cover in the first period of human occupation in the region largely explain the absence of radiocarbon dates from this earliest period.
+The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, possibly as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].
 
@@
 Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g.
-@lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigatory effects that might impact the present SPDs inordinately [see e.g.
+@lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigations that might impact the present SPDs inordinately [see e.g.
 @crema2022; @hoebe2023; @rick1987].
@@
 
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-
 <div id="refs"></div>

### 2023-09-20 R/01_shoreline_spd.R

@@ -6,7 +6,7 @@ library(tidyverse)
 library(sf)
 library(terra)
 library(here)
-library(shoredate)
+library(shoredate) # Note that shoredate has to be v.1.0.2
 library(patchwork)
 
 # For reproducibility
@@ -55,7 +55,7 @@ sites <- rbind(rcolexcshore, rcolsurvq)
 ###### Shoreline dating ######
 step_reso = 0.1
 
-# Find weighted means dates for figure X in text
+# Find weighted means dates for Figure 3 in the main text
 wmean_dates <- shoreline_date(sites, elevation = dtm,
                               elev_reso = step_reso,
                               cal_reso = 5,
@@ -68,6 +68,7 @@ for(i in 1:length(wmean_dates)){
 }
 wmeans <- wmeans[wmeans  <= -2500]
 
+# Perform shoreline dating using the probabilistic method
 shorelinedates <- shoreline_date(sites,
                                  elevation = dtm,
                                  elev_reso = step_reso,
@@ -79,8 +80,6 @@ shorelinedates <- shoreline_date(sites,
 save(step_reso, wmeans, shorelinedates,
      file = here("analysis/data/derived_data/sdates.RData"))
 
-load(file = here("analysis/data/derived_data/sdates.RData"))
-
 # Sum the probability distributions for the shoreline dates
 sumdates <- sum_shoredates(shorelinedates)
 sumdatesdf <- as.data.frame(sumdates) %>%
@@ -89,7 +88,8 @@ sumdatesdf <- filter(sumdatesdf, sum.bce <= -2500)
 
 # Exclude sites that were given a NA date (elevation equates to a date older
 # than 9465 BCE or younger than 2500 BCE).
-                  # Younger start date than 2500 BCE
+
+                  # Later start date than 2500 BCE
 out_of_range <- c("159969", "94301-1", "144111-1", "144113-1", "144114-1",
                   "138457-1", "230584-0", "265781-0", "265784-0",
                   # Earlier start date than 9465 BCE
@@ -99,12 +99,23 @@ out_of_range <- c("159969", "94301-1", "144111-1", "144113-1", "144114-1",
 
 combined_sites <- filter(sites, !ask_id %in% out_of_range)
 
+st_write(combined_sites,
+         here("analysis/data/derived_data/combined_sites.gpkg"),
+         append = FALSE)
+
+# Exclude out of bounds sites from the two data sets for plotting
+surveyed <- filter(surveyed, !askeladden_id %in% out_of_range)
[…]

### 2023-09-20 R/02_displacement_prep.R

@@ -40,7 +40,7 @@ endline <- st_cast(c(st_point(c(xx,yy)),  st_point(c(xx2,yy2))), "LINESTRING")
 # Find line between closest points on the two lines
 perpline <- st_nearest_points(startline, endline)
 # Find number of increments given the distance between the lines and a set
-# increment (5km)
+# increment (2km)
 ninc <- round(st_length(perpline)/increment)
 # Create ninc number of points at regular intervals along this line
 incpts <- st_line_sample(perpline, ninc) %>%
@@ -108,12 +108,15 @@ for(i in 1:nrow(incpolys)){
   incpolys[i,]$disp <- list(shoredate::interpolate_curve(target = inccents[i,],
                                                          cal_reso = 5))
 }
-save(incpolys,
-         file = here("analysis/data/derived_data/displacement_polygons.RData"))
 
+# Assign weight to the polygon features based on the distribution of sites
 incpolys$dens <- lengths(st_intersects(incpolys, sites)) /
   sum(lengths(st_intersects(incpolys, sites)))
 
+save(incpolys,
+     file = here("analysis/data/derived_data/displacement_polygons.RData"))
+
+# Retrieve displacement curves and weights for plotting
 displist <- list()
 for(i in 1:nrow(incpolys)){
   tmp_curve <- incpolys[i,]$disp[[1]][[1]]
@@ -124,7 +127,7 @@ for(i in 1:nrow(incpolys)){
 
 dispcurves <- do.call(rbind, displist)
 
-displt <- ggplot(dispcurves) +
+displt <- ggplot(as.data.frame(dispcurves)) +
   ggplot2::geom_ribbon(ggplot2::aes(x = bce,
                                     ymin = lowerelev,
                                     ymax = upperelev,

### 2023-09-20 R/03_functions.R

@@ -2,7 +2,7 @@
 # the results.
 
 # Reverse a single calendar date to a corresponding elevation
-# Repurposed from Bchron::unCalibrate()
+# Repurposed from Bchron.
 reverse_shoredate <- function(shoreline_date,
                               displacement_curve,
                               elev_reso){

### 2023-09-20 R/04_shoreline_monte_carlo.R

@@ -1,12 +1,14 @@
 # This script performs the Monte Carlo simulations for the shoreline dates
 
 library(ggplot2)
-library(shoredate)
 library(dplyr)
 library(sf)
 library(here)
 library(ADMUR)
 library(DEoptimR)
+
+# Due to how results are returned, shoredate has to be version  < 1.1.0
+library(shoredate)
 
 set.seed(1)
 
@@ -84,11 +86,14 @@ logip <- convertPars(pars = logi$par, years = minage2:maxage2, type = 'logistic'
 logip$model <- "Logistic"
 logip$year <- logip$year + 3900 # Transforming back to BP
 
+# Save models for BIC
+save(exp, unif, logi,
+     file = here("analysis/data/derived_data/shore_models.RData"))
 
-# Save
-# save(pd, expp, logip, unifp, maxage, minage,
-#      file = here("analysis/data/derived_data/shore_models_pd.RData"))
-load(here("analysis/data/derived_data/shore_models_pd.RData"))
+# Save pd array, models and min and max age
+save(pd, expp, logip, unifp, maxage, minage,
+     file = here("analysis/data/derived_data/shore_models_pd.RData"))
+# load(here("analysis/data/derived_data/shore_models_pd.RData"))
 
 # Combine models for plotting
 shoremodels <- rbind(expp, logip, unifp)

### 2023-09-20 R/05_radiocarbon_monte_carlo.R

@@ -1,5 +1,4 @@
-# This script perform Monte Carlo simulation and model comparison using the
-# radiocarbon data.
+# This script performs Monte Carlo simulation using the radiocarbon data.
 
 library(dplyr)
 library(rcarbon)
@@ -16,13 +15,21 @@ source(here("R/03_functions.R"))
 c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
 c14 <- c14 %>% filter(context != "Food crust")
 
-names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
-
 # Use the calibration procedure from rcarbon to identify min and max ages
-minage <- 4500
-maxage <- 10555
+caldates <- calibrate(x = c14$c14_bp, normalised = FALSE,
+                      errors = c14$error, calCurves = "intcal20")
+c14spd <- spd(caldates, timeRange = c(12000, 4500))
+
+max(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
+min(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
+
+minage <- min(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
+maxage <- max(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
+
 
 # Switch to ADMUR
+names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
+
 CalArray <- makeCalArray(intcal20, calrange = c(minage, maxage))
 PD <- phaseCalibrator(c14, CalArray, remove.external = TRUE)
 
@@ -34,7 +41,8 @@ uniform <- objectiveFunction(NULL, PD, type = 'uniform')
 
 save(exp, log, uniform,
      file = here("analysis/data/derived_data/rcarbon_models.RData"))
-load(file = here("analysis/data/derived_data/rcarbon_models.RData"))
+save(exp, log, uniform, PD,
+     file = here("analysis/data/derived_data/rcarbon_models_pd.RData"))
 
 c14$datingType <- '14C'
 
@@ -47,10 +55,6 @@ logsum <- SPDsimulationTest(c14, calcurve = intcal20,
 unisum <- SPDsimulationTest(c14, calcurve = intcal20,
                             calrange = c(minage, maxage), pars = NULL,
                             type = 'uniform', N = 10000)
-
-save(expsum, logsum, unisum,
-     file = here("analysis/data/derived_data/admur_model_test.RData"))
[…]

### 2023-09-20 R/06_bic.R

@@ -1,7 +1,9 @@
+# Compare model performances for the SPD of each dating method using BIC
+
 library(ggplot2)
 library(patchwork)
 
-# Load models fit to raadiocarbon SPD
+# Load models fit to radiocarbon SPD
 load(file = here::here("analysis/data/derived_data/rcarbon_models.RData"))
 
 # Find BIC values

### 2023-09-20 R/08_map_plots.R

@@ -1,0 +1,169 @@
+#### Create maps and plots providing example of shoreline dating #####
+
+library(ggplot2)
+library(patchwork)
+library(sf)
+library(here)
+library(shoredate)
+
+load(here("analysis/data/derived_data/prepared_sitedata.RData"))
+load(here("analysis/data/derived_data/rcarbon_models_pd.RData"))
+surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
+excavated <- st_zm(read_sf(here("analysis/data/raw_data/excavated_sites.gpkg")))
+
+# Create label column for plotting
+surveyed <- surveyed %>%  dplyr::mutate(lab = ifelse(quality < 4, "t", "f"))
+excavated$lab <- "y"
+
+worldmap <- st_read(here("analysis/data/raw_data/naturalearth_countries.gpkg"))
+norway <- st_read(here("analysis/data/raw_data/naturalearth_norway.gpkg"))
+muncipalities <- st_read(here("analysis/data/raw_data/municipalities.gpkg"))
+
+sitbbox <- st_bbox(sites)
+sitbbox[1:2] <- sitbbox[1:2] - 1000000
+sitbbox[3:4] <- sitbbox[3:4] + 1000000
+boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
+
+studyareabox <- st_bbox(sites)
+studyareabox[1] <- studyareabox[1] - 15000
+studyareabox[3] <- studyareabox[3] + 15000
+studyareabox[2] <- studyareabox[2] - 5000
+studyareabox[4] <- studyareabox[4] + 5000
+bboxpolyspoly <- st_as_sf(st_as_sfc(studyareabox))
+
+# Reproject the bounding box to match world map, and crop the world map
+# with the bounding box polygon.
+bound_reproj <- st_transform(boundingpoly, st_crs(worldmap))
+mapcountries <- worldmap %>%
+  dplyr::filter(st_intersects(., bound_reproj, sparse = FALSE))
+count_reproj <- st_transform(mapcountries, st_crs(surveyed))
+
+overview <-
+  ggplot() +
+  geom_sf(data = count_reproj, fill = "grey", colour = NA) +
+  geom_sf(data = bboxpolyspoly,
+          fill = NA, colour = "black", size = 1) +
+  coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
[…]

### 2023-09-20 R/0x_radiocarbon_mc2.R

@@ -1,373 +1,0 @@
-library(dplyr)
-library(rcarbon)
-library(ADMUR)
-library(patchwork)
-library(ggplot2)
-library(DEoptimR)
-
-set.seed(42)
-
-c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
-c14 <- c14 %>% filter(context != "Food crust")
-
-caldates <- calibrate(x = c14$c14_bp, normalised = FALSE,
-                      errors = c14$error, calCurves = "intcal20")
-c14bins <- binPrep(sites =  c14$site_name, ages = c14$c14_bp, h = 200)
-
-# SPD of entire range in use for fitting models
-c14spd <- spd(caldates, bins = c14bins, timeRange = c(12000, 3500))
-
-
-# Using rcarbon, calibrate each date individually or sum dates that are binned
-subcal2 <- subset(caldates, BP <= 12000 & BP >= 4500, p = 0.05)
-c14sub2 <- c14[which.CalDates(caldates, BP <= 12000 &
-                                BP >= 4500, p = 0.05),]
-subbins2 = binPrep(sites =  c14sub2$site_name, ages = c14sub2$c14_bp, h = 200)
-
-# Retrieve binned and unbinned dates (based on code from rcarbon::spd())
-x <- subcal2
-binNames <- unique(subbins2)
-
-calyears <- data.frame(calBP = seq(12000, 4500, -1))
-binnedMatrix <- matrix(NA, nrow= nrow(calyears), ncol = length(binNames))
-colnames(binnedMatrix) <- binNames
-rownames(binnedMatrix) <- rev(calyears[,1])
-
-for(b in 1:length(binNames)){
-  index <- which(subbins2 == binNames[b])
-  slist <- x$grids[index]
-  slist <- lapply(slist,FUN = function(x) merge(calyears, x, all.x = TRUE))
-  slist <- rapply(slist, f = function(x) ifelse(is.na(x), 0, x),
-                  how = "replace")
-  slist <- lapply(slist, FUN = function(x) x[with(x, order(-calBP)), ])
-
-  tmp <- do.call(rbind, slist)
-  tmp <- aggregate(tmp$PrDens, by = list(Category = tmp$calBP), FUN = sum)
-  spdtmp <- tmp[,-1]
-  spdtmp <- as.data.frame(spdtmp)
-  rownames(spdtmp) <- tmp[, 1]
[…]

### 2023-09-20 R/0x_shoreline_cpl.R

@@ -1,92 +1,0 @@
-library(ADMUR)
-library(DEoptimR)
-library(ggplot2)
-library(patchwork)
-
-set.seed(42)
-
-# Load data from shoreline_monte_carlo.R: models fit for MC, the pd and the
-# min and max ages
-load(file = here::here("analysis/data/derived_data/shore_models_pd.RData"))
-
-# Fit CPL-models to shoreline dates
-cpl_1 <- JDEoptim(lower = 0, upper = 1, fn = objectiveFunction, PDarray = pd,
-                  type = 'CPL', NP = 20, trace = TRUE)
-cpl_2 <- JDEoptim(lower = rep(0, 3), upper = rep(1, 3), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 60, trace = TRUE)
-cpl_3 <- JDEoptim(lower = rep(0, 5), upper = rep(1, 5), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 100, trace = TRUE)
-cpl_4 <- JDEoptim(lower = rep(0, 7), upper = rep(1, 7), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 140, trace = TRUE)
-cpl_5 <- JDEoptim(lower = rep(0, 9), upper = rep(1, 9), fn = objectiveFunction,
-                  PDarray = pd, type='CPL', NP = 180, trace = TRUE)
-cpl_6 <- JDEoptim(lower = rep(0, 11), upper = rep(1,11), fn = objectiveFunction,
-                  PDarray = pd, type = 'CPL', NP = 220, trace = TRUE)
-
-save(cpl_1, cpl_2, cpl_3, cpl_4, cpl_5, cpl_6,
-     file = here::here("analysis/data/derived_data/shore_models.RData"))
-load(file = here::here("analysis/data/derived_data/shore_models1.RData"))
-
-cpl1 <- convertPars(pars = cpl_1$par, years = minage:maxage, type = 'CPL')
-cpl1$model <- "1-CPL"
-cpl2 <- convertPars(pars = cpl_2$par, years = minage:maxage, type = 'CPL')
-cpl2$model <- "2-CPL"
-cpl3 <- convertPars(pars = cpl_3$par, years = minage:maxage, type = 'CPL')
-cpl3$model <- "3-CPL"
[…]

### 2023-09-21 README.Rmd
Minor edits

@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Roalkvam, Isak and Solheim, Steinar (2023).
+_`r Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway`_.
+Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Roalkvam, Isak and Solheim, Steinar (`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/xxx/xxx>
 

### 2023-09-21 analysis/paper/paper.Rmd
Minor edits

@@
 The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD.
-Following @timpson2021, the model parameters were identified by MLE search with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006].
+Following @timpson2021, the model parameters were identified by MLE estimation with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006].
 The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*.
@@
 While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation.
-This is also relevant for the treatment of RSPDs in Scandinavia more broadly, as we suggested that a combination of mobility patterns, pyrotechnologies and potentially limited vegetational cover in the first period of human occupation in the region largely explain the absence of radiocarbon dates from this earliest period.
+This is also relevant for the treatment of RSPDs in Scandinavia more broadly, as we suggested that mobility patterns have relevance for the kind of pyrotechnologies that were utilised in the first period of human occupation in the region, and that the technologies utilised largely explain the absence of radiocarbon dates from this earliest period.
 The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, possibly as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].

### 2023-09-21 R/01_shoreline_spd.R

@@ -1,6 +1,6 @@
 # The first part of this script prepares data for shoreline dating and the
-# creates the SPD. The second part creates maps and plots exemplifying the
-# method for shoreline dating.
+# creates the SPD. The second part creates plots exemplifying different ways of
+# summing shoreline dates.
 
 library(tidyverse)
 library(sf)

### 2023-09-21 R/02_displacement_prep.R

@@ -1,4 +1,4 @@
-# This script creates 2.5km wide line segments that run perpendicular to the
+# This script creates 2km wide line segments that run perpendicular to the
 # shoreline gradient, interpolates a displacement curve to centre of these,
 # and finds the distribution of sites across them.
 

### 2023-09-21 R/03_functions.R

@@ -42,7 +42,7 @@ reverse_shoredate <- function(shoreline_date,
   }
 }
 
-# Repurposed from ADMUR
+# Return summary of a Monte Carlo simulation run. Repurposed from ADMUR
 simulation_summary <- function(spd, simulation_results,
                                cut_offs, mod, sample_size){
 
@@ -105,7 +105,8 @@ simulation_summary <- function(spd, simulation_results,
               n.phases.internal=n.phases.internal))
 }
 
-
+# Custom function for plotting results of a Monte Carlo simulation.
+# Based partially on rcarbon.
 plot_mc <- function(mc_summary, xppos, yppos, title = NULL) {
   booms <- mc_summary$timeseries$calBP[mc_summary$timeseries$index == 1]
   busts <- mc_summary$timeseries$calBP[mc_summary$timeseries$index == -1]

### 2023-09-21 R/04_shoreline_monte_carlo.R

@@ -6,9 +6,7 @@ library(sf)
 library(here)
 library(ADMUR)
 library(DEoptimR)
-
-# Due to how results are returned, shoredate has to be version  < 1.1.0
-library(shoredate)
+library(shoredate) # Note that shoredate has to be v.1.0.2
 
 set.seed(1)
 
@@ -160,6 +158,9 @@ random_dates$reverse_elevation <- mapply(reverse_shoredate,
                                          random_dates$displacement_curve,
                                          step_reso)
 
+# To rerun the code below the directory external_data/shorespd/exp/
+# has to be created one level above the location of this R project
+
 simdates <- vector("list", ssize)
 start_time <- Sys.time()
 for (i in 1:nrow(random_dates)){
@@ -216,13 +217,13 @@ mod <- approx(x = (expp$bce * -1) + 1950, y = expp$prob_dens,
 exp_summary <- simulation_summary(SPD, simulation_results,
                                   c(-2500, -9445), mod, ncol(pd))
 
-# ADMUR plot
-plotSimulationSummary(exp_summary)
-
 # Custom plot
 expplts <- plot_mc(exp_summary)
 
-save(expplts, file = "../external_data/shorespd/expplts.rda")
+# ADMUR plot as sanity check
+plotSimulationSummary(exp_summary)
+
+save(expplts, file = here("analysis/data/derived_data/expplts.rda"))
 
 ##### Monte Carlo simulation - Logistic #####
 
@@ -308,7 +309,7 @@ plotSimulationSummary(log_summary)
 # Custom plot
 logplts <- plot_mc(log_summary)
 
-save(logplts, file = "../external_data/shorespd/logplts.rda")
+save(logplts, file = here("analysis/data/derived_data/logplts.rda"))
 
 ##### Monte Carlo simulation - Uniform #####
 
@@ -402,4 +403,4 @@ plotSimulationSummary(uni_summary)
 # Custom plot
 uniplts <- plot_mc(uni_summary)
 
-save(uniplts, file = "../external_data/shorespd/uniplts.rda")
+save(uniplts, file = file = here("analysis/data/derived_data/uniplts.rda"))

### 2023-09-21 R/05_radiocarbon_monte_carlo.R

@@ -1,4 +1,5 @@
-# This script performs Monte Carlo simulation using the radiocarbon data.
+# This script performs Monte Carlo simulation using the radiocarbon data and
+# plots results from both simulations with both shoreline and radiocarbon dates.
 
 library(dplyr)
 library(rcarbon)
@@ -67,9 +68,9 @@ logplotr <- plot_mc(logsum)
 uniplotr <- plot_mc(unisum)
 
 # Load MC plots for shoreline dates, stored externally
-load("../external_data/shorespd/expplts.rda")
-load("../external_data/shorespd/logplts.rda")
-load("../external_data/shorespd/uniplts.rda")
+load(here("analysis/data/derived_data/expplts.rda"))
+load(here("analysis/data/derived_data/logplts.rda"))
+load(here("analysis/data/derived_data/uniplts.rda"))
 
 (expplts + logplts + uniplts)/
 (expplotr + logplotr + uniplotr) +

### 2023-09-22 R/00_dtm_prep.R
Set up DTM prep script

@@ -1,0 +1,49 @@
+library(here)
+library(supportR)
+
+# Due to its size (4.1 GB), the DTM used for shoreline dating in
+# 01_shoreline_spd.R is not provided directly with the repository of this paper,
+# but is instead downloaded in tiled form from a GitHub repository associated
+# with another paper at https://github.com/isakro/assessing.sealevel.dating.
+# As an alternative it is also possible to download the raster data directly
+# from the Norwegian Mapping Authority (https://hoydedata.no/LaserInnsyn2/),
+# making sure the downloaded raster covers the sites provided in
+# surveyed_sites.gpkg and excavated_sites.gpkg in the raw_data directory
+# of this repository.
+
+# The DTM tiles and file are stored in a local directory external from the R
+# project. This has to be created, but can also be changed to another
+# directory of choice. However, this will require editing of the file path to
+# the dtm specified in 01_shoreline_spd.R
+
+# Uncomment to create external directory if does not already exist
+# dir.create(here("../external_data/data/"))
+
+# Define destination of DTM
+destdir <- here("../external_data/data/")
+
+# List files in the GitHub repository folder
+dtm_tiles <- github_ls("https://github.com/isakro/assessing.sealevel.dating",
+                       folder = "analysis/data/raw_data/tiled_dtm")
+
+# Download the files to the destination directory
+for(i in 1:nrow(dtm_tiles)) { #
+  print(paste0(i, "/", nrow(dtm_tiles)))
+  download.file(url = paste0("https://raw.github.com/isakro/assessing.sealevel.dating/master/analysis/data/raw_data/tiled_dtm/",
+                       dtm_tiles$name[i]), destfile = paste0(destdir, dtm_tiles$name[i]))
+}
+
+# Retrieve path to DTM tiles
+dtmtiles <- list.files(destdir, pattern= ".*.tif$", full.names = TRUE)
+
+# Load the tiles and merge
+tiles <- sapply(dtmtiles, terra::rast)
+names(tiles)[1:2] <- c('x', 'y')
+dtm <- do.call(terra::merge, tiles)
+
+# Delete the DTM tiles
+files <- list.files(destdir, full.names = TRUE)
+sapply(files, file.remove)
+
+# Save merged raster in the external directory
+terra::writeRaster(dtm, file.path(destdir, "dtm10.tif"), overwrite = TRUE)

### 2023-09-22 R/04_shoreline_monte_carlo.R

@@ -121,10 +121,6 @@ sumdates <- sum_shoredates(shorelinedates)
 sumdatesdf <- as.data.frame(sumdates) %>%
   filter(sum.probability != 0)
 
-# Find density of sites across polygons
-incpolys$dens <- lengths(st_intersects(incpolys, sites)) /
-  sum(lengths(st_intersects(incpolys, sites)))
-
 # Sample size
 ssize = sumdates$dates_n
 

### 2023-09-23 analysis/paper/paper.Rmd
Cleaning repository

@@
 These null models are typically a uniform, exponential or logistic distribution.
-These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE).
+These are generally chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE).
 The result from these simulations is then used to create a 95% critical envelope representing the null model.
@@
 All analyses done in this study were performed using the R programming language [@rcoreteam].
-Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5.
+Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.17605/osf.io/fqwt5.
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
 Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021].
-The R package *shoredate* v.1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.
+The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.
 
@@
 This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE.
-However, properly assessing this relationship and the degree to which it has held constant throughout the period will, as mentioned, depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
+However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.
 
@@
 In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
-@hoebe2023; @jorgensen2020; @ordonez2022; @lundstrom2023].
+@hoebe2023; @manninen2018; @jorgensen2020; @ordonez2022; @lundstrom2023].
 To account for the confounding effects these might have, a more principled and comprehensive evaluation of how investigatory practices might impact the SPDs would also be beneficial.

### 2023-09-23 R/00_dtm_prep.R

@@ -9,20 +9,22 @@ library(supportR)
 # from the Norwegian Mapping Authority (https://hoydedata.no/LaserInnsyn2/),
 # making sure the downloaded raster covers the sites provided in
 # surveyed_sites.gpkg and excavated_sites.gpkg in the raw_data directory
-# of this repository.
+# of this repository. Finally, as this is only used in 01_shoreline_spd.R
+# and the results of this is as saved files with the repository, it is also
+# possible to skip this step if downloading the DTM is undesirable.
 
 # The DTM tiles and file are stored in a local directory external from the R
-# project. This has to be created, but can also be changed to another
+# project. This directory has to be created, but can also be changed to another
 # directory of choice. However, this will require editing of the file path to
 # the dtm specified in 01_shoreline_spd.R
 
-# Uncomment to create external directory if does not already exist
+# Uncomment to create external directory if this does not already exist
 # dir.create(here("../external_data/data/"))
 
-# Define destination of DTM
-destdir <- here("../external_data/data/")
+# Define external destination of DTM
+destdir <- here("../external_data/dtm/")
 
-# List files in the GitHub repository folder
+# List files in the GitHub repository folder to be downloaded
 dtm_tiles <- github_ls("https://github.com/isakro/assessing.sealevel.dating",
                        folder = "analysis/data/raw_data/tiled_dtm")
 

### 2023-09-23 R/01_shoreline_spd.R

@@ -1,6 +1,6 @@
-# The first part of this script prepares data for shoreline dating and the
+# The first part of this script prepares data for shoreline dating and then
 # creates the SPD. The second part creates plots exemplifying different ways of
-# summing shoreline dates.
+# summing shoreline dates. Note that
 
 library(tidyverse)
 library(sf)
@@ -15,7 +15,7 @@ set.seed(1)
 # Load data
 surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
 excavated <- st_zm(read_sf(here("analysis/data/raw_data/excavated_sites.gpkg")))
-dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+dtm <- rast(here("../external_data/dtm/dtm10.tif")) # see 00_dtm_prep.R
 site_dat <- read.csv(here("analysis/data/raw_data/sites.csv"))
 
 ###### Prepare site data for shoreline dating ######

### 2023-09-23 R/04_shoreline_monte_carlo.R

@@ -1,4 +1,9 @@
-# This script performs the Monte Carlo simulations for the shoreline dates
+# This script performs the Monte Carlo simulations for the shoreline dates.
+# Note that this is highly inefficient and takes several days, if not weeks to
+# execute. To assess the script it is possible to adjust the
+# nsim object below, which is set to 10000. Setting nsim <- 100 means that
+# each of the three runs should, depending on hardware, execute in an hour or
+# so.
 
 library(ggplot2)
 library(dplyr)
@@ -9,6 +14,21 @@ library(DEoptimR)
 library(shoredate) # Note that shoredate has to be v.1.0.2
 
 set.seed(1)
+
+# Due to memory and storage consideration, the results of the Monte Carlo
+# simulations are stored externally, in the same directory as that created for
+# 00_dtm_prep.R. The results after passing them through
+# simulation_summary(), defined in 03_functions.R, are provided with the
+# repository, and so it is possible to load in these in to inspect the final
+# results (indicated by commented out code for each of the three models below).
+
+# Uncomment code to create directories to hold simulation results, if these do
+# not already exist. Alternatively, respecify folder paths to a desired location
+# in the simulation loops below.
+# subfolders <- c("exp", "logi", "uni")
+# for(i in 1:length(subfolders)){.
+#   print(paste0(here("../external_data/data/shorespd/"), subfolders[i]))
+# }
 
 # Source functions
 source(here("R/03_functions.R"))
@@ -41,10 +61,6 @@ pd <- as.data.frame(sdates) %>%
   tidyr::pivot_wider(names_from = site, values_from = probability) %>%
   tibble::column_to_rownames("bp")
 pd <- sweep(pd, 2, colSums(pd),"/")
-
-# Create SPD using the ADMUR approach
-SPD <- as.data.frame(rowSums(pd))
-SPD <- SPD/( sum(SPD) *5 )
 
 # Find mininum and maximum ages (BP)
 minage <- min(as.numeric(row.names(pd)))
@@ -80,7 +96,9 @@ logi <- JDEoptim(lower = c(0, 0000), upper = c(1, 10000),
                  fn = objectiveFunction, PDarray = pd2, type = 'logistic',
                  NP = 40, trace = TRUE)
 
[…]

### 2023-09-23 R/05_radiocarbon_monte_carlo.R

@@ -67,7 +67,7 @@ expplotr <- plot_mc(expsum)
 logplotr <- plot_mc(logsum)
 uniplotr <- plot_mc(unisum)
 
-# Load MC plots for shoreline dates, stored externally
+# Load MC plots for shoreline dates
 load(here("analysis/data/derived_data/expplts.rda"))
 load(here("analysis/data/derived_data/logplts.rda"))
 load(here("analysis/data/derived_data/uniplts.rda"))

### 2023-09-23 R/07_tpq_ranges.R

@@ -14,7 +14,7 @@ centrepoints <- sf::st_read(
 # Retreieve the point on the Horten isobase
 horten <- centrepoints[1,]
 
-# Create a sequence of eleveation values from well above the marine limit for
+# Create a sequence of elevation values from well above the marine limit for
 # the curves and down to 20 masl
 elevations <- seq(190, 1, -1)
 
@@ -94,6 +94,6 @@ for(i in 1:length(tsdates)){
 tpqmean <- mean(c(prange, hrange, arange, trange), na.rm = TRUE)
 tpqsd <- sd(c(prange, hrange, arange, trange), na.rm = TRUE)
 
-# save results
+# Save results to be used in the main text
 save(tpqmean, tpqsd,
      file = here("analysis/data/derived_data/disp_tpq_ranges.RData"))

### 2023-09-23 R/08_map_plots.R

@@ -74,7 +74,8 @@ sa <- ggplot() +
                                paste0("Surveyed sites,\nexcluded (n = ",
                                       nrow(dplyr::filter(surveyed,
                                                          quality > 3)) ,")"),
-                               paste0("Excavated, shoreline\ndated sites (n = ", nrow(excshore) ,
+                               paste0("Excavated, shoreline\ndated sites (n = ",
+                                      nrow(excshore) ,
                                       ")")),
                     values = c("t" = "darkgoldenrod1",
                                "f" = "black",

### 2023-09-24 README.Rmd
Set up renv, updated readme

--- 2023-01-23 README.Rmd
@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
 ```
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Roalkvam, Isak <a href="https://orcid.org/0000-0001-6974-1374"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> and Solheim, Steinar <a href="https://orcid.org/0000-0001-8293-8147"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> (2023).
+_`r Title`_.
+Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Roalkvam, Isak and Solheim, Steinar (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 

### 2023-09-24 README.Rmd
Updated description, added DOI to Zenodo repository

@@
 
-# se.norway.shoredate.14c
+# Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway
 
@@
 _`r Title`_.
-Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`.
 Online at <https://doi.org/xxx/xxx>
@@
 
-> Roalkvam, Isak and Solheim, Steinar (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Roalkvam, Isak <a href="https://orcid.org/0000-0001-6974-1374"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> and Solheim, Steinar <a href="https://orcid.org/0000-0001-8293-8147"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> (`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/10.5281/zenodo.8373857>
 
 ## Contents
+
+The **R** directory contains scripts for performing the analyses and create the
+figures used in the paper.
 
@@
   - [:file\_folder: paper](/analysis/paper): R Markdown source document
-    for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
-    and tables in this file)
+    for manuscript. It also has a rendered version of the manuscript,
+    `paper.pdf`, suitable for reading.
   - [:file\_folder: data](/analysis/data): Data used in the analysis.
@@
     illustrations
-  - [:file\_folder:
-    supplementary-materials](/analysis/supplementary-materials):
-    Supplementary materials including notes and other documents
-    prepared and collected during the analysis.
 

### 2023-09-25 analysis/paper/paper.Rmd
Updated description, added DOI to Zenodo repository
Minor edits to paper. Updated gitignore

@@
 All analyses done in this study were performed using the R programming language [@rcoreteam].
-Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.17605/osf.io/fqwt5.
+Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.5281/zenodo.8373857.
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
@@
 
+# Declaration of competing interest{-}
+The authors have no competing interests to declare.
+
 \newpage

### 2023-11-28 analysis/paper/paper.Rmd
Additional analyses after PCI review

@@
 The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period.
-Although the precise nature of this discrepancy will require further substantiation, we tentatively suggest that while not devoid of a demographic signal, the number of shoreline dated sites is heavily influenced by mobility patterns.
+Although the precise nature of this discrepancy will require further substantiation, we tentatively suggest that while not devoid of a demographic signal, the number of shoreline dated sites could be more heavily influenced by mobility patterns.
 Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
@@
 This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details].
-Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf.
-@weninger2015, 545], each shoreline date is normalised to sum to unity before being summed.
+Given that the shoreline displacement curves have no inversions and should therefore be commutative [see @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed.
 To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.
@@
 Finally, a positive deviation occurs again around 5300--5000 BCE.
-While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf.
-@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.
+While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.
 
@@
 These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @åstveit2014; @bjerck2017; @damm2022].
-A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf.
-@damm2022; @sørensen2014].
+A possible lack of reliable access to firewood has also been seen as a possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; see also @damm2022; @sørensen2014].
 While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.
@@
 The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
-The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf.
-@damm2022] and low population numbers.
+The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [@damm2022] and low population numbers.
 
@@
 However, after this the population dynamics indicated by the radiocarbon record is found to have been characterised by a cessation of growth, rather than the clear decrease suggested by the regional studies in southern Norway.
-Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [cf.
-@crema2017].
+Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [see @crema2017].
 
@@
 While it is possible that the limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
-A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf.
-@tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.
+A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [@tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.
 
@@
 While certainly also pertinent for the RSPD, this has generally been considered more extensively for radiocarbon data than for shoreline dated sites.
-Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf.
-@timpson2021].
+Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [see @timpson2021].
 Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g.

### 2023-11-28 R/00_dtm_prep.R

@@ -10,8 +10,8 @@ library(supportR)
 # making sure the downloaded raster covers the sites provided in
 # surveyed_sites.gpkg and excavated_sites.gpkg in the raw_data directory
 # of this repository. Finally, as this is only used in 01_shoreline_spd.R
-# and the results of this is as saved files with the repository, it is also
-# possible to skip this step if downloading the DTM is undesirable.
+# and the results of this provided as saved files with the repository, it is
+# also possible to skip this step if downloading the DTM is undesirable.
 
 # The DTM tiles and file are stored in a local directory external from the R
 # project. This directory has to be created, but can also be changed to another
@@ -21,7 +21,7 @@ library(supportR)
 # Uncomment to create external directory if this does not already exist
 # dir.create(here("../external_data/data/"))
 
-# Define external destination of DTM
+# Define external destination for DTM
 destdir <- here("../external_data/dtm/")
 
 # List files in the GitHub repository folder to be downloaded

### 2023-11-28 R/03_functions.R

@@ -50,7 +50,7 @@ simulation_summary <- function(spd, simulation_results,
   SPD.obs <- spd[,1]
 
   # Format simulation results for ADMUR
-  SPD.sims <- simresults %>%
+  SPD.sims <- simulation_results %>%
     dplyr::filter(bce <= cut_offs[1] & bce >= cut_offs[2]) %>%
     dplyr::mutate(bp = (bce * -1) + 1950) %>%
     arrange(bp) %>%

### 2023-11-28 R/04_shoreline_monte_carlo.R

@@ -148,7 +148,7 @@ sumdatesdf <- as.data.frame(sumdates) %>%
   filter(sum.probability != 0)
 
 # Find sample size
-ssize = sumdates$dates_n
+ssize <- sumdates$dates_n
 
 # Set number of simulations
 nsim <- 10000
@@ -237,7 +237,7 @@ mod <- approx(x = (expp$bce * -1) + 1950, y = expp$prob_dens,
               ties = 'ordered', rule = 2)$y
 
 # Model summary (can also be plotted with ADMUR)
-exp_summary <- simulation_summary(SPD, simulation_results,
+exp_summary <- simulation_summary(SPD, simresults,
                                   c(-2500, -9445), mod, ncol(pd))
 
 save(exp_summary,
@@ -327,7 +327,7 @@ mod <- approx(x = (logip$bce * -1) + 1950, y = logip$prob_dens,
               ties = 'ordered', rule = 2)$y
 
 # Summary results
-log_summary <- simulation_summary(SPD, simulation_results,
+log_summary <- simulation_summary(SPD, simresults,
                                   c(-2500, -9445), mod, ncol(pd))
 
 save(log_summary,
@@ -425,7 +425,7 @@ mod <- approx(x = unifp$bce, y = unifp$prob_dens,
               ties = 'ordered', rule = 2)$y
 
 # Model summary (can also be plotted with ADMUR)
-uni_summary <- simulation_summary(SPD, simulation_results,
+uni_summary <- simulation_summary(SPD, simresults,
                                   c(-2500, -9445), mod, ncol(pd))
 
 save(uni_summary,

### 2023-11-28 R/05_radiocarbon_monte_carlo.R

@@ -21,9 +21,6 @@ caldates <- calibrate(x = c14$c14_bp, normalised = FALSE,
                       errors = c14$error, calCurves = "intcal20")
 c14spd <- spd(caldates, timeRange = c(12000, 4500))
 
-max(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
-min(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
-
 minage <- min(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
 maxage <- max(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
 

### 2023-11-28 R/09_sdate_c14sites.R

@@ -1,0 +1,358 @@
+library(here)
+library(shoredate)
+library(terra)
+library(sf)
+library(ggplot2)
+library(dplyr)
+library(ADMUR)
+
+dtm <- rast(here("../external_data/dtm/dtm10.tif")) # see 00_dtm_prep.R
+# Load RSPD models and probability distribution
+load(here("analysis/data/derived_data/rcarbon_models_pd.RData"))
+# Load excavated sites data
+excavated <- st_zm(read_sf(here("analysis/data/raw_data/excavated_sites.gpkg")))
+# Background map for plotting
+bmap <- st_read(here("analysis/data/raw_data/naturalearth_norway.gpkg")) %>%
+  st_transform(st_crs(excavated))
+
+# Find names of sites with radiocarbon dates
+c14_names <- tools::file_path_sans_ext(colnames(PD))
+
+# Filter for sites with radiocarbon dates
+c14_sites <- excavated %>%
+  dplyr::filter(site_name %in% c14_names)
+
+# Find the elevations of the sites with radiocarbon dates
+c14_sites$elev <- terra::extract(dtm, vect(c14_sites), fun = mean)[, -1]
+
+# Define here for use in Monte Carlo simulation as well
+step_reso = 0.1
+
+# Perform shoreline dating
+c14sites_shoredates <- shoreline_date(c14_sites,
+                                 elevation = c14_sites$elev,
+                                 elev_reso = step_reso,
+                                 cal_reso = 5,
+                                 sparse = TRUE,
+                                 verbose = FALSE)
+
+# Save results
+save(c14sites_shoredates,
+     file = here("analysis/data/derived_data/sdates_c14.RData"))
+load(here("analysis/data/derived_data/sdates_c14.RData"))
+
+# Create ADMUR compatible PD and SPD
+dates_dfs <- lapply(c14sites_shoredates, as.data.frame)
+sdates <- do.call(rbind, dates_dfs)
+sdates$site <- rep(1:length(dates_dfs),
+                   each = nrow(sdates)/length(dates_dfs))
+
+pd <- as.data.frame(sdates) %>%
+  dplyr::filter(bce <= -2550 & bce >= -9445) %>%
+  dplyr::mutate(bp = (bce * -1) + 1950) %>%
+  dplyr::select(-bce) %>%
+  dplyr::group_by(bp) %>%
+  dplyr::filter(!is.na(probability)) %>%
[…]

### 2023-11-28 R/09_sdate_c14sites.R
Minor edit to SPD

@@ -56,7 +56,7 @@ pd <- as.data.frame(sdates) %>%
   dplyr::group_by(site) %>%
   tidyr::pivot_wider(names_from = site, values_from = probability) %>%
   tibble::column_to_rownames("bp")
-pd <- sweep(pd, 2, colSums(pd),"/")
+# pd <- sweep(pd, 2, colSums(pd),"/")
 
 SPD <- as.data.frame(rowSums(pd))
 SPD <- SPD/( sum(SPD) *5 )
@@ -69,7 +69,7 @@ c14_shoresumdf <- as.data.frame(c14_shoresum[[1]]) %>%
 c14_shoresumdf <- filter(c14_shoresumdf, bce <= -2550 & bce >= -9445)
 # Normalise after cut-off
 c14_shoresumdf$probability <- c14_shoresumdf$probability /
-                              sum(c14_shoresumdf$probability)
+                              (sum(c14_shoresumdf$probability)* 5)
 
 # Retrieve logistic model used for RSPD
 log_c14 <- convertPars(pars = log$par,

### 2023-12-31 analysis/paper/paper.Rmd
Final edits before re-submission for PCI

@@
 The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period.
-Although the precise nature of this discrepancy will require further substantiation, we tentatively suggest that while not devoid of a demographic signal, the number of shoreline dated sites could be more heavily influenced by mobility patterns.
-Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
+The precise nature of this discrepancy will necessitate further substantiation and the method of shoreline dating itself still requires further testing and assessment.
+Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns.
+Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
 keywords: | 
@@
 During the last decade, the frequency distribution of radiocarbon dates has become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022].
-This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2023; see also @persson2009].
+This has also been the case for Norwegian Stone Age archaeology [e.g.
+@bergsvik2021; @solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @jorgensen2020b; @lundstrom2023; @persson2018; see also @persson2009].
 
@@
 All sites with a quality score of 4 or worse were excluded from further analysis.
-Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded.
+Any sites situated at elevations that would result in a shoreline date earlier than 9465 BCE were then excluded.
 This marks the latest start-date among the employed displacement curves, and no sites in the region have yet been conclusively verified to have an older date than around 9300 BCE [@glorstad2016].
@@
 # Map prepared in 01_shoreline_spd.R 
-knitr::include_graphics(here("analysis/figures/map2.png"))
+knitr::include_graphics(here("analysis/figures/map.png"))
 ```
@@
 Following @timpson2021, the model parameters were identified by MLE estimation with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006].
-The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*.
+The Monte Carlo simulations were then performed using the *SPDsimulationTest* function from *ADMUR*.
 For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD.
@@
 Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B).
-These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987, and map in Figure \@ref(fig:shoredate)A].
+These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Figure \@ref(fig:shoredate)A].
 The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated.
 This is done by inverse distance weighting [e.g.
-@conolly2020], where the relative sea-level change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.
+@conolly2020], where the RSL change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.
 
@@
 
-When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
+When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
 As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years.
@@
 In the same way as with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above.
-Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site.
+Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of RSL change effectively functions as the calibration curve for each site.
 As the shoreline displacement curves are interpolated to the sites based on their location along a south-west/north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase.
@@
 A possible lack of reliable access to firewood has also been seen as a possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; see also @damm2022; @sørensen2014].
+To assess the potential influence of taphonomy on the logistic model for the RSPD, we fit a version of the model that accounts for taphonomy, using the workflow from *ADMUR*.
+As this did not alter the MLE parameters of the model, the results of this exercise have been relegated to the supplementary material.
 While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.
@@
 Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic [e.g.
-@bang-andersen2005; @fuglestvedt2018, 16--17; @lindblom1984; @bjerck2008, 103--105], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.
+@bang-andersen2005; @fuglestvedt2018, 16--17; @lindblom1984; @bjerck2008, 103--105], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population dynamics.
 
@@
 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018].
-@boaz1999 related this to an economic reorganisation, involving a more focussed exploitation of the interior and a reduction in the utilisation of coastal areas.
+@boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in the utilisation of coastal areas.
 This period is followed by a marked reduction in inland activity, lasting until c.
@@
 @boaz1999 associated this period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022].
-The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
+The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233].
 The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.
@@
 There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE.
-However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data.
-While it is possible that the limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].
+An immediate candidate explanation for this decrease could be the so-called Finse cold event, occurring around 6200 BCE, which has been identified in a range of environmental proxy records from Norway [e.g.
+@nesje2005; @nesje2009], and which corresponds to the 8.2 ka cal BP cold event evident in North Atlantic climate records more widely [e.g.
+@alley1997; @alley2005; @seppa2009].
+The possible societal relevance of the Finse event for the Norwegian Mesolithic has been discussed extensively [e.g.
+@breivik2018; @fossum2020; @persson2018].
+However, as the logistic model could not be rejected for the RSPD it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data.
+While it is possible that the limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE that would correspond to the positive deviation in SSPD [see also @solheim2018].
 A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [@tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.
@@
 
-Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g.
+It is important to underscore that underpinning the potential relevance of demographic and cultural factors that have been highlighted here is the fact that the implemented method for shoreline dating is hefted with considerable unresolved questions concerning its applicability and reliability, and the degree to which the site-sea relationship has been stationary through the Mesolithic and within the study area.
+For example, it was in @roalkvam2023 found that the vertical distances between site and sea were associated with a higher uncertainty in the earliest part of the Holocene, due to the rapid shoreline displacement in this period.
+This uncertainty is not properly accounted for in the implemented method, as the MLE parameters of the gamma distribution in use was found when treating all sites older than 2500 BCE in aggregate.
+Additionally, the MLE parameters for the gamma distribution are point-estimates that do not account for the underlying sample size used for deriving them.
+Furthermore, the rapid shoreline displacement in the early Holocene is also a challenge for the geological reconstructions themselves, and there is reason to question whether this is properly accounted for in the compiled time-series for RSL-change in the region [see e.g.
+@romundset2018; @sørensen2014; cf.
+@ashe2019].
+Although these factors will hamper the traditionally employed TPQ approach even more, there is therefore a danger that the severe right skew of the gamma distribution, combined with the rapid shoreline displacement in the earliest period, leads to an overestimation of the antiquity of the sites and the precision of the earliest dates.
+
+While the apparent applicability of shoreline dating is a testament to the importance the sea has played for movement, communication and resource exploitation through the Mesolithic in the region, the immediate inland has also seen some use through the Mesolithic in south-eastern Norway [e.g.
+@brogger1905; @gundersen2013].
+However, with one notable exception [@mjærum2018], radiocarbon dates reflecting the use of coastal hinterland sites are associated with a few site features that are otherwise unrelated to typo-chronological indicators in the lithic inventories on the same sites, which are instead associated with an earlier date than the radiocarbon dates.
+Consequently, these later radiocarbon dates are typically interpreted as either being of a non-anthropogenic nature, or the result of visits of a more ephemeral nature that have succeeded the settlement on the sites [e.g.
+@breivik2018; @schulke2020; @schulke2023; see also @roalkvam2023] -- potentially reflecting an increase in logistic utilisation of the immediate inland areas in the later stages of the Mesolithic.
+While this conclusion has been reached by several researchers independently of each other, thus increasing the confidence that shoreline dating captures the date of settlement on the sites, there is still reason to question whether the available radiocarbon record reflects the true use of the hinterland in later periods.
+If the coastal hinterland has been utilised more actively than what is currently evidenced by the radiocarbon record, the SSPD would be biased towards overestimating the intensity of site frequency in earlier periods.
+
+Another relevant point to mention here is that the RSL change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g.
 @crema2022].
-The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D).
+The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of RSL change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D).
 Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models.
-An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline.
-If this is the case, the proxy would still reflect the degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
+An increased movement instantiated by increased rates of RSL change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline.
+If this is the case, the proxy would still reflect the degree of occupational duration, but this could be a response to the rate of RSL change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments.
 Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.
@@
 @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g.
-@barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change.
+@barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of RSL change.
 In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important.
-In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
+In the same vein, environmental conditions, constraints and developments beyond RSL change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g.
 @hoebe2023; @manninen2018; @jorgensen2020; @ordonez2022; @lundstrom2023].
@@
 Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [see @timpson2021].
-Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g.
+Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed developments [see e.g.
 @lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigations that might impact the present SPDs inordinately [see e.g.
@@
 <div id="refs"></div>
+
+```{r taph, echo = FALSE, fig.cap = "The result of accounting for taphonomic loss in the analysis of the radiocarbon data.
+A) Comparison of MLE parameters for the logistic population dynamics models when both accounting for and disregarding taphonomy using the ADMUR approach.
+These are superimposed on the RSPD.
+As indicated by their perfect overlap, accounting for taphonomy does not impact the MLE parameters.
+B) The taphonomic component of the logistic model estimated when accounting for taphonomy.", out.width = "500px", fig.align="center", regfloat = TRUE}
+knitr::include_graphics(here("analysis/figures/taphonomic.png"))
+```

### 2023-12-31 R/05_radiocarbon_monte_carlo.R

@@ -1,5 +1,8 @@
 # This script performs Monte Carlo simulation using the radiocarbon data and
 # plots results from both simulations with both shoreline and radiocarbon dates.
+# Based on the failed rejection of the logistic model, a version of the model
+# fit when accounting for taphonomy is then created (figure included in
+# as supplementary material)
 
 library(dplyr)
 library(rcarbon)
@@ -24,6 +27,7 @@ c14spd <- spd(caldates, timeRange = c(12000, 4500))
 minage <- min(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
 maxage <- max(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
 
+# c14spd_taph <- transformSPD(c14spd)
 
 # Switch to ADMUR
 names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
@@ -59,6 +63,7 @@ plotSimulationSummary(expsum)
 plotSimulationSummary(logsum)
 plotSimulationSummary(unisum)
 
+
 # Plotting using custom plot function
 expplotr <- plot_mc(expsum)
 logplotr <- plot_mc(logsum)
@@ -75,3 +80,71 @@ load(here("analysis/data/derived_data/uniplts.rda"))
 
 ggsave(here::here("analysis/figures/mc.png"),
        units = "px", width = 4000, height = 1250*2)
+
+
+#### Logistic model with taphonomic loss ####
+
+# Use the default parameter settings from ADMUR, as there are no available
+# prior data on which to adjust these from the region (see the ADMUR vignette)
+logtaph <- JDEoptim(lower = c(0, 0000, 0, -3), upper = c(1, 10000, 20000,0),
+                    fn = objectiveFunction, PDarray = PD, type = 'logistic',
+                    NP = 40, taphonomy = TRUE)
+
+# Retrieve the logistic model fit without accoutning for taphonomy
+pop_log <- convertPars(pars=log$par,
+                       years=minage:maxage,
+                       type='logistic')
+
+# Retrieve the population dynamics component of the logistic model fit when
+# accounting for taphonomy
+pop_log_taph <- convertPars(pars=logtaph$par[1:2],
+                            years=minage:maxage,
+                            type='logistic')
+
+# Retrieve the taphonomic component
+taph_log_taph <- convertPars(pars=logtaph$par[3:4],
+                            years=minage:maxage,
[…]

### 2023-12-31 R/09_sdate_c14sites.R

@@ -1,358 +1,0 @@
-library(here)
-library(shoredate)
-library(terra)
-library(sf)
-library(ggplot2)
-library(dplyr)
-library(ADMUR)
-
-dtm <- rast(here("../external_data/dtm/dtm10.tif")) # see 00_dtm_prep.R
-# Load RSPD models and probability distribution
-load(here("analysis/data/derived_data/rcarbon_models_pd.RData"))
-# Load excavated sites data
-excavated <- st_zm(read_sf(here("analysis/data/raw_data/excavated_sites.gpkg")))
-# Background map for plotting
-bmap <- st_read(here("analysis/data/raw_data/naturalearth_norway.gpkg")) %>%
-  st_transform(st_crs(excavated))
-
-# Find names of sites with radiocarbon dates
-c14_names <- tools::file_path_sans_ext(colnames(PD))
-
-# Filter for sites with radiocarbon dates
-c14_sites <- excavated %>%
-  dplyr::filter(site_name %in% c14_names)
-
-# Find the elevations of the sites with radiocarbon dates
-c14_sites$elev <- terra::extract(dtm, vect(c14_sites), fun = mean)[, -1]
-
-# Define here for use in Monte Carlo simulation as well
-step_reso = 0.1
-
-# Perform shoreline dating
-c14sites_shoredates <- shoreline_date(c14_sites,
-                                 elevation = c14_sites$elev,
-                                 elev_reso = step_reso,
-                                 cal_reso = 5,
-                                 sparse = TRUE,
-                                 verbose = FALSE)
-
-# Save results
-save(c14sites_shoredates,
-     file = here("analysis/data/derived_data/sdates_c14.RData"))
-load(here("analysis/data/derived_data/sdates_c14.RData"))
-
-# Create ADMUR compatible PD and SPD
-dates_dfs <- lapply(c14sites_shoredates, as.data.frame)
-sdates <- do.call(rbind, dates_dfs)
-sdates$site <- rep(1:length(dates_dfs),
-                   each = nrow(sdates)/length(dates_dfs))
-
-pd <- as.data.frame(sdates) %>%
-  dplyr::filter(bce <= -2550 & bce >= -9445) %>%
-  dplyr::mutate(bp = (bce * -1) + 1950) %>%
-  dplyr::select(-bce) %>%
-  dplyr::group_by(bp) %>%
-  dplyr::filter(!is.na(probability)) %>%
[…]

### 2024-01-05 analysis/paper/paper.Rmd
Added link to new version of Zenodo archive
Updates before PCI recommendation

@@
   \usepackage[left]{lineno}
-  \linenumbers
+  # \linenumbers
 # always_allow_html: true
@@
 All analyses done in this study were performed using the R programming language [@rcoreteam].
-Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.5281/zenodo.8373857.
+Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.5281/zenodo.10449324.
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
@@
 
-# Declaration of competing interest{-}
-The authors have no competing interests to declare.
+# Acknowledgement{-}
+We would like to thank David Wright for commenting on an earlier draft of this paper.
+We would also like to thank the reviewers for their constructive feedback and PCI archaeology for facilitating their evaluation of our manuscript.
+
+# Funding{-}
+This research received no external funding.
+
+# Conflict of interest disclosure{-}
+The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article.
+
+# Data, scripts, code, and supplementary information availability{-}
+All data, code, text and figures used for this paper is freely available in a version-controlled repository on GitHub at https://github.com/isakro/se.norway.shoredate.14c.
+This has been archived on Zenodo at https://doi.org/10.5281/zenodo.10449324.
 
@@
 <div id="refs"></div>
+
+# Supplementary material{-}
 

### 2024-01-18 analysis/paper/paper.Rmd
Updated figure formating for JAMT

@@
 bibliography: references.bib
-csl: "../templates/saa.csl" # Insert path for the bib-style
+csl: "../templates/apa.csl" # Insert path for the bib-style
 abstract: |
@@
 In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites.
-The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)).
+The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Fig.
+\@ref(fig:map)).
 The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility.
@@
 However, idiosyncrasies in the dating method did necessitate some adjustments.
-To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate) and outlined below.
+To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Fig.
+\@ref(fig:shoredate) and outlined below.
 
-Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B).
-These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Figure \@ref(fig:shoredate)A].
+Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Fig.
+\@ref(fig:shoredate)B).
+These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Fig.
+\@ref(fig:shoredate)A].
 The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated.
@@
 With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline.
-Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve.
+Fig.
+\@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve.
 As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
@@
 The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833.
-This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D).
+This is incorporated in the procedure for shoreline dating used for this study (Fig.
+\@ref(fig:shoredate)D).
 When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve.
@@
 
-The implications of different approaches to shoreline dating and summing the results are illustrated in Figure \@ref(fig:sumshore).
-In Figure \@ref(fig:sumshore)A the sites to be shoreline dated are plotted according to their elevation.
+The implications of different approaches to shoreline dating and summing the results are illustrated in Fig.
+\@ref(fig:sumshore).
+In Fig.
+\@ref(fig:sumshore)A the sites to be shoreline dated are plotted according to their elevation.
 Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution.
@@
 
-When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C).
+When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Fig.
+\@ref(fig:shoredate)C).
 As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years.
-This approach is illustrated in Figure \@ref(fig:sumshore)B.
-The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c.
+This approach is illustrated in Fig.
+\@ref(fig:sumshore)B.
+The SSPD resulting from using the probabilistic method of shoreline dating, given in Fig.
+\@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c.
 8000 BCE.
@@
 As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites.
-These intervals were assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).
+These intervals were assigned a weight based on how the density of observed sites is distributed among them (Fig.
+\@ref(fig:incpolys)).
 
@@
 ```
-The results of the Monte Carlo simulations are given in Figure \@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values.
-The comparison of the models based on BIC scores is given in Figure \@ref(fig:bic).
+The results of the Monte Carlo simulations are given in Fig.
+\@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values.
+The comparison of the models based on BIC scores is given in Fig.
+\@ref(fig:bic).
 
-For the RSPD, the logistic model achieved a p-value of 0.072 and could therefore not be rejected (Figure \@ref(fig:spdmc)E).
+For the RSPD, the logistic model achieved a p-value of 0.072 and could therefore not be rejected (Fig.
+\@ref(fig:spdmc)E).
 As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development.
@@
 Of the contrasted models, the BIC values indicate that the exponential model is the best candidate.
-Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A).
+Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Fig.
+\@ref(fig:spdmc)A).
 Some significant deviations occur with the model over-predicting the frequency of sites in the period from c.
@@
 @crema2022].
-The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of RSL change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D).
+The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Fig.
+\@ref(fig:incpolys)A), the corresponding local trajectories of RSL change (Fig.
+\@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Fig.
+\@ref(fig:shoredate)D).
 Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models.
@@
 
-# Acknowledgement{-}
+# Acknowledgements{-}
 We would like to thank David Wright for commenting on an earlier draft of this paper.

### 2024-01-18 R/01_shoreline_spd.R

@@ -142,4 +142,6 @@ eplot <- ggplot() +
   theme_bw()
 
 
-eplot + wplot + splot + plot_annotation(tag_levels = 'A')
+eplot + wplot + splot + plot_annotation(tag_levels = 'a')
+ggsave(here::here("analysis/figures/summed_examples.png"),
+       units = "px", width = 4800, height = 1800, dpi = 400)

### 2024-01-18 R/02_displacement_prep.R

@@ -133,7 +133,7 @@ displt <- ggplot(as.data.frame(dispcurves)) +
                                     ymax = upperelev,
                                     group = id,
                                     alpha = dens)) +
-  labs(title = "B", x = "BCE/CE", y = "Meters above present sea-level") +
+  labs(title = "b", x = "BCE/CE", y = "Meters above present sea-level") +
   scale_x_continuous(expand = c(0,0), limits = c(-9469, 2000)) +
   theme_bw() +
   theme(legend.position = "none")
@@ -155,7 +155,7 @@ mplt <- ggplot() +
   #                border.size = 0.1, st.dist = 0.03,
   #                anchor = c(x = anc[2] - 15500, y = anc[1]) + 8000) +
   scale_fill_gradient(low = "white", high = "black", name = "Site density") +
-  ggtitle("A") +
+  ggtitle("a") +
   coord_sf(xlim = c(bboxpolys[1], bboxpolys[3]),
            ylim = c(bboxpolys[2], bboxpolys[4]),
            expand = FALSE) +
@@ -170,4 +170,4 @@ mplt <- ggplot() +
 mplt + displt
 
 ggsave(here::here("analysis/figures/incpolys.png"),
-       units = "px", width = 2650, height = 1600)
+       units = "px", width = 3975, height = 2400, dpi = 400)

### 2024-01-18 R/05_radiocarbon_monte_carlo.R

@@ -76,10 +76,10 @@ load(here("analysis/data/derived_data/uniplts.rda"))
 
 (expplts + logplts + uniplts)/
 (expplotr + logplotr + uniplotr) +
-  plot_annotation(tag_levels = "A")
+  plot_annotation(tag_levels = "a")
 
 ggsave(here::here("analysis/figures/mc.png"),
-       units = "px", width = 4000, height = 1250*2)
+       units = "px", width = 6000, height = 3750, dpi = 400)
 
 
 #### Logistic model with taphonomic loss ####
@@ -90,7 +90,7 @@ logtaph <- JDEoptim(lower = c(0, 0000, 0, -3), upper = c(1, 10000, 20000,0),
                     fn = objectiveFunction, PDarray = PD, type = 'logistic',
                     NP = 40, taphonomy = TRUE)
 
-# Retrieve the logistic model fit without accoutning for taphonomy
+# Retrieve the logistic model fit without accounting for taphonomy
 pop_log <- convertPars(pars=log$par,
                        years=minage:maxage,
                        type='logistic')
@@ -144,7 +144,7 @@ taphplt <- ggplot() +
   theme_bw()
 
 popplt + taphplt +
-  plot_annotation(tag_levels = 'A')
+  plot_annotation(tag_levels = 'a')
 
 ggsave(here::here("analysis/figures/taphonomic.png"),
-       units = "px", width = 3000, height = 1500)
+       units = "px", width = 4500, height = 2250, dpi = 400)

### 2024-01-18 R/06_bic.R

@@ -41,4 +41,4 @@ bicplt_shore + bicplt_rcarbon
 
 # Save plot
 ggsave(filename = here::here("analysis/figures/bic.png"),
-       units = "px", width = 2000, height = 900)
+       units = "px", width = 3000, height = 1350, dpi = 400)

### 2024-01-18 R/08_map_plots.R

@@ -95,7 +95,8 @@ sa <- ggplot() +
                      axis.ticks = element_blank(),
                      panel.grid.major = element_blank(),
                      legend.position = "bottom",
-                     legend.text=element_text(size = 11))
+                     legend.text=element_text(size = 11),
+                     legend.spacing.x = unit(5, "mm"))
 
 # Retrieve site names for plotting of radiocarbon data
 c14_names <- tools::file_path_sans_ext(colnames(PD))
@@ -136,7 +137,7 @@ plt1 <- cowplot::ggdraw() +
                      y = 0.7, width = 0.3, height = 0.3)
 
 (plt1)/
-  pltc14 + plot_annotation(tag_levels = "A")
+  pltc14 + plot_annotation(tag_levels = "a")
 
 # Example date plot
 
@@ -167,4 +168,7 @@ shoredate_plot(shdate, greyscale = TRUE)
 splt <- last_plot()
 
 (tplt + dplt) /
-  (tpqplt + splt) + plot_annotation(tag_levels = "A")
+  (tpqplt + splt) + plot_annotation(tag_levels = "a")
+
+ggsave(here("analysis/figures/example_shoredate.png"),
+       width = 4200, height = 3400, dpi = 400, units = "px")

### 2025-01-14 analysis/paper/paper.Rmd
Final edits after peer review

@@
     bookdown::pdf_document2:
+      extra_dependencies: ["float"]
       fig_caption: true
@@
 Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns.
-Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
+Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data could be reflecting population dynamics more directly.
 keywords: | 
@@
 ---
+#' \pagenumbering{gobble}
+
 Keywords: `r rmarkdown::metadata$keywords`
@@
 library(sf)
-library(shoredate)
 ```
@@
 sites <- st_read(here("analysis/data/derived_data/combined_sites.gpkg"))
+log_params <- read.csv(here("analysis/data/derived_data/logistic_parameters.csv"))
+load(here("analysis/data/derived_data/c14_details.RData"))
+
 ```
@@
 # Introduction
-Variations in population patterns are regarded as important drivers and potential outcomes of cultural variation, and are thus of critical importance to our understanding of past human societies [e.g.
+Population patterns are regarded as important drivers and potential outcomes of cultural variation, and are thus of critical importance to our understanding of past human societies [e.g.
 @french2021; @shennan2000; @shennan2021].
@@
 In this paper, we report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD).
+While fundamental questions concerning the reliability of shoreline dating still persist, our analysis draws on recent developments and ongoing efforts to formalise and better understand the dating method [@roalkvam2023; @rosenvinge2022, 21--41].
 We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c.
@@
 @mörner1979; @steffen2011].
-The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or located close to the contemporaneous shoreline at their time of use.
-Thus, as coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites.
+The majority of Mesolithic sites along the Norwegian Skagerrak coast are interpreted as having been shore bound or located close to the contemporaneous shoreline at their time of use [see e.g.
+@darmark2018a; @solheim2020a; @mjærum2022; @roalkvam2023].
+Consequently, as coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the occupation of the sites.
 This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating.
@@
 @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021].
-However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
+These point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method.
 In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites.
@@
 The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility.
-Thus, this geographical and temporal limit are also used for this study.
+Consequently, these geographical and temporal limits are also used for this study.
+
+Limiting this study to the same temporal and geographical constraints as was used in @roalkvam2023 should mitigate some issues that could follow from undue extrapolation of the technique to areas where the reliability of the method and the parametrisation employed here has not been assessed and therefore might not apply.
+However, while we view the long history of use and research into shoreline dating as a testament to its general applicability, and these recent developments as an improvement of this framework, it should also be underscored that the method for shoreline dating employed in this study is still hefted with significant questions regarding its reliability [see @roalkvam2023].
 
 The present study is decidedly exploratory.
-While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses.
-Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables.
+While some thoughts concerning the relationship between the frequency of radiocarbon dates, shoreline dated sites and population dynamics underlie the analysis, these could not be instantiated as concrete hypotheses.
+Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between these variables.
 The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics.
@@
 2500 BCE, the Mesolithic is the main focus of this analysis.
-Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates.
+Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of `r all_included` ^14^C-dates.
 Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.
+Of the final sample, `r char_nut` (`r round((char_nut/all_included)*100)`%) dates are made on charcoal and burnt nutshell and `r other` (`r round((other/all_included)*100)`%) on other material such as bone.
 
@@
 This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017].
-Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021].
-The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.
+Analysis of the ^14^C-dates was mainly done using the R package *ADMUR* v1.0.3 [@admur2020; @timpson2021] and with some use of *rcarbon* v1.4.4 [@crema2021].
+The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* used for subsequent model fitting, Monte Carlo simulation and model comparison.
 
@@
 
-```{r map, echo = FALSE, fig.cap = "A) Map displaying the location of the study area and shoreline dated sites.
+```{r map, echo = FALSE, fig.cap = "a) Map displaying the location of the study area and shoreline dated sites.
 Black lines indicate the borders between municipalities.
 The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1.
-B) The distribution of radiocarbon dates employed in the study.", out.width = "500px", fig.align="center", regfloat = TRUE}
+b) The distribution of radiocarbon dates employed in the study.", out.width = "500px", fig.align="center", regfloat = TRUE}
 
@@
 All calibrations were performed using the IntCal20 calibration curve [@reimer2020].
+It can be noted that while several authors have pointed to the fact that normalising the dates can lead to spikes in the RSPD [e.g.
+@crema2021, 27--29; @weninger2015], not normalising the dates did not result in any meaningful difference in the resulting best-fit models for the purposes of this paper (see supplementary figure).
 
@@
 The Monte Carlo simulations were then performed using the *SPDsimulationTest* function from *ADMUR*.
-For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD.
+For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of site phases in the observed RSPD.
 These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date.
@@
 Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Fig.
-\@ref(fig:shoredate)B).
+\@ref(fig:shoredate)b).
 These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Fig.
-\@ref(fig:shoredate)A].
+\@ref(fig:shoredate)a].
 The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated.
@@
 Fig.
-\@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve.
+\@ref(fig:shoredate)c indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve.
 As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date.
 However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken.
-Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
+Furthermore, the common practice of adding a constant error or binning point estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.
 
@@
 This is incorporated in the procedure for shoreline dating used for this study (Fig.
-\@ref(fig:shoredate)D).
+\@ref(fig:shoredate)d).
 When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve.
@@
 
-```{r shoredate, echo = FALSE, fig.cap = "A) The location of an example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.
-B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site.
-C) The black line on the y-axis represents the elevation of the site.
+```{r shoredate, echo = FALSE, fig.cap = "a) The location of an example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.
+b) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site.
+c) The black line on the y-axis represents the elevation of the site.
 This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea and represents a \\textit{terminus post quem} date for the site.
-D) The shoreline dating approach used for this study, where the gamma function on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
+d) The shoreline dating approach used for this study, where the gamma function on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}
 
@@
 In Fig.
-\@ref(fig:sumshore)A the sites to be shoreline dated are plotted according to their elevation.
+\@ref(fig:sumshore)a the sites to be shoreline dated are plotted according to their elevation.
 Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution.
@@
 
-When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Fig.
-\@ref(fig:shoredate)C).
-As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years.
+When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Fig.
+\@ref(fig:shoredate)c).
+As outlined above,  these point estimates are then typically aggregated in bins of 200 or 500 years.
 This approach is illustrated in Fig.
-\@ref(fig:sumshore)B.
+\@ref(fig:sumshore)b.
 The SSPD resulting from using the probabilistic method of shoreline dating, given in Fig.
-\@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c.
+\@ref(fig:sumshore)c, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c.
 8000 BCE.
 
-```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by altitude above present sea-level.
+```{r sumshore, echo = FALSE, fig.cap = "a) The distribution of the shoreline dated sites (n = 921) by altitude above present sea-level.
 The binwidth is 4 m.
-B) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates.
-C) The result of instead summing the probability of the shoreline dates achieved with the method employed here, which accounts for the likely distance between sites and shoreline.
+b) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates.
+c) The result of instead summing the probability of the shoreline dates achieved with the method employed here, which accounts for the likely distance between sites and shoreline.
 The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}
@@
 
-In the same way as with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above.
+In the same way as with an RSPD, the SSPD cannot merely be assessed visually due to the summation issues introduced above.
 Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of RSL change effectively functions as the calibration curve for each site.
@@
 
-```{r incpolys, echo = FALSE, fig.cap = "A) Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 921) as distributed across 2 km wide line segments that run perpendicular to the shoreline gradient.
-B) The displacement curves interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
+```{r incpolys, echo = FALSE, fig.cap = "a) Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 921) as distributed across 2 km wide line segments that run perpendicular to the shoreline gradient.
+b) The displacement curves interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
 knitr::include_graphics(here("analysis/figures/incpolys.png"))
@@
 From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model.
-This penalises the models for the number of parameters in use to avoid over-fitting, and provides a measure for the relative performance of the models.
+This penalises the models for the number of parameters in use to avoid over-fitting and provides a measure for the relative performance of the models.
 
@@
 ```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates.
-A) Summed probability of shoreline-dated sites (n = 921) compared to an exponential null model.
-B) SSPD compared to a logistic null model.
-C) SSPD compared to a uniform null model.
-D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model.
-E) RSPD compared to a logistic null model.
-F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
+a) Summed probability of shoreline-dated sites (n = 921) compared to an exponential null model.
+b) SSPD compared to a logistic null model.
+c) SSPD compared to a uniform null model.
+d) Summed probability of calibrated radiocarbon dates (n = 352, bins = 153) compared to an exponential null model.
+e) RSPD compared to a logistic null model.
+f) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
 knitr::include_graphics(here("analysis/figures/mc.png"))
 ```
+
 The results of the Monte Carlo simulations are given in Fig.
@@
 
-For the RSPD, the logistic model achieved a p-value of 0.072 and could therefore not be rejected (Fig.
-\@ref(fig:spdmc)E).
+For the RSPD, the logistic model achieved a p-value of 0.073 and could therefore not be rejected (Fig.
+\@ref(fig:spdmc)e).
 As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development.
@@
 Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Fig.
-\@ref(fig:spdmc)A).
+\@ref(fig:spdmc)a).
 Some significant deviations occur with the model over-predicting the frequency of sites in the period from c.
@@
 
+We start out here by presenting some potentially substantive implications of the results before moving on to consider methodological limitations that could undermine our findings.
 Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g.
@@
 A possible lack of reliable access to firewood has also been seen as a possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; see also @damm2022; @sørensen2014].
-To assess the potential influence of taphonomy on the logistic model for the RSPD, we fit a version of the model that accounts for taphonomy, using the workflow from *ADMUR*.
-As this did not alter the MLE parameters of the model, the results of this exercise have been relegated to the supplementary material.
-While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.
+The earliest dates in our dataset are made on charcoal, which dominates the dated material overall (see figure in the supplementary material), as other materials such as bone make up a very small portion of the dated material.
+Furthermore, to assess the potential influence of taphonomy on the logistic model for the RSPD, we fit a version of the model that attempts to account for taphonomy using the workflow from *ADMUR*.
+Unlike approaches that adjust the SPD directly, using for example the global point estimates for taphonomic correction presented by @surovell2009b and @bluhm2019, *ADMUR* instead treats this is as additional parameters to be included in the model itself.
+We used the default constraints for the taphonomy parameters recommended by *ADMUR*, as there is no readily available data that can inform these in our context.
+As the result of including taphonomy did not alter the MLE parameters of the model, owing also to the fairly small sample of radiocarbon dates with which to narrow the taphonomic parameters, this exercise has been relegated to the supplementary material.
+In sum therefore, while taphonomic and ecological factors cannot be excluded as part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind the observed pattern.
 
@@
 High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018].
-The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
+The initial peak in the SSPD and the trajectory of the exponential model could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period.
 The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [@damm2022] and low population numbers.
@@
 The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE.
+
+Given that it is well established that people were present in Norway before the earliest ^14^C-dates in our dataset, the identified logistic model does not capture an initial colonisation.
+The above-mentioned explanations of low population size and/or mobility patterns that are potentially causing the lacking radiocarbon record in the earliest period of human presence could therefore both be causing the under-prediction of the antiquity of human presence, and, as a consequence, also lead to an exaggeration in the steepens of the growth in the logistic model.
+If these factors are then counteracted by migratory events around the time of the earliest ^14^C-dates, the subsequent plateau of the logistic growth curve could nonetheless represent an average carrying capacity, around which the population density has likely fluctuated (e.g.
+@tallavaara2021).
+
 The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.
-
 Subsequently, the SSPD deviates negatively from the exponential null model from c.
@@
 This is followed by a positive deviation in the last centuries before 5000 BCE.
-These developments should possibly be seen in relation to simultaneous developments in the inland areas of southern Norway.
+
+These developments should possibly be seen in relation to simultaneous developments in the inland areas of southern Norway that are located beyond the the highest post-glacial elevation of the sea.
 The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c.
@@
 This uncertainty is not properly accounted for in the implemented method, as the MLE parameters of the gamma distribution in use was found when treating all sites older than 2500 BCE in aggregate.
-Additionally, the MLE parameters for the gamma distribution are point-estimates that do not account for the underlying sample size used for deriving them.
+Additionally, the MLE parameters for the gamma distribution are point estimates that do not account for the underlying sample size used for deriving them.
 Furthermore, the rapid shoreline displacement in the early Holocene is also a challenge for the geological reconstructions themselves, and there is reason to question whether this is properly accounted for in the compiled time-series for RSL-change in the region [see e.g.
@@
 
+Furthermore, as was suggested by two reviewers of this paper, it could be possible to evaluate the method of shoreline dating by performing the dating procedure on the radiocarbon dates that underlie the RSPD.
+By comparing the RSPD with the resulting SSPD, this could make it possible to assess how well shoreline dating captures anthropogenic activity represented by ^14^C-dates.
+However, the RSPD is the sum of all radiocarbon dates with an anthropogenic origin from within the study area, irrespective of their context beyond this.
+Any mismatch between the RSPD and a SSPD that would result from shoreline dating the radiocarbon dates would therefore have had to be followed by a *post hoc* evaluation of the potential causes for this deviation -- whether related to the unreliability of shoreline dating, issues or non-relevance of individual radiocarbon dates, or issues to do with topographic disturbances or uncertain elevation data that might influence the shoreline date in each case.
+Consequently, as any observed deviations in such a comparison would have to be explored in a *post hoc* manner, we have instead opted to save this data for a more comprehensive and rigorous test of shoreline dating that is currently in process of being preregistered [see e.g.
+@ross2022], where the reliability and relevance of available radiocarbon data will be evaluated before being subjected to the test [following e.g.
+@pettitt2003].
+
 Another relevant point to mention here is that the RSL change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g.
@@
 The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Fig.
-\@ref(fig:incpolys)A), the corresponding local trajectories of RSL change (Fig.
-\@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Fig.
-\@ref(fig:shoredate)D).
+\@ref(fig:incpolys)a), the corresponding local trajectories of RSL change (Fig.
+\@ref(fig:incpolys)b), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Fig.
+\@ref(fig:shoredate)d).
 Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models.
@@
 Some of the deviations from the model were then narratively explored to suggest some possible reasons for these deviations.
-However, given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.
+However, given the lack of a model that adequately describe the data, and consequently the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.
 
@@
 # Data, scripts, code, and supplementary information availability{-}
-All data, code, text and figures used for this paper is freely available in a version-controlled repository on GitHub at https://github.com/isakro/se.norway.shoredate.14c.
-This has been archived on Zenodo at https://doi.org/10.5281/zenodo.10449324.
+All data, code, text and figures used for this paper are freely available in a version-controlled repository on GitHub at https://github.com/isakro/se.norway.shoredate.14c.
+This has been archived on Zenodo at https://doi.org/10.5281/zenodo.8373856
 
@@
 <div id="refs"></div>
+\newpage
 
@@
 
-```{r taph, echo = FALSE, fig.cap = "The result of accounting for taphonomic loss in the analysis of the radiocarbon data.
-A) Comparison of MLE parameters for the logistic population dynamics models when both accounting for and disregarding taphonomy using the ADMUR approach.
-These are superimposed on the RSPD.
-As indicated by their perfect overlap, accounting for taphonomy does not impact the MLE parameters.
-B) The taphonomic component of the logistic model estimated when accounting for taphonomy.", out.width = "500px", fig.align="center", regfloat = TRUE}
+```{r material, echo = FALSE, fig.cap = "The dated material in the data set divided into the categories 'charcoal and burnt nutshell' and 'other', which primarily consists of bone.", out.width = "500px", fig.align="center", regfloat = TRUE}
+knitr::include_graphics(here("analysis/figures/dated_material.png"))
+```
+
+```{r norm, echo = FALSE, fig.cap = "Comparison of the effect of normalising and not normalising calibrated dates when identifying the best logistic fit.
+As not normalising calibrated dates is not currently possible with ADMUR (see https://github.com/UCL/ADMUR/issues/1) rcarbon was used for this step.
+The slight discrepancy between the MLE parameters of the two models has no meaningful relevance for the purposes of this paper.", out.width = "500px", fig.align="center", regfloat = TRUE}
+knitr::include_graphics(here("analysis/figures/normalisation.png"))
+```
+
+```{r mc-nonnorm, echo = FALSE, fig.cap = "Monte Carlo simulation using non-normalised dates for the RSPD.
+Not normalising calibrated dates is not currently possible with ADMUR (see also figure above).
+Consequently, the figure has been produced using rcarbon for the calibration of dates, the assemblage of the RSPD and for the Monte Carlo simulations.
+All which might have slight algorithmic differences with ADMUR and the approach used in the main text.
+The null models have been fit using the ADMUR approach.
+a) Exponential null model, b) Logistic null model, c) Uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
+knitr::include_graphics(here("analysis/figures/mc_nonnorm.png"))
+```
+
+```{r taph, echo = FALSE, fig.cap = "The result of accounting for taphonomic loss in the analysis of the radiocarbon data using the ADMUR approach.
+a) Comparison of ML parameters and samples from the joint posterior distributions for the population dynamics components of the logistic models when both accounting for and disregarding taphonomy.
+The ML parameters perfectly overlap (see Table 2 below for credible intervals).
+Note that unlike in the main text we here use the cal BP calendar scale to be consistent with the model parameters.
+b) The ML and joint posterior distributions of the taphonomic parameters b and c (following the ADMUR notation), constituting the taphonomic component of the logistic model that accounts for taphonomy.
+This figure indicates that the underlying data provides limited information with which to constrain the default priors.", out.width = "500px", fig.align="center", regfloat = TRUE}
 knitr::include_graphics(here("analysis/figures/taphonomic.png"))
 ```
+
+\newpage
+```{r tab, fig.pos='H', out.extra = ""}
+
+logp1 <- paste0(log_params[1, 2], " (", log_params[1,3],
+                " to ", log_params[1, 4], ")")
+logp2 <- paste0(log_params[1, 5], " (", log_params[1, 6],
+                " to ", log_params[1, 7], ")")
+
+taphp1 <- paste0(log_params[2, 2], " (", log_params[2, 3],
+                " to ", log_params[2, 4], ")")
+taphp2 <- paste0(log_params[2, 5], " (", log_params[2, 6],
+                " to ", log_params[2, 7], ")")
+
+params <- data.frame(m = c(log_params[1,1], log_params[2,1]),
+           p1 = c(logp1, taphp1),
+           p2 = c(logp2, taphp2))
+names(params) <- c("", "Rate (\\textit{k})", "Centre (\\textit{x}\\textsubscript{0})")
+
+kableExtra::kable(params, format = "latex", booktabs = TRUE, escape = FALSE,
+      caption = "Population dynamics parameters for the logistic model when disregarding and including a taphonomic component.
+The 95\\% credible intervals in the parentheses are based on quantiles.") %>%
+  kable_styling(latex_options = "HOLD_position")
+```

### 2025-01-14 R/03_functions.R

@@ -64,7 +64,7 @@ simulation_summary <- function(spd, simulation_results,
 
   calBP <- as.numeric(rownames(spd))
 
-  # Code below repurposed from ADMUR
+  # Code below re-purposed from ADMUR
 
   # expected simulation
   expected.sim <- rowMeans(SPD.sims)

### 2025-01-14 R/05_radiocarbon_monte_carlo.R

@@ -1,16 +1,19 @@
 # This script performs Monte Carlo simulation using the radiocarbon data and
-# plots results from both simulations with both shoreline and radiocarbon dates.
-# Based on the failed rejection of the logistic model, a version of the model
-# fit when accounting for taphonomy is then created (figure included in
-# as supplementary material)
+# plots results from simulations with both shoreline and radiocarbon dates.
+# The impact of normalising and not normalising the dates is then explored
+# (included as a supplementary figure), and, based on the failed rejection of
+# the logistic model, a version of the model fit when accounting for taphonomy
+# is then explored (included as supplementary figures).
 
 library(dplyr)
 library(rcarbon)
 library(ADMUR)
 library(patchwork)
 library(ggplot2)
+library(ggnewscale)
 library(DEoptimR)
 library(here)
+library(ggplotify)
 
 set.seed(42)
 
@@ -26,8 +29,6 @@ c14spd <- spd(caldates, timeRange = c(12000, 4500))
 
 minage <- min(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
 maxage <- max(c14spd$grid[c14spd$grid$PrDens > 0,]$calBP)
-
-# c14spd_taph <- transformSPD(c14spd)
 
 # Switch to ADMUR
 names(c14) <- c("site", "context", "material", "lab_ref", "age", "sd")
@@ -64,7 +65,7 @@ plotSimulationSummary(logsum)
 plotSimulationSummary(unisum)
 
 
-# Plotting using custom plot function
+# Plot using custom plot function
 expplotr <- plot_mc(expsum)
 logplotr <- plot_mc(logsum)
 uniplotr <- plot_mc(unisum)
@@ -81,11 +82,298 @@ load(here("analysis/data/derived_data/uniplts.rda"))
 ggsave(here::here("analysis/figures/mc.png"),
        units = "px", width = 6000, height = 3750, dpi = 400)
 
+##### Fitting models to non-normalised dates #####
+
+# The code below using the rcarbon approach for calibrating and summing without
+# normalising, and then fits the models using the ADMUR approach
+
+# Read in 14C-data again
+c14 <- read.csv(here::here("analysis/data/raw_data/radiocarbon.csv"))
[…]

### 2025-10-03 README.Rmd
Additional analyses after PCI review
Update README.Rmd

@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Roalkvam, Isak <a href="https://orcid.org/0000-0001-6974-1374"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> and Solheim, Steinar <a href="https://orcid.org/0000-0001-8293-8147"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> (2025).
+_`r Title`_.
+*Journal of Archaeological Method and Theory* 32(1):26 <https://doi.org/10.1007/s10816-025-09696-7>
 
-Our pre-print is online here:
+Our pre-print that was reviewed and recommended through [PCI Archaeology](https://archaeo.peercommunityin.org/articles/rec?id=417) is online here:
 
-> Roalkvam, Isak <a href="https://orcid.org/0000-0001-6974-1374"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> and Solheim, Steinar <a href="https://orcid.org/0000-0001-8293-8147"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> (2023).
+> Roalkvam, Isak <a href="https://orcid.org/0000-0001-6974-1374"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> and Solheim, Steinar <a href="https://orcid.org/0000-0001-8293-8147"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> (2024).
 _`r Title`_.
-Online at <https://doi.org/xxx/xxx>
+*Peer Community in Archaeology*.
+Online at <https://doi.org/10.31235/osf.io/2f8ph>
 
@@
 
-The **R** directory contains scripts for performing the analyses and create the
+The **R** directory contains scripts for performing the analyses and creating the
 figures used in the paper.
